---
title: "DDfathers_figures"
author: "Matt Neville"
date: "09/07/2026"
output: html_document
---

## a) Prep
### Paths
```{r paths}
# All input files needed to run this notebook live in ./data
dataPath <- "data/"
# All figures produced by this notebook are written to ./output
outputPath <- "output/"
dir.create(outputPath, showWarnings = FALSE)
```

### Libraries
```{r libraries}
library(scales)
library(readxl)
library(ggrepel)
library(patchwork)
library(grid)
library(ggpubr)
library(png)
library(tidyverse)
```

### Sperm data
```{r samples}

# Combined Table S1-S10 workbook. Table S6 (Sperm Samples) and Table S8 (Sperm Vars)
supTables_list <- excel_sheets(paste0(dataPath, "Supplementary_Tables.xlsx")) |>
  set_names() |>
  map(~ read_xlsx(paste0(dataPath, "Supplementary_Tables.xlsx"), sheet = .x))

# Per-sample sperm metadata and mutation burdens (Table S6 SpermSamples).
spermSamplesFiltered <- supTables_list[["Table S6 SpermSamples"]] |>
  mutate(Sample_PD_ID = paste0(PD_ID, "_tds0001")) |>
  mutate(dataset = Cohort)

# Targeted variants: derived from Table S8 (Sperm Vars)
spermVariants <- supTables_list[["Table S8 SpermVars"]] |>
  mutate(Sample_PD_ID = paste0(PD_ID, "_tds0001")) |>
  rename(start = start_hg19, csq = VEP_csq, IMPACT = VEP_IMPACT,
         TIMES_CALLED = DUPLEX_MUT, PHEN = ClinVar_CLNDN, CLNSIG = ClinVar_CLNSIG,
         CLNSIGCONF = ClinVar_CLNSIGCONF) |>
  mutate(class = if_else(TYPE == "snv", "snv", "indel"))

# Per-family diagnostic DNM calls, with duplex and BAM support (Table S4 DiagDNMs).
diagnosticVarsCalls <- supTables_list[["Table S4 DiagDNMs"]] |>
  rename(duplexCalls = sperm_duplexCalls, duplexCov = sperm_duplexCov, duplexVAF = sperm_duplexVAF,
         BAM_MUT = sperm_BAM_calls, BAM_COV = sperm_BAM_cov, BAM_VAF = sperm_BAM_VAF) |>
  mutate(varLabel = if_else(coalesce(duplexCalls, 0) > 0, gene, ""))

table(spermVariants$TYPE)

# Note: "German"/"germanPanel" (variable and chunk names, filter strings,
# tukGermanPanelVafSums.tsv) refers to the DD-father cohort throughout - as
# opposed to Apert fathers or TwinsUK Controls.

# Colour palette
dataset_colours <- c("DD fathers" = "#ef8a62",
                     "Controls" = "#67a9cf",
                     "Apert fathers" = "#7fbf7b",
                    "DD fathers & Controls" = "#654E4E",
                    "NS" = "grey60",
                    "Trio DNMs" = "#666A86")


```

### Trio data
```{r dnms}

# Trio de novo mutation calls, including parent-of-origin phase (Table S2 DNMs).
trioDNM_Vars <- supTables_list[["Table S2 DNMs"]] |>
  rename(start = start_hg19, PD_ID = PD_ID_Proband, IMPACT = VEP_IMPACT)

table(trioDNM_Vars$phase)
dim(trioDNM_Vars)[1]
dim(trioDNM_Vars |> filter(ref %in% c("A", "T", "C", "G") & alt %in% c("A", "T", "C", "G")))[1]

# Trinucleotide-context counts of trio DNMs, for the mutational signature analysis (computed in DDfathers_processing.Rmd, Signatures section).
trinucDNMs <- read_tsv(paste0(dataPath, "trinucDNMs.tsv"), col_types = cols())
sum(trinucDNMs$dnmCount)


# Trio-level DNM burdens and covariates for DD-father families (Table S3 Trios).
trioBurdens <- supTables_list[["Table S3 Trios"]] |>
  filter(Cohort == "DD fathers") |>
  mutate(dataset = "Trio DNMs") |>
  rename(PD_ID = PD_ID_Proband, Pat_Age_at_birth = Pat_age_at_birth, Mat_Age_at_birth = Mat_age_at_birth,
         MEDIAN_COVERAGE = Proband_median_cov, Pat_seq_x = Pat_median_cov, Mat_seq_x = Mat_median_cov,
         sample.median.0.3vaf = DNM_0.3_medianVAF)
mean(trioBurdens$Pat_Age_at_birth)
mean(trioBurdens$Mat_Age_at_birth)
median(trioBurdens$MEDIAN_COVERAGE)

# Sample overlap between sperm + Trios
dim(spermSamplesFiltered |> filter(dataset == "DD fathers" & FAM_ID %in% trioBurdens$FAM_ID))[1]

# Trio-level DNM burdens for the SFHS control cohort (Table S3 Trios, ControlTrios rows).
trioControlBurdens <- supTables_list[["Table S3 Trios"]] |>
  filter(Cohort == "ControlTrios") |>
  mutate(dataset = "Controls") |>
  rename(Pat_Age_at_birth = Pat_age_at_birth) |>
  mutate(Indel_DNMs_0.3VAF = coalesce(Indel_DNMs_0.3VAF, 0))
mean(trioControlBurdens$Pat_Age_at_birth)
range(trioControlBurdens$Pat_Age_at_birth)
sd(trioControlBurdens$Pat_Age_at_birth)

mean(trioBurdens$Mat_Age_at_birth)

trioSigs <- trioBurdens |> 
  summarize(SBS1 = sum(SBS1), SBS5 = sum(SBS5))
```

### Stats
```{r sampleStats}
statsVars <- spermVariants |> filter(str_detect(Sample_PD_ID, "PD6")) |> 
  mutate(coding = if_else(csq %in% c("missense", "synonymous", "stop_lost", "start_lost", "splice_acceptor", "splice_donor", "stop_gained", "inframe_insertion", "inframe_deletion", "coding_sequence"), "coding region", "non-coding"))
print(paste("Total targeted vars:", dim(statsVars)[1]))
table(statsVars |> pull(coding))

#VAFs
dim(statsVars |> filter(TIMES_CALLED == 1))[1]/dim(statsVars)[1]
dim(statsVars |> filter(TIMES_CALLED > 1 & BAM_VAF>0.01))[1]/dim(statsVars)[1]

mean(spermVariants |> filter(str_detect(Sample_PD_ID, "PD6")) |> pull(DUPLEX_COV))
mean(spermVariants |> filter(str_detect(Sample_PD_ID, "PD6")) |> pull(BAM_COV))

statsVarsApert <- spermVariants |> filter(str_detect(Sample_PD_ID, "PD5")) |> 
  mutate(coding = if_else(csq %in% c("missense", "synonymous", "stop_lost", "start_lost", "splice_acceptor", "splice_donor", "stop_gained", "inframe_insertion", "inframe_deletion", "coding_sequence"), "coding region", "non-coding"))
print(paste("Total targeted vars:", dim(statsVarsApert)[1]))

targetVarsStats <- diagnosticVarsCalls |>
  filter(Cohort == "DD fathers", duplexCov > 100) |>
  mutate(
    excl95_pct = 100 * (1 - 0.05^(1 / duplexCov))
  )
```

### TwinsUK controlled-access data
```{r twinsUKAvailability}
# TwinsUK sperm NanoSeq data (Neville et al.) is under TwinsUK's controlled-access
# policy - see README for how to request it. If you have that access and have
# placed the individual-level TwinsUK files in data/, this notebook reproduces the
# manuscript figures exactly (DD fathers + Apert fathers + TwinsUK Controls).
# Otherwise it falls back to DD father (+ Apert father) data only, drawing the
# TwinsUK trend/fit lines from pre-computed aggregate summaries
# (data/twinsUK_*Fits.tsv, data/twinsUK_exomePanelCorrection_stats.tsv) where
# available - these contain only fitted Age-grid values and reported statistics,
# never individual TwinsUK data points.
twinsUKDataAvailable <- file.exists(paste0(dataPath, "tukGermanPanelVafSums.tsv"))

# Matches the grid the fits were saved on (by 1 year of age)
age_grid_fine <- tibble(Age = seq(10, 80, by = 1))

# Predict from a saved Age-grid fit table (columns: model, Age, fit, ci_lo, ci_hi)
# at arbitrary ages via linear interpolation - exact for the simple lm() fits
# saved here, and a good approximation for the quasibinomial smooth since the
# saved grid is by single year of age. rule = 2 extrapolates using the nearest
# boundary value instead of returning NA - needed because ages back-inferred to
# conception (Age - paternalAgeGap) can fall outside the observed TwinsUK range.
predict_from_fit_table <- function(fitTable, modelName, ages) {
  sub <- fitTable |> filter(model == modelName) |> arrange(Age)
  tibble(
    Age = ages,
    fit = approx(sub$Age, sub$fit, xout = ages, rule = 2)$y,
    ci_lo = approx(sub$Age, sub$ci_lo, xout = ages, rule = 2)$y,
    ci_hi = approx(sub$Age, sub$ci_hi, xout = ages, rule = 2)$y
  )
}
```

### SCN1A-ascertained families
```{r scn1a_ids}
# Identify SCN1A-ascertained FAM_IDs and their corresponding sperm Sample_PD_IDs.
# These are families enrolled specifically because of an SCN1A diagnostic variant, and are excluded in some analyses

scn1a_fam_ids   <- diagnosticVarsCalls |>
  filter(ANALYSIS_PERFORMED == "SCN1A ascertained") |>
  pull(FAM_ID) |> unique()

scn1a_sperm_ids <- spermSamplesFiltered |>
  filter(FAM_ID %in% scn1a_fam_ids) |>
  pull(Sample_PD_ID)

cat("SCN1A-ascertained families (n =", length(scn1a_fam_ids), "):", paste(scn1a_fam_ids, collapse = ", "), "\n")
cat("Corresponding sperm samples (n =", length(scn1a_sperm_ids), "):", paste(scn1a_sperm_ids, collapse = ", "), "\n")
```

## b) Burdens
### Trio Burdens
```{r triBurden, fig.height = 5, fig.width = 10}
# Grid of ages to predict over
age_grid <- tibble(Pat_Age_at_birth = seq(20, 65, by = 5))

# Fit linear model and predict across age grid
fit_model_and_predict_dnms <- function(df, age_grid) {
  model <- lm(DNM ~ Pat_Age_at_birth, data = df)
  
  pred_mat <- sapply(age_grid$Pat_Age_at_birth, function(a) {
    predict(model, newdata = tibble(Pat_Age_at_birth = a), se.fit = TRUE, interval = "confidence", level = 0.95)$fit[1:3]
  })
  
  pred_df <- tibble(
    Pat_Age_at_birth = age_grid$Pat_Age_at_birth,
    fit = pred_mat[1,],
    ci_lo = pred_mat[2,],
    ci_hi = pred_mat[3,]
  )
  
  list(model = model, predictions = pred_df)
}

dnm_data_long <- trioBurdens |> 
  bind_rows(trioControlBurdens) |> 
  select(dataset, Pat_Age_at_birth, SNP_DNMs_0.3VAF, Indel_DNMs_0.3VAF) |> 
  rename(snv = SNP_DNMs_0.3VAF, indel = Indel_DNMs_0.3VAF) |> 
  pivot_longer(cols = c(snv, indel), names_to = "class", values_to = "DNM") |> 
  drop_na(DNM) |> 
  mutate(dataset = factor(dataset, levels = c("Controls", "Trio DNMs")))

all_preds <- dnm_data_long |> 
  group_by(dataset, class) |> 
  group_map(~ {
    result <- fit_model_and_predict_dnms(.x, age_grid)
    
    result$predictions |> 
      mutate(
        dataset = .y$dataset,
        class = .y$class
      )
  }) |> 
  bind_rows() |> 
    mutate(dataset = factor(dataset, levels = c("Trio DNMs", "Controls")))

# Filter only Trio DNMs, for SNVs and indels separately
dnm_trio_snv <- dnm_data_long |> filter(dataset == "Trio DNMs", class == "snv")
dnm_trio_indel <- dnm_data_long |> filter(dataset == "Trio DNMs", class == "indel")

# Fit models
lm_snv_trio <- lm(DNM ~ Pat_Age_at_birth, data = dnm_trio_snv)
lm_indel_trio <- lm(DNM ~ Pat_Age_at_birth, data = dnm_trio_indel)

# Extract slope + 95% CI for Trio SNVs
slope_snv <- coef(lm_snv_trio)["Pat_Age_at_birth"]
ci_snv <- confint(lm_snv_trio)["Pat_Age_at_birth", ]

# Extract slope + 95% CI for Trio Indels
slope_indel <- coef(lm_indel_trio)["Pat_Age_at_birth"]
ci_indel <- confint(lm_indel_trio)["Pat_Age_at_birth", ]

# Print summary
cat("=== Trio DNMs (SNVs) ===\n")
cat("Slope:", signif(slope_snv, 3), "\n")
cat("95% CI:", signif(ci_snv[1], 3), "to", signif(ci_snv[2], 3), "\n\n")

cat("=== Trio DNMs (Indels) ===\n")
cat("Slope:", signif(slope_indel, 3), "\n")
cat("95% CI:", signif(ci_indel[1], 3), "to", signif(ci_indel[2], 3), "\n")

# Fit interaction model for SNVs
lm_snv_interaction <- lm(DNM ~ Pat_Age_at_birth * dataset, data = dnm_data_long |> filter(class == "snv"))


# Fit interaction model for Indels
lm_indel_interaction <- lm(DNM ~ Pat_Age_at_birth * dataset, data = dnm_data_long |> filter(class == "indel"))


# Extract p-values for interaction term (slope difference)
pval_snv_interaction <- summary(lm_snv_interaction)$coefficients["Pat_Age_at_birth:datasetTrio DNMs", "Pr(>|t|)"]
pval_indel_interaction <- summary(lm_indel_interaction)$coefficients["Pat_Age_at_birth:datasetTrio DNMs", "Pr(>|t|)"]

cat("\n=== Interaction test ===\n")
cat("P-value for SNV slope difference (Trio vs Control):", signif(pval_snv_interaction, 3), "\n")
cat("P-value for Indel slope difference (Trio vs Control):", signif(pval_indel_interaction, 3), "\n")

# SNV panel using model predictions
snv_DNM_plot <- ggplot(trioBurdens |> bind_rows(trioControlBurdens) |> mutate(dataset = factor(dataset, levels = c("Trio DNMs", "Controls"))), 
                       aes(x = Pat_Age_at_birth, y = SNP_DNMs_0.3VAF, fill = dataset)) +
  geom_point(pch = 21, size = 3.5, stroke = 0.25, colour = "white") +
  geom_ribbon(
    data = all_preds |> filter(class == "snv"),
    aes(x = Pat_Age_at_birth, ymin = ci_lo, ymax = ci_hi, fill = dataset),
    inherit.aes = FALSE, alpha = 0.25
  ) +
  geom_line(
    data = all_preds |> filter(class == "snv"),
    aes(x = Pat_Age_at_birth, y = fit, colour = dataset),
    inherit.aes = FALSE, linewidth = 1.1
  ) +
  xlim(c(20, 75)) +
  ylim(c(0, NA)) +
  scale_fill_manual(values = dataset_colours, guide = guide_legend(title = NULL)) +
  scale_colour_manual(values = dataset_colours, guide = guide_legend(title = NULL)) +
  xlab("Paternal age") +
  ylab("De novo SNV count") +
  theme_pubr() +
  theme(
    text = element_text(size = 16),
    legend.position = c(0.4, 0.4),
    legend.justification = c("left", "top"),
    legend.background = element_blank(),
    legend.key.height = unit(1.2, "lines"),
    legend.key.width = unit(1.2, "lines")
  )

# Indel panel using model predictions
indel_DNM_plot <- ggplot(trioBurdens |> bind_rows(trioControlBurdens) |> mutate(dataset = factor(dataset, levels = c("Trio DNMs", "Controls"))), 
                         aes(x = Pat_Age_at_birth, y = Indel_DNMs_0.3VAF, fill = dataset)) +
  geom_point(pch = 21, size = 3.5, stroke = 0.25, colour = "white") +
  geom_ribbon(
    data = all_preds |> filter(class == "indel"),
    aes(x = Pat_Age_at_birth, ymin = ci_lo, ymax = ci_hi, fill = dataset),
    inherit.aes = FALSE, alpha = 0.25
  ) +
  geom_line(
    data = all_preds |> filter(class == "indel"),
    aes(x = Pat_Age_at_birth, y = fit, colour = dataset),
    inherit.aes = FALSE, linewidth = 1.1
  ) +
  xlim(c(20, 75)) +
  ylim(c(0, NA)) +
  scale_fill_manual(values = dataset_colours, guide = guide_legend(title = NULL)) +
  scale_colour_manual(values = dataset_colours, guide = guide_legend(title = NULL)) +
  xlab("Paternal age") +
  ylab("De novo indel count") +
  theme_pubr() +
  theme(
    text = element_text(size = 16),
    legend.position = "none",
    legend.justification = c("left", "top"),
    legend.background = element_blank(),
    legend.key.height = unit(1.2, "lines"),
    legend.key.width = unit(1.2, "lines")
  )

# Combine plots
snv_DNM_plot + indel_DNM_plot
```

### Sperm Burdens 
#### i) Prep mutations
```{r targBurden, fig.height=5, fig.width=8}

# --- 1. Mutations ---
spermVariants1 <- spermVariants |>
  filter(BAM_VAF < 0.01, DUPLEX_VAF < 0.1)

# TwinsUK sperm NanoSeq data (individual-level mutation calls, coverage and
# burdens from the TwinsUK sperm control cohort) is under TwinsUK's
# controlled-access policy - see README. When twinsUKDataAvailable, mutRegionCovs
# below includes the TwinsUK "Controls" group exactly as in the manuscript.
# Otherwise it's DD fathers + Apert fathers only; the TwinsUK trend line is added
# back into the model predictions (not the raw points) further down from a
# pre-computed aggregate fit.
if (twinsUKDataAvailable) {
  twinsUKTargVars1 <- read_tsv(
    paste0(dataPath, "twinsUK_analysisTargVars.tsv"),
    col_types = cols()
  ) |>
    filter(nanoseq == "Exome") |>
    select(-meth)
}

# Mutations per region
germanMuts <- bind_rows(
  spermVariants1 |> mutate(region = "all"),
  spermVariants1 |> mutate(region = "germanPanel"),
  spermVariants1 |>
    left_join(
      read_tsv(paste0(dataPath, "exclDrivers.txt"),
               col_types = 'cdc') |> distinct(),
      by = join_by(chr, start)
    ) |> drop_na(region)
) |>
  mutate(dataset = "DD fathers")

if (twinsUKDataAvailable) {
  twinsMuts <- bind_rows(
    twinsUKTargVars1 |> filter(BAM_VAF < 0.01) |> mutate(region = "all"),
    twinsUKTargVars1 |>
      left_join(
        bind_rows(
          read_tsv(paste0(dataPath, "twinsUK_coding.txt"),
                   col_types = cols(), col_names = c("chr", "start", "region")),
          read_tsv(paste0(dataPath, "twinsUK_nonCoding.txt"),
                   col_types = cols(), col_names = c("chr", "start", "region"))
        ) |> mutate(region = "germanPanel") |> distinct(),
        by = join_by(chr, start)
      ) |> drop_na(region),
    twinsUKTargVars1 |>
      left_join(
        read_tsv(paste0(dataPath, "twinsUK_exclDrivers.txt"),
                 col_types = cols(), col_names = c("chr", "start", "region")) |> distinct(),
        by = join_by(chr, start)
      ) |> drop_na(region)
  ) |>
    mutate(dataset = "Controls")
}

mutRegionMuts <- (if (twinsUKDataAvailable) bind_rows(germanMuts, twinsMuts) else germanMuts) |>
  group_by(Sample_PD_ID, class, region, dataset) |>
  summarise(muts = sum(TIMES_CALLED), .groups = "drop") |>
  complete(nesting(Sample_PD_ID, dataset), class, region, fill = list(muts = 0))

# --- 2. Coverage ---
regions <- c("all", "germanPanel", "exclDrivers")

germanCov <- bind_rows(
  read_tsv(paste0(dataPath, "mutRegionsCovs.tsv"),
           col_types = cols()),
  spermSamplesFiltered |> select(Sample_PD_ID, total = total_duplex_bp_cov) |> mutate(region = "all"),
  spermSamplesFiltered |> select(Sample_PD_ID, total = total_duplex_bp_cov) |> mutate(region = "germanPanel")
)

if (twinsUKDataAvailable) {
  twinsCov_raw <- read_tsv(
    paste0(dataPath, "twinsUK_mutRegionsCovs.tsv"),
    col_types = cols(), col_names = c("total", "Sample_PD_ID", "region")
  )

  twinsCov <- bind_rows(
    twinsCov_raw,
    twinsCov_raw |>
      filter(region %in% c("coding", "nonCoding")) |>
      group_by(Sample_PD_ID) |>
      summarise(total = sum(total), .groups = "drop") |>
      mutate(region = "germanPanel")
  )

  burdenAll <- read_tsv(
    paste0(dataPath, "twinsUK_burdensFiltered_exome.tsv"),
    col_types = cols()
  ) |>
    select(Sample_PD_ID, total) |>
    mutate(region = "all")
}

mutRegionCovs_base <- (if (twinsUKDataAvailable) bind_rows(germanCov, twinsCov, burdenAll) else germanCov) |>
  left_join(mutRegionMuts, by = join_by(Sample_PD_ID, region)) |>
  drop_na() |>
  mutate(burden = muts / total)


# --- 3. Add metadata ---
# TwinsUK sample metadata (age at sampling) is part of the same controlled-access
# dataset - see note above. sampleMeta below is DD fathers + Apert fathers only
# unless twinsUKDataAvailable.
sampleMeta <- spermSamplesFiltered |> select(Sample_PD_ID, Age)
if (twinsUKDataAvailable) {
  sampleMeta <- bind_rows(
    sampleMeta,
    read_tsv(paste0(dataPath, "targetedSamplesFiltered.tsv"), col_types = cols()) |>
      select(Sample_PD_ID, Age = age_at_sampling)
  )
}

mutRegionCovs <- mutRegionCovs_base |>
  left_join(sampleMeta, by = "Sample_PD_ID") |>
  mutate(region = fct_relevel(region, regions)) |>
  mutate(dataset = if_else(str_detect(Sample_PD_ID, "PD59"), "Apert fathers", dataset))

```

#### ii) Linear models
```{r targBurdenModels, fig.height=5, fig.width=8}
# -- Label/summary helpers for lm() models --

metricsText_lm <- function(model, roundTo = 2) {
  slope <- coef(model)[2]
  intercept <- coef(model)[1]
  conf <- confint(model)
  cat(paste0("Slope of ", round(slope, roundTo), " * Age (95% CI: ",
             round(conf[2,1], roundTo), "–", round(conf[2,2], roundTo), ")\n"))
  cat(paste0("Intercept of ", round(intercept, roundTo), " (95% CI: ",
             round(conf[1,1], roundTo), "–", round(conf[1,2], roundTo), ")\n"))
}

metricsLabel_lm <- function(model, roundTo = 2) {
  sprintf("%0.0f + %.*f × age", coef(model)[1], roundTo, coef(model)[2])
}

# -- Create grid of ages for prediction --
ages_grid <- tibble(Age = seq(20, 75, by = 5))


# -- Fit a model and return predictions with CI --
fit_model_and_predict <- function(df, ages = ages_grid$Age) {
  model <- lm(burden ~ Age, data = df)
  
  pred_mat <- sapply(ages, function(a) {
    predict(model, newdata = tibble(Age = a), se.fit = TRUE, interval = "confidence", level = 0.95)$fit[1:3]
  })
  
  pred_df <- tibble(
    Age = ages,
    fit = pred_mat[1,],
    ci_lo = pred_mat[2,],
    ci_hi = pred_mat[3,]
  )
  
  list(model = model, predictions = pred_df)
}


# -- Filter and prepare input data --
mutData <- mutRegionCovs |> 
  filter(region != "all") |> 
  mutate(region = as.character(region))


# -- Run models and collect predictions + metrics --
model_outputs <- mutData |> 
  group_by(class, region, dataset) |> 
  group_map(~ {
    result <- fit_model_and_predict(.x)
    label <- metricsLabel_lm(result$model)
    
    predictions <- result$predictions |> mutate(
      class = .y$class,
      region = .y$region,
      dataset = .y$dataset,
      label = label
    )
    
    metrics <- tibble(
      class = .y$class,
      region = .y$region,
      dataset = .y$dataset,
      label = label,
      slope = coef(result$model)[2],
      intercept = coef(result$model)[1]
    )
    
    list(model = result$model, predictions = predictions, metrics = metrics)
  })


# -- Combine outputs for plotting and summary --
all_preds <- map_dfr(model_outputs, ~ .x$predictions)
if (!twinsUKDataAvailable) {
  # Add the TwinsUK "Controls" trend line back in from the pre-computed
  # aggregate fit (data/twinsUK_mutRegionFits.tsv - Age grid + fit + CI per
  # class/region, no individual TwinsUK points) so plots still show it below.
  all_preds <- bind_rows(all_preds, read_tsv(paste0(dataPath, "twinsUK_mutRegionFits.tsv"), col_types = cols()))
}
all_preds <- all_preds |>
    mutate(dataset = factor(dataset, levels = c("DD fathers", "Apert fathers", "Controls")))
all_metrics <- map_dfr(model_outputs, ~ .x$metrics)

```

#### iii) Test burden dif
```{r burdenDif, fig.height=5, fig.width=8}
compare_slopes_verbose <- function(df, mutation_class) {
  # Subset by mutation class and region
  subset_data <- df |> filter(class == mutation_class, region == "germanPanel")
  
  # Fit models
  full_model <- lm(burden ~ Age * dataset, data = subset_data)
  null_model <- lm(burden ~ Age + dataset, data = subset_data)
  
  # Compare models
  anova_result <- anova(null_model, full_model)
  interaction_p <- anova_result$`Pr(>F)`[2]
  
  # Extract joint slope + CI from null model
  slope <- coef(null_model)["Age"]
  ci <- confint(null_model)["Age", ]
  
  # Format outputs
  cat("\n=== Results for", toupper(mutation_class), "(germanPanel) ===\n")
  cat("P-value for Age × dataset interaction:", signif(interaction_p, 3), "\n")
  if (!is.na(interaction_p) && interaction_p < 0.05) {
    cat("⟶ Slopes differ significantly between datasets.\n")
  } else {
    cat("⟶ No significant difference in slope between datasets.\n")
  }
  cat("Joint slope estimate:", signif(slope, 3), "mutations per year\n")
  cat("95% CI for joint slope: [", signif(ci[1], 3), "–", signif(ci[2], 3), "]\n\n")
  
  # Return values (optional)
  return(invisible(list(
    p_value = interaction_p,
    slope = slope,
    ci_lower = ci[1],
    ci_upper = ci[2]
  )))
}
compare_slopes_verbose(mutRegionCovs, "snv")
compare_slopes_verbose(mutRegionCovs, "indel")

```

#### iv) Plots
```{r plotBurdens, fig.height=4, fig.width=12}

# Points come from mutRegionCovs (real data only, so no individual TwinsUK
# points unless twinsUKDataAvailable); the ribbon/line come from all_preds,
# which always includes the TwinsUK "Controls" trend - either fitted live or
# read back from the pre-computed aggregate fit (see chunks above).
pointDatasetLevels <- if (twinsUKDataAvailable) c("DD fathers", "Controls") else c("DD fathers")
pointDatasetLevelsApert <- if (twinsUKDataAvailable) c("DD fathers", "Apert fathers", "Controls") else c("DD fathers", "Apert fathers")

# SNV panel
snv_plot <- ggplot(
  mutRegionCovs |> filter(class == "snv", region == "germanPanel", dataset != "Apert fathers") |>
    mutate(dataset = factor(dataset, levels = pointDatasetLevels)),
  aes(x = Age, y = burden, fill = dataset)
) +
  geom_point(pch = 21, size = 3.5, stroke = 0.25, colour = "white") +
  geom_ribbon(
    data = all_preds |> filter(region == "germanPanel" & class == "snv", dataset != "Apert fathers"),
    aes(x = Age, ymin = ci_lo, ymax = ci_hi, fill = dataset),
    inherit.aes = FALSE, alpha = 0.25
  ) +
  geom_line(
    data = all_preds |> filter(region == "germanPanel" & class == "snv", dataset != "Apert fathers"),
    aes(x = Age, y = fit, colour = dataset),
    inherit.aes = FALSE, linewidth = 1.1
  ) +
  ylim(0,NA) +
  scale_fill_manual(values = dataset_colours, guide = guide_legend(title = NULL)) +
  scale_colour_manual(values = dataset_colours, guide = guide_legend(title = NULL)) +
  xlab("Paternal age") +
  ylab("Sperm SNVs per bp") +
  theme_pubr() +
  theme(
    text = element_text(size = 16),
    legend.position = c(0.02, 0.98),
    legend.justification = c("left", "top"),
    legend.background = element_blank(),
    legend.key.height = unit(1.2, "lines"),
    legend.key.width = unit(1.2, "lines")
  )

# Indel panel
indel_plot <- ggplot(
  mutRegionCovs |> filter(class == "indel", region == "germanPanel", dataset != "Apert fathers")|>
    mutate(dataset = factor(dataset, levels = pointDatasetLevels)),
  aes(x = Age, y = burden, fill = dataset)
) +
  geom_point(pch = 21, size = 3.5, stroke = 0.25, colour = "white") +
  geom_ribbon(
    data = all_preds |> filter(region == "germanPanel" & class == "indel", dataset != "Apert fathers"),
    aes(x = Age, ymin = ci_lo, ymax = ci_hi, fill = dataset),
    inherit.aes = FALSE, alpha = 0.25
  ) +
  geom_line(
    data = all_preds |> filter(region == "germanPanel" & class == "indel", dataset != "Apert fathers"),
    aes(x = Age, y = fit, colour = dataset),
    inherit.aes = FALSE, linewidth = 1.1
  ) +
  scale_fill_manual(values = dataset_colours, guide = guide_legend(title = NULL)) +
  scale_colour_manual(values = dataset_colours, guide = guide_legend(title = NULL)) +
  xlab("Paternal age") +
  ylab("Sperm indels per bp") +
  theme_pubr() +
  theme(
    text = element_text(size = 16),
    legend.position = "none"
  )

snv_plot + indel_plot

# SNV panel (DD fathers vs Apert fathers, +TwinsUK Controls if available)
snv_plotApert <- ggplot(
  mutRegionCovs |> filter(class == "snv", region == "germanPanel") |>
    mutate(dataset = factor(dataset, levels = pointDatasetLevelsApert)),
  aes(x = Age, y = burden, fill = dataset)
) +
  geom_point(pch = 21, size = 3.5, stroke = 0.25, colour = "white") +
  geom_ribbon(
    data = all_preds |> filter(region == "germanPanel" & class == "snv"),
    aes(x = Age, ymin = ci_lo, ymax = ci_hi, fill = dataset),
    inherit.aes = FALSE, alpha = 0.25
  ) +
  geom_line(
    data = all_preds |> filter(region == "germanPanel" & class == "snv"),
    aes(x = Age, y = fit, colour = dataset),
    inherit.aes = FALSE, linewidth = 1.1
  ) +
  ylim(0,NA) +
  scale_fill_manual(values = dataset_colours, guide = guide_legend(title = NULL)) +
  scale_colour_manual(values = dataset_colours, guide = guide_legend(title = NULL)) +
  xlab("Paternal age") +
  ylab("Sperm SNVs per bp") +
  theme_pubr() +
  theme(
    text = element_text(size = 16),
    legend.position = c(0.02, 0.98),
    legend.justification = c("left", "top"),
    legend.background = element_blank(),
    legend.key.height = unit(1.2, "lines"),
    legend.key.width = unit(1.2, "lines")
  )

# Indel panel (DD fathers vs Apert fathers, +TwinsUK Controls if available)
indel_plotApert <- ggplot(
  mutRegionCovs |> filter(class == "indel", region == "germanPanel")|>
    mutate(dataset = factor(dataset, levels = pointDatasetLevelsApert)),
  aes(x = Age, y = burden, fill = dataset)
) +
  geom_point(pch = 21, size = 3.5, stroke = 0.25, colour = "white") +
  geom_ribbon(
    data = all_preds |> filter(region == "germanPanel" & class == "indel"),
    aes(x = Age, ymin = ci_lo, ymax = ci_hi, fill = dataset),
    inherit.aes = FALSE, alpha = 0.25
  ) +
  geom_line(
    data = all_preds |> filter(region == "germanPanel" & class == "indel"),
    aes(x = Age, y = fit, colour = dataset),
    inherit.aes = FALSE, linewidth = 1.1
  ) +
  scale_fill_manual(values = dataset_colours, guide = guide_legend(title = NULL)) +
  scale_colour_manual(values = dataset_colours, guide = guide_legend(title = NULL)) +
  xlab("Paternal age") +
  ylab("Sperm indels per bp") +
  theme_pubr() +
  theme(
    text = element_text(size = 16),
    legend.position = "none"
  )

snv_plotApert + indel_plotApert
```

### Parental Phase test
```{r phaseBurden, fig.height=5, fig.width=10}

distinctPhasing <- diagnosticVarsCalls |> 
  select(FAM_ID, phasing) |> 
  distinct() |> 
  group_by(FAM_ID) |> 
  summarise(
    phasing = case_when(
      all(phasing %in% c("paternal", "unknown")) & "paternal" %in% phasing ~ "paternal",
      all(phasing %in% c("maternal", "unknown")) & "maternal" %in% phasing ~ "maternal",
      TRUE ~ "unknown"
    ),
    .groups = "drop"
  )
table(distinctPhasing$phasing)

burdenPhasing <- spermSamplesFiltered |>
  mutate(burdenCorr = snv_burden_corr_0.01vaf, indelBurden = indel_burden_0.01vaf)|>
  left_join(distinctPhasing, by = join_by(FAM_ID)) |>
  mutate(phasing = coalesce(phasing, "unknown"))
table(burdenPhasing$phasing)

# 1. Filter to samples with phasing: paternal or maternal only
phasedOnly <- burdenPhasing |> 
  filter(phasing %in% c("paternal", "maternal")) |> 
  mutate(phasing = factor(phasing, levels = c("maternal", "paternal")))

# 2. Linear model for SNV burden
lm_snv_phase <- lm(burdenCorr ~ Age + phasing, data = phasedOnly)
summary(lm_snv_phase)

# 3. Linear model for indel burden
lm_indel_phase <- lm(indelBurden ~ Age + phasing, data = phasedOnly)
summary(lm_indel_phase)

# 4. Extract and report p-values for the phasing term
pval_snv_phase <- summary(lm_snv_phase)$coefficients["phasingpaternal", "Pr(>|t|)"]
pval_indel_phase <- summary(lm_indel_phase)$coefficients["phasingpaternal", "Pr(>|t|)"]

cat("P-value for phasing effect (SNV burden):", signif(pval_snv_phase, 3), "\n")
cat("P-value for phasing effect (Indel burden):", signif(pval_indel_phase, 3), "\n")

p_germanTrioPhasingBurdens <- ggplot(burdenPhasing, aes(x = Age, y = burdenCorr, fill = phasing)) + 
  geom_point(size=3.5, pch=21, col='white', stroke=0.25) + 
  geom_smooth(method = "lm") + 
   theme(text = element_text(size=16)) + 
  xlab('Age (years)') + 
  ylab('SNVs per bp') + 
  labs(fill='Phasing of\ndiagnostic variant') + 
  theme_pubr() + 
  ylim(0,NA) +
  theme(legend.position = "right")

p_germanTrioPhasingBurdens

p_indelPhasingBurden<- ggplot(burdenPhasing, aes(x = Age, y = indelBurden, fill =  phasing)) + 
  geom_point(size=3.5, pch=21, col='white', stroke=0.25) + 
  scale_alpha_manual(values = c(1, 1, 0.8), guide = "none") + 
   theme(text = element_text(size=16)) + 
  geom_smooth(method = "lm") + 
  xlab('Age (years)') + 
  ylab('Indels per bp') + 
  labs(fill='Phasing of\ndiagnostic variant') + 
  theme_pubr() + 
  ylim(0,NA)  +
  theme(legend.position = "none")

p_indelPhasingBurden


```

#### SupFig 2
```{r supFig2, fig.height=4, fig.width=11}

p_germanTrioPhasingBurdens + p_indelPhasingBurden + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")  &
  theme(plot.tag = element_text(size = 14, face = "bold"))


pdf(paste0(outputPath, 'SupFig2_phasedBurdens.pdf'), height=4, width=11, onefile = F)
p_germanTrioPhasingBurdens + p_indelPhasingBurden + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")  &
  theme(plot.tag = element_text(size = 14, face = "bold"))
dev.off()

```

### DNM vs Sperm burdens
```{r dnmBurden, fig.height=5, fig.width=8}

burdenComp <- spermSamplesFiltered |> select(FAM_ID, burdenCorr = snv_burden_corr_0.01vaf, indelBurden = indel_burden_0.01vaf, spermAge = Age) |>
  left_join(trioBurdens |> select(FAM_ID, indelDNMs = Indel_DNMs_0.3VAF, snvDNMs = SNP_DNMs_0.3VAF, Pat_Age_at_birth, Mat_Age_at_birth, sample.median.0.3vaf, Proband_seq_x = MEDIAN_COVERAGE, Pat_seq_x, Mat_seq_x), by = join_by(FAM_ID)) |> 
  drop_na(snvDNMs) |> 
  mutate(paternalAgeGap = spermAge - Pat_Age_at_birth)


# Drop rows with missing QC data
burdenComp <- burdenComp |> drop_na(sample.median.0.3vaf, Proband_seq_x, Pat_seq_x, Mat_seq_x)

# === SNV ANALYSIS ===

# 1. Raw correlation
cor_test_snv <- cor.test(burdenComp$burdenCorr, burdenComp$snvDNMs)

# 2. Base model: parental age + QC
lm_base_snv <- lm(snvDNMs ~ Pat_Age_at_birth + Mat_Age_at_birth +
                    sample.median.0.3vaf + Proband_seq_x + Pat_seq_x + Mat_seq_x,
                  data = burdenComp)

# 3. Full model: add sperm burden + age gap
lm_full_snv <- lm(snvDNMs ~ Pat_Age_at_birth + Mat_Age_at_birth + paternalAgeGap +
                    sample.median.0.3vaf + Proband_seq_x + Pat_seq_x + Mat_seq_x +
                    burdenCorr,
                  data = burdenComp)

# 4. Extract R² and p-values
r2_base_snv <- summary(lm_base_snv)$r.squared
r2_full_snv <- summary(lm_full_snv)$r.squared
delta_r2_snv <- r2_full_snv - r2_base_snv

pval_burden_snv <- summary(lm_full_snv)$coefficients["burdenCorr", "Pr(>|t|)"]
pval_anova_snv <- anova(lm_base_snv, lm_full_snv)$`Pr(>F)`[2]

# === INDEL ANALYSIS ===

# 1. Raw correlation
cor_test_indel <- cor.test(burdenComp$indelBurden, burdenComp$indelDNMs)

# 2. Base model
lm_base_indel <- lm(indelDNMs ~ Pat_Age_at_birth + Mat_Age_at_birth +
                      sample.median.0.3vaf + Proband_seq_x + Pat_seq_x + Mat_seq_x,
                    data = burdenComp)

# 3. Full model
lm_full_indel <- lm(indelDNMs ~ Pat_Age_at_birth + Mat_Age_at_birth + paternalAgeGap +
                      sample.median.0.3vaf + Proband_seq_x + Pat_seq_x + Mat_seq_x +
                      indelBurden,
                    data = burdenComp)

# 4. Extract R² and p-values
r2_base_indel <- summary(lm_base_indel)$r.squared
r2_full_indel <- summary(lm_full_indel)$r.squared
delta_r2_indel <- r2_full_indel - r2_base_indel

pval_burden_indel <- summary(lm_full_indel)$coefficients["indelBurden", "Pr(>|t|)"]
pval_anova_indel <- anova(lm_base_indel, lm_full_indel)$`Pr(>F)`[2]

# === REPORT ===

cat("=== SNV Model ===\n")
cat("Raw correlation (burdenCorr vs snvDNMs): p =", signif(cor_test_snv$p.value, 3), "\n")
cat("P-value for sperm burden in full model:", signif(pval_burden_snv, 3), "\n")
cat("Model improvement (ANOVA): p =", signif(pval_anova_snv, 3), "\n")
cat("R² (base model):", round(r2_base_snv, 4), "\n")
cat("R² (with sperm burden):", round(r2_full_snv, 4), "\n")
cat("Additional variance explained:", round(delta_r2_snv * 100, 2), "%\n\n")

cat("=== INDEL Model ===\n")
cat("Raw correlation (indelBurden vs indelDNMs): p =", signif(cor_test_indel$p.value, 3), "\n")
cat("P-value for sperm burden in full model:", signif(pval_burden_indel, 3), "\n")
cat("Model improvement (ANOVA): p =", signif(pval_anova_indel, 3), "\n")
cat("R² (base model):", round(r2_base_indel, 4), "\n")
cat("R² (with sperm burden):", round(r2_full_indel, 4), "\n")
cat("Additional variance explained:", round(delta_r2_indel * 100, 2), "%\n")

ggplot(burdenComp, aes(x = burdenCorr, y = snvDNMs)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = "#1b9e77") +
  labs(x = "Sperm SNV burden", y = "Offspring SNV DNM count") +
  theme_pubr()

burdenComp$pred_snv <- predict(lm_full_snv)

ggplot(burdenComp, aes(x = pred_snv, y = snvDNMs)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "Predicted SNV DNM count", y = "Observed SNV DNM count") +
  theme_pubr()

r2_df <- tibble(
  model = c("Base (age + QC)", "Full (+ sperm burden)"),
  r2 = c(r2_base_snv, r2_full_snv)
)

ggplot(r2_df, aes(x = model, y = r2)) +
  geom_col(fill = "#1b9e77", width = 0.6) +
  ylim(0, 1) +
  labs(y = "R² (variance explained)", x = NULL) +
  theme_pubr()

ggplot(burdenComp, aes(x = pred_snv, y = residuals(lm_full_snv))) +
  geom_point(alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = "Predicted DNM count", y = "Residuals") +
  theme_pubr()

```


### Signatures
```{r sigs, fig.width=6, fig.height=5}
trioSigsProp <- trioSigs |> 
  mutate(SBS1 = trioSigs$SBS1/(trioSigs$SBS1 + trioSigs$SBS5)) |> 
  mutate(SBS5 = trioSigs$SBS5/(trioSigs$SBS1 + trioSigs$SBS5))

sigs <- read_tsv(paste0(dataPath, "sigs.txt"), col_types = cols()) |> 
  pivot_longer(cols = c(SBS1, SBS5), names_to = "sig", values_to = "sigCount") |> 
  group_by(Samples) |> 
  mutate(sigCount = sigCount / sum(sigCount)) |> 
  ungroup() |> 
  mutate(sig = fct_relevel(sig, "SBS5", "SBS1")) |> 
  filter(Samples == "GermanTrio") |> 
  mutate(sigCountTrioDNM = c(trioSigsProp$SBS1, trioSigsProp$SBS5))


p_sigs <- ggplot(sigs, aes(x = Samples, y = sigCount, fill = sig)) +
  geom_bar(stat = "identity") +
  ylab("Signature contribution sperm") +
  scale_y_continuous(expand = c(0, 0)) +
  theme_pubr() +
  scale_fill_manual(values = c("SBS5" = "#80b1d3", "SBS1" = "#fdb462")) +
  theme(
    legend.title = element_blank(),
    legend.position = "none",
    strip.background = element_blank(),
    strip.text = element_text(size = 12),
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.line.x = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
p_sigs

p_DNMsigs <- ggplot(sigs, aes(x = Samples, y = sigCountTrioDNM, fill = sig)) +
  geom_bar(stat = "identity") +
  ylab("Signature contribution trio DNMs") +
  scale_y_continuous(expand = c(0, 0)) +
  theme_pubr() +
  scale_fill_manual(values = c("SBS5" = "#80b1d3", "SBS1" = "#fdb462")) +
  theme(
    legend.title = element_blank(),
    legend.position = "none",
    strip.background = element_blank(),
    strip.text = element_text(size = 12),
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.line.x = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
p_sigs + p_DNMsigs

p_sigs_lbl <- ggplot(sigs, aes(x = Samples, y = sigCount, fill = sig)) +
  geom_bar(stat = "identity") +
  geom_label(aes(label = sig),
             position = position_stack(vjust = 0.5),
             size = 3,
             fill = "white",       # label background for contrast
             alpha = 0.9,          # slight transparency
             label.size = 0) +     # no border
  ylab("Signature contribution sperm") +
  scale_y_continuous(expand = c(0, 0)) +
  theme_pubr() +
  scale_fill_manual(values = c("SBS5" = "#80b1d3", "SBS1" = "#fdb462")) +
  theme(
    legend.title = element_blank(),
    legend.position = "none",
    strip.background = element_blank(),
    strip.text = element_text(size = 12),
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.line.x = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

p_DNMsigs_lbl <- ggplot(sigs, aes(x = Samples, y = sigCountTrioDNM, fill = sig)) +
  geom_bar(stat = "identity") +
  geom_label(aes(label = sig),
             position = position_stack(vjust = 0.5),
             size = 3,
             fill = "white",
             alpha = 0.9,
             label.size = 0) +
  ylab("Signature contribution trio DNMs") +
  scale_y_continuous(expand = c(0, 0)) +
  theme_pubr() +
  scale_fill_manual(values = c("SBS5" = "#80b1d3", "SBS1" = "#fdb462")) +
  theme(
    legend.title = element_blank(),
    legend.position = "none",
    strip.background = element_blank(),
    strip.text = element_text(size = 12),
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.line.x = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

p_sigs_lbl + p_DNMsigs_lbl
```

### Trinuc profiles
```{r trinuc, fig.height=5, fig.width=5}

# Create the necessary vectors
sub_vec <- c("C>A", "C>G", "C>T", "T>A", "T>C", "T>G")
ctx_vec <- paste(rep(c("A", "C", "G", "T"), each = 4), rep(c("A", "C", "G", "T"), times = 4), sep = "-")
full_vec <- paste(rep(sub_vec, each = 16), rep(ctx_vec, times = 6), sep = ",")
xstr <- paste(substr(full_vec, 5, 5), substr(full_vec, 1, 1), substr(full_vec, 7, 7), sep = "")
ordered_names <- paste(xstr, ">", rep(c("A", "G", "T", "A", "C", "G"), each = 16), sep = "")

# Define colors for each sub_vec
colours <- c("deepskyblue", "black", "firebrick2", "gray", "darkolivegreen3", "rosybrown2")
names(colours) <- sub_vec  # Assign names to colors for mapping

# Create a combined data frame
trinucCounts <- tibble(
  TRI = ordered_names,
  xstr = xstr,
  sub_vec = rep(sub_vec, each = 16)) |> 
  left_join(spermVariants |> filter(str_detect(Sample_PD_ID, "PD6")) |>  group_by(TRI) |> summarize(mutation_counts = n(), .groups =  "drop") |> 
  drop_na(), by = join_by(TRI)) |> 
  left_join(spermVariants |> filter(str_detect(Sample_PD_ID, "PD5")) |>  group_by(TRI) |> summarize(apertFather_counts = n(), .groups =  "drop") |> 
  drop_na(), by = join_by(TRI)) |> 
  mutate(apertFather_counts = coalesce(apertFather_counts, 0)) |> 
  left_join(trinucDNMs, by = join_by(TRI))

trinucSperm <- ggplot(trinucCounts, aes(x = xstr, y = mutation_counts, fill = sub_vec)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.8) +
  facet_wrap(~sub_vec, ncol = 6, scales = "free_x") +
  scale_fill_manual(values = colours) +
  scale_y_continuous(limits=c(0, NA), expand = c(0, 0)) +
  labs(y = "Mutation counts", x = "Sperm trinucleotide mutations") +
  theme_pubr() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.line.x = element_blank(),
    legend.position = "none",
    strip.text = element_blank(),
    panel.spacing = unit(0, "lines"),
    panel.grid = element_blank()
  ) 

trinucApert <- ggplot(trinucCounts, aes(x = xstr, y = apertFather_counts, fill = sub_vec)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.8) +
  facet_wrap(~sub_vec, ncol = 6, scales = "free_x") +
  scale_fill_manual(values = colours) +
  scale_y_continuous(limits=c(0, NA), expand = c(0, 0)) +
  labs(y = "Mutation counts", x = "Apert father sperm trinucleotide mutations") +
  theme_pubr() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.line.x = element_blank(),
    legend.position = "none",
    strip.text = element_blank(),
    panel.spacing = unit(0, "lines"),
    panel.grid = element_blank()
  ) 


trinucDNM <- ggplot(trinucCounts, aes(x = xstr, y = dnmCount, fill = sub_vec)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.8) +
  facet_wrap(~sub_vec, ncol = 6, scales = "free_x") +
  scale_fill_manual(values = colours) +
  scale_y_continuous(limits=c(0, NA), expand = c(0, 0)) +
  labs(y = "Mutation counts", x = "Trio trinucleotide DNMs") +
  theme_pubr() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.line.x = element_blank(),
    legend.position = "none",
    strip.text = element_blank(),
    panel.spacing = unit(0, "lines"),
    panel.grid = element_blank()
  ) 


# Data frame for rectangles
domainsPlotObject <- data.frame(
  recStart = 1:6,
  recEnd = 2:7,
  base = 0,
  height = rep(1, 6),
  type = sub_vec)
# Create rectangles using ggplot
pDomains <- ggplot(domainsPlotObject, aes(xmin = recStart, xmax = recEnd, ymin = base, ymax = height)) +
    geom_rect(aes(fill = type), colour = "white") +
    geom_text(aes(x = (recStart + recEnd) / 2, y = height + 0.1, label = type), color = "black", vjust = 0) +  # Adjust y position
    scale_fill_manual(values = colours) +
    theme_void() + 
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0), limits = c(0, max(domainsPlotObject$height) + 0.2)) +  # Adjust y limits
    guides(fill = "none") + coord_cartesian(clip = "off")
plot_spacer()/pDomains/trinucDNM/trinucSperm + plot_layout(heights = c(0.1,1, 15, 15))
# pDomains/trinuc + plot_layout(heights = c(1, 15))

plot_spacer()/pDomains/trinucDNM/trinucSperm/trinucApert + plot_layout(heights = c(0.1,1, 15, 15, 15))

```

##### Figure 1
```{r fig1, fig.width=20, fig.height=8}


p_graphic <- ggplot2::ggplot() + ggplot2::annotation_custom(grid::rasterGrob(readPNG(paste0(dataPath, "GraphicFig1a.png")),
                                                width=ggplot2::unit(1,"npc"),
                                                height=ggplot2::unit(1,"npc"),
                                                interpolate = FALSE),
                               -Inf, Inf, -Inf, Inf) + coord_fixed(ratio = 412/238, expand = FALSE, clip = "off") +
  theme_void() +
  theme(panel.background = element_rect(fill = "white", colour = NA),
        plot.margin = margin(0, 0, 0, 0))

p_graphic
layoutFig1 <- "
ABCDF
ABCEF
AGHIK
AGHJK"

p_graphic  + labs(tag = "A") + snv_DNM_plot  + labs(tag = "B") + indel_DNM_plot  + labs(tag = "C") + pDomains + labs(tag = "D") + trinucDNM + p_DNMsigs_lbl + labs(tag = "E") + snv_plot + labs(tag = "F") + indel_plot + labs(tag = "G") + pDomains + labs(tag = "H") + trinucSperm + p_sigs_lbl + labs(tag = "I") +
   plot_layout(design = layoutFig1, heights = c(1,15,1,15), widths = c(3,2,2,4,0.5)) &
  theme(plot.tag = element_text(size = 14, face = "bold"))

pdf(paste0(outputPath, 'Fig1_mutRates.pdf'), height=8, width=19, onefile = F)
p_graphic  + labs(tag = "A") + snv_DNM_plot  + labs(tag = "B") + indel_DNM_plot  + labs(tag = "C") + pDomains + labs(tag = "D") + trinucDNM + p_DNMsigs_lbl + labs(tag = "E") + snv_plot + labs(tag = "F") + indel_plot + labs(tag = "G") + pDomains + labs(tag = "H") + trinucSperm + p_sigs_lbl + labs(tag = "I") +
   plot_layout(design = layoutFig1, heights = c(1,15,1,15), widths = c(3,2,2,4,0.5)) &
  theme(plot.tag = element_text(size = 14, face = "bold"))
dev.off()
```

### Review Analysis: SCN1A exclusion sensitivity
#### DNM burden: SCN1A exclusion
```{r scn1a_DNMburden, fig.width=8, fig.height=5}
# DD fathers only — controls removed. Rebuild from trioBurdens to retain FAM_ID.
# Black dashed line (manuscript) plotted on top of blue solid (SCN1A excluded)
# by ordering layers: blue first, black second.

dnm_data_scn1a <- trioBurdens |>
  select(dataset, FAM_ID, Pat_Age_at_birth, SNP_DNMs_0.3VAF, Indel_DNMs_0.3VAF) |>
  rename(snv = SNP_DNMs_0.3VAF, indel = Indel_DNMs_0.3VAF) |>
  pivot_longer(cols = c(snv, indel), names_to = "class", values_to = "DNM") |>
  drop_na(DNM) |>
  mutate(retained = if_else(FAM_ID %in% scn1a_fam_ids, "SCN1A excluded", "Retained"))

p_dnmBurden_scn1a <- ggplot(
  dnm_data_scn1a,
  aes(x = Pat_Age_at_birth, y = DNM, fill = retained, colour = retained)
) +
  geom_point(size = 3, pch = 21, col = "white", stroke = 0.25) +
  # SCN1A-excluded fit (blue solid — rendered first, underneath)
  geom_smooth(
    method = "lm", alpha = 0.15, inherit.aes = FALSE,
    aes(x = Pat_Age_at_birth, y = DNM, linetype = "SCN1A excluded"),
    colour = "#2166AC", linewidth = 1,
    data = dnm_data_scn1a |> filter(retained == "Retained")
  ) +
  # All DD fathers fit (black dashed — rendered second, on top)
  geom_smooth(
    method = "lm", alpha = 0.10, inherit.aes = FALSE,
    aes(x = Pat_Age_at_birth, y = DNM, linetype = "All DD fathers (manuscript)"),
    colour = "black", linewidth = 1.2,
    data = dnm_data_scn1a
  ) +
  facet_wrap(~ class, scales = "free_y",
             labeller = labeller(class = c(snv = "SNV DNMs", indel = "Indel DNMs"))) +
  scale_fill_manual(name = "Sample status",
                    values = c("Retained" = "#56B4E9", "SCN1A excluded" = "#E69F00")) +
  scale_colour_manual(name = "Sample status",
                      values = c("Retained" = "#56B4E9", "SCN1A excluded" = "#E69F00")) +
  scale_linetype_manual(name = "Linear fit",
                        values = c("All DD fathers (manuscript)" = "dashed", "SCN1A excluded" = "solid")) +
  xlab("Paternal age at birth") + ylab("De novo mutation count") +
  ylim(0, NA) +
  theme_pubr() +
  theme(legend.position = "right", strip.background = element_blank(),
        strip.text = element_text(size = 11))

p_dnmBurden_scn1a
```

#### Sperm burden: SCN1A exclusion
```{r scn1a_spermBurden, fig.width=8, fig.height=5}
# DD fathers only — controls removed. Pivot to long format for SNV + indel facets.

burdenCombined_scn1a <- spermSamplesFiltered |>
  mutate(retained = if_else(Sample_PD_ID %in% scn1a_sperm_ids, "SCN1A excluded", "Retained")) |>
  select(Sample_PD_ID, Age, retained, snv = snv_burden_corr_0.01vaf, indel = indel_burden_0.01vaf) |>
  pivot_longer(cols = c(snv, indel), names_to = "class", values_to = "burden")

p_spermBurden_scn1a <- ggplot(
  burdenCombined_scn1a,
  aes(x = Age, y = burden, fill = retained, colour = retained)
) +
  geom_point(size = 3.5, pch = 21, col = "white", stroke = 0.25) +
  # SCN1A-excluded fit (blue solid — underneath)
  geom_smooth(
    method = "lm", alpha = 0.15, inherit.aes = FALSE,
    aes(x = Age, y = burden, linetype = "SCN1A excluded"),
    colour = "#2166AC", linewidth = 1,
    data = burdenCombined_scn1a |> filter(retained == "Retained")
  ) +
  # All DD fathers fit (black dashed — on top)
  geom_smooth(
    method = "lm", alpha = 0.10, inherit.aes = FALSE,
    aes(x = Age, y = burden, linetype = "All DD fathers (manuscript)"),
    colour = "black", linewidth = 1.2,
    data = burdenCombined_scn1a
  ) +
  facet_wrap(~ class, scales = "free_y",
             labeller = labeller(class = c(snv = "SNV burden", indel = "Indel burden"))) +
  scale_fill_manual(name = "Sample status",
                    values = c("Retained" = "#56B4E9", "SCN1A excluded" = "#E69F00")) +
  scale_colour_manual(name = "Sample status",
                      values = c("Retained" = "#56B4E9", "SCN1A excluded" = "#E69F00")) +
  scale_linetype_manual(name = "Linear fit",
                        values = c("All DD fathers (manuscript)" = "dashed", "SCN1A excluded" = "solid")) +
  xlab("Age (years)") + ylab("Burden (corrected)") +
  ylim(0, NA) +
  theme_pubr() +
  theme(legend.position = "right", strip.background = element_blank(),
        strip.text = element_text(size = 11))

p_spermBurden_scn1a
```

#### Sperm vs DNM burden: SCN1A exclusion
```{r scn1a_spermVsDNM, fig.width=9, fig.height=5}
# Refit all four models (base + full, SNV + indel) on SCN1A-excluded burdenComp.
# Report change in p-value and R² as a grouped bar chart.

burdenComp_scn1a <- burdenComp |>
  mutate(retained = if_else(FAM_ID %in% scn1a_fam_ids, "SCN1A excluded", "Retained"))

dd_excl <- burdenComp_scn1a |> filter(retained == "Retained")

lm_base_snv_scn1a  <- lm(snvDNMs ~ Pat_Age_at_birth + Mat_Age_at_birth +
                            sample.median.0.3vaf + Proband_seq_x + Pat_seq_x + Mat_seq_x,
                          data = dd_excl)
lm_full_snv_scn1a  <- lm(snvDNMs ~ Pat_Age_at_birth + Mat_Age_at_birth + paternalAgeGap +
                            sample.median.0.3vaf + Proband_seq_x + Pat_seq_x + Mat_seq_x +
                            burdenCorr,
                          data = dd_excl)
lm_base_indel_scn1a <- lm(indelDNMs ~ Pat_Age_at_birth + Mat_Age_at_birth +
                             sample.median.0.3vaf + Proband_seq_x + Pat_seq_x + Mat_seq_x,
                           data = dd_excl)
lm_full_indel_scn1a <- lm(indelDNMs ~ Pat_Age_at_birth + Mat_Age_at_birth + paternalAgeGap +
                             sample.median.0.3vaf + Proband_seq_x + Pat_seq_x + Mat_seq_x +
                             indelBurden,
                           data = dd_excl)

# Compile p-values and R²
r2_summary <- tibble(
  class       = rep(c("SNV", "Indel"), each = 4),
  model_set   = rep(c("Manuscript", "Manuscript", "SCN1A excluded", "SCN1A excluded"), 2),
  model_type  = rep(c("Base model", "Full model\n(+sperm burden)"), 4),
  r2          = c(
    summary(lm_base_snv)$r.squared,   summary(lm_full_snv)$r.squared,
    summary(lm_base_snv_scn1a)$r.squared, summary(lm_full_snv_scn1a)$r.squared,
    summary(lm_base_indel)$r.squared,  summary(lm_full_indel)$r.squared,
    summary(lm_base_indel_scn1a)$r.squared, summary(lm_full_indel_scn1a)$r.squared
  ),
  p_burden    = c(
    NA, summary(lm_full_snv)$coefficients["burdenCorr", "Pr(>|t|)"],
    NA, summary(lm_full_snv_scn1a)$coefficients["burdenCorr", "Pr(>|t|)"],
    NA, summary(lm_full_indel)$coefficients["indelBurden", "Pr(>|t|)"],
    NA, summary(lm_full_indel_scn1a)$coefficients["indelBurden", "Pr(>|t|)"]
  )
) |> mutate(
  class      = factor(class,      levels = c("SNV", "Indel")),
  model_set  = factor(model_set,  levels = c("Manuscript", "SCN1A excluded")),
  model_type = factor(model_type, levels = c("Base model", "Full model\n(+sperm burden)"))
)

cat("=== p-values for sperm burden term ===\n")
r2_summary |> filter(!is.na(p_burden)) |>
  select(class, model_set, p_burden) |>
  mutate(p_burden = signif(p_burden, 3)) |>
  print()

p_r2_scn1a <- ggplot(r2_summary,
       aes(x = model_set, y = r2, fill = model_type)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.65, colour = "black") +
  geom_text(
    data = r2_summary |> filter(!is.na(p_burden)),
    aes(label = paste0("p=", signif(p_burden, 2)), group = model_type),
    position = position_dodge(width = 0.7),
    vjust = -0.4, size = 3
  ) +
  facet_wrap(~ class) +
  scale_fill_manual(name = NULL,
                    values = c("Base model" = "grey70",
                               "Full model\n(+sperm burden)" = "#2166AC")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = expression("Variance explained (" * R^2 * ")")) +
  theme_pubr() +
  theme(strip.background = element_blank(), strip.text = element_text(face = "bold"),
        legend.position = "right")

p_r2_scn1a
```

#### Combined SCN1A sensitivity figure
```{r scn1a_combined, fig.width=10, fig.height=10}

make_dnm_panel_scn1a <- function(cls) {
  dat    <- dnm_data_scn1a |> filter(class == cls)
  ylab_str <- if (cls == "snv") "DNM SNV burden" else "DNM indel burden"
  ggplot(dat, aes(x = Pat_Age_at_birth, y = DNM, fill = retained, colour = retained)) +
    geom_point(size = 3, pch = 21, col = "white", stroke = 0.25) +
    geom_smooth(
      method = "lm", alpha = 0.15, inherit.aes = FALSE,
      aes(x = Pat_Age_at_birth, y = DNM, linetype = "SCN1A excluded"),
      colour = "#2166AC", linewidth = 1,
      data = dat |> filter(retained == "Retained")
    ) +
    geom_smooth(
      method = "lm", alpha = 0.10, inherit.aes = FALSE,
      aes(x = Pat_Age_at_birth, y = DNM, linetype = "All DD fathers (manuscript)"),
      colour = "black", linewidth = 1.2,
      data = dat
    ) +
    scale_fill_manual(name = "Sample status",
                      values = c("Retained" = "#56B4E9", "SCN1A excluded" = "#E69F00")) +
    scale_colour_manual(name = "Sample status",
                        values = c("Retained" = "#56B4E9", "SCN1A excluded" = "#E69F00")) +
    scale_linetype_manual(name = "Linear fit",
                          values = c("All DD fathers (manuscript)" = "dashed",
                                     "SCN1A excluded" = "solid")) +
    xlab("Paternal age at birth") + ylab(ylab_str) +
    ylim(0, NA) +
    theme_pubr() +
    theme(legend.position = "right")
}

make_sperm_panel_scn1a <- function(cls) {
  dat      <- burdenCombined_scn1a |> filter(class == cls)
  ylab_str <- if (cls == "snv") "Sperm SNV burden" else "Sperm indel burden"
  ggplot(dat, aes(x = Age, y = burden, fill = retained, colour = retained)) +
    geom_point(size = 3.5, pch = 21, col = "white", stroke = 0.25) +
    geom_smooth(
      method = "lm", alpha = 0.15, inherit.aes = FALSE,
      aes(x = Age, y = burden, linetype = "SCN1A excluded"),
      colour = "#2166AC", linewidth = 1,
      data = dat |> filter(retained == "Retained")
    ) +
    geom_smooth(
      method = "lm", alpha = 0.10, inherit.aes = FALSE,
      aes(x = Age, y = burden, linetype = "All DD fathers (manuscript)"),
      colour = "black", linewidth = 1.2,
      data = dat
    ) +
    scale_fill_manual(name = "Sample status",
                      values = c("Retained" = "#56B4E9", "SCN1A excluded" = "#E69F00")) +
    scale_colour_manual(name = "Sample status",
                        values = c("Retained" = "#56B4E9", "SCN1A excluded" = "#E69F00")) +
    scale_linetype_manual(name = "Linear fit",
                          values = c("All DD fathers (manuscript)" = "dashed",
                                     "SCN1A excluded" = "solid")) +
    xlab("Age (years)") + ylab(ylab_str) +
    ylim(0, NA) +
    theme_pubr() +
    theme(legend.position = "right")
}

# Legends suppressed on panels a, b, c at call site; panel d keeps its legend (right).
# Panel e keeps its own legend (right). No guide collection used.
no_leg <- guides(fill = "none", colour = "none", linetype = "none")
p_snvDNM_scn1a     <- make_dnm_panel_scn1a("snv")   + no_leg
p_indelDNM_scn1a   <- make_dnm_panel_scn1a("indel") + no_leg
p_snvSperm_scn1a   <- make_sperm_panel_scn1a("snv")   + no_leg
p_indelSperm_scn1a <- make_sperm_panel_scn1a("indel")

rebuttal_fig1 <- (p_snvDNM_scn1a + p_indelDNM_scn1a) /
  (p_snvSperm_scn1a + p_indelSperm_scn1a) /
  p_r2_scn1a +
  plot_layout(heights = c(1, 1, 0.8)) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(size = 14, face = "bold"))

rebuttal_fig1
```

## c) Comp DNMs
### Plot VAFs
```{r plotVAFs}


pDiagVarCov <- ggplot(diagnosticVarsCalls |> filter(duplexCov > 0) |> 
                        mutate(Cohort = factor(Cohort, levels = c("DD fathers", "Apert fathers"))),
       aes(x = Cohort, y = duplexCov, fill = Cohort)) +
  geom_boxplot(outlier.shape = NA, width = 0.3) +
  geom_jitter(width = 0.15, size = 3.5, pch = 21, col = 'white', stroke = 0.25, alpha = 0.7) +
  scale_fill_manual(values = dataset_colours) +
  theme_pubr() +
  ylab("Sperm duplex coverage of causal DNM site") +
  theme(
    axis.title.x = element_blank(), 
    legend.position = "none", 
    legend.title = element_blank()    
  )

pDiagVarCov

median(diagnosticVarsCalls |> filter(Cohort == "DD fathers") |> pull(duplexCov))
median(diagnosticVarsCalls |> filter(Cohort == "Apert fathers") |> pull(duplexCov))

```

### Compare UMI vs Duplex
```{r compareUmiDuplex}
# Build UMI table
umi_tbl <- read_csv(file.path(dataPath, "UMIdiagVarFreqs.csv"), col_types = cols()) |>
  separate(Var, into = c("IND_ID", "VAR_ID", "gene"), sep = "\\|") |>
  mutate(across(c(IND_ID, VAR_ID, gene), str_trim)) |>
  transmute(
    Variant = paste0(VAR_ID, " | ", gene),
    Tissue,
    UMI_Neuser2025 = as.numeric(str_remove(`VAF with UMI collapsing`, "%")) / 100
  )

# Build NanoSeq table (diagnostic calls)
nano_tbl <- diagnosticVarsCalls |>
  filter(varLabel != "") |>
  transmute(
    Variant = paste0(VAR_ID, " | ", gene),
    Tissue = "semen father",
    duplexVAF = duplexVAF,
    BAM_VAF = na_if(BAM_VAF, -0.01)
  )

# Join and reshape; keep UMI and duplex VAFs for plotting
umi_vs_duplex <- full_join(umi_tbl, nano_tbl, by = c("Variant", "Tissue")) |>
  select(Variant, Tissue, UMI_Neuser2025, duplexVAF) |>
  pivot_longer(
    cols = c(UMI_Neuser2025, duplexVAF),
    names_to = "Method",
    values_to = "VAF"
  ) |>
  drop_na(VAF) |>
  mutate(
    SeqType = if_else(Method == "UMI_Neuser2025", "UMI_Neuser2025", "NanoSeq"),
  ) |> 
  bind_rows(tibble(Variant = c("VAR_223 | EZH2", "VAR_249 | SCN1A"),
                   Tissue = "semen father",
                   Method = "Case report",
                   SeqType = "UMI_Neuser2025",
                   VAF = c(0.0311,0.1054561)))

# Plot
pUMIcomp <- umi_vs_duplex |>
  filter(Tissue == "semen father", Variant != "VAR_155 | PRPF38A") |>
  ggplot(aes(x = Variant, y = VAF, fill = SeqType)) +
  geom_point(size = 3.5, pch = 21, colour = "white", stroke = 0.25,
             position = position_dodge(width = 0.6)) +
  theme_pubr() +
  scale_fill_manual(values = c("UMI_Neuser2025" = "#0072B2", "NanoSeq" = "#E69F00")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank(),
        legend.position = c(0.2,0.7))
pUMIcomp
```

#### SupFig 4
```{r extFigDiag, fig.height=5, fig.width=12}

pDiagVarCov + pUMIcomp + plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(size = 14, face = "bold"))

pdf(paste0(outputPath, 'SupFig4_VAF.pdf'), height=5, width=10, onefile = F)
pDiagVarCov + pUMIcomp + plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(size = 14, face = "bold"))
dev.off()

```

## d) dNdS
### Sig genes
```{r dndsTargeted}
# Gene-level and hotspot dN/dS results (Tables S9/S10). SEMG1 and MECP2 are
# excluded as spurious signals (see DDfathers_processing.Rmd).
sig_dndsAll <- supTables_list[["Table S9 dNdSgeneLevel"]]
sig_dndsHotspot <- supTables_list[["Table S10 dNdShotspot"]] |> filter(!is.na(significant))

# Sig twinsUK
sigTwinsUK <- read_tsv(paste0(dataPath, "twinsUK_sigGenes.tsv"), col_types = cols()) |> filter(testSig == "Exome-wide")
sigTwinsUKall <- read_tsv(paste0(dataPath, "twinsUK_SupTable3_sig_dndsAll.tsv"), col_types = cols())

# Gene set of known germline sig genes in TwinsUK/Goriely
knownPosSelection <- c("PTPN11", "FGFR3", "FGFR2", "HRAS", "RET", "BRAF", "KRAS", "MAP2K1", "MAP2K2", "SOS1", "CBL", "RAF1", "SMAD4")
knownGermline <- unique(c(sigTwinsUK  |> pull(gene_name), knownPosSelection))

sigGenes <- sig_dndsAll |> filter(qglobal_cv < 0.01 | gene_name %in% sig_dndsHotspot$gene) |>
  mutate(sigTwinsUK = if_else(gene_name %in% c(sigTwinsUK$gene_name), "Yes", "No"))

paste((sigGenes |> filter(sigTwinsUK == "Yes") |> pull(gene_name)), collapse = ", ")
paste((sigGenes |> filter(sigTwinsUK == "No") |> pull(gene_name)), collapse = ", ")

```

### dnds Plots
#### Objects
```{r objectsdnds}
classOrder <- c("syn", "mis", "non", "spl", "non+spl", "ind")
classOrderLong <- c("synonymous", "missense", "nonsense", "splice", "nonsense+splice", "indel")
geneGroupOrder <- c("excess_other", "excess_dNdS_novel",  "excess_known", "expected")
mutType_colours <- c("synonymous" = "grey70",
                     "missense" = "cadetblue",
                    "nonsense" = "darkorchid4",
                    "nonsense+splice" = "darkorchid3",
                    "splice" = "darkorchid2",
                    "indels" = "chocolate3",
                    "indel" = "chocolate3",
                    "syn" = "grey70",
                     "mis" = "cadetblue",
                    "non" = "darkorchid4",
                    "non+spl" = "darkorchid3",
                    "spl" = "darkorchid2",
                    "ind" = "chocolate3")


```

#### Global dNdS
```{r fig_global_dNdS, fig.height = 4, fig.width = 5}

# Global dN/dS values 
p_globaldnds <- read_tsv(paste0(dataPath, "globaldnds_TriMeth_Panel_withControls.tsv"), col_types = cols()) |> 
  filter(name %in% c("wall")) |> 
  mutate(dataset = factor(dataset, levels = c("DD fathers","Controls"))) |> 
  ggplot(aes(x = dataset, y = mle, fill = dataset, group = dataset)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(
    aes(ymin = cilow, ymax = cihigh),
    position = position_dodge(width = 0.8),
    width = 0.2, color = "black"
  ) +
  geom_hline(yintercept = 1, linetype = 'dashed') + 
  scale_fill_manual(values = dataset_colours) +
  scale_y_continuous(expand = c(0,0)) +
  ylab("Global dN/dS (220 gene panel)") +
  theme_pubr() +
  theme(
    axis.title.x = element_blank(),
    legend.position = "none",
    legend.title = element_blank(),
    strip.background = element_blank()
  )

p_globaldnds
```


#### Gene dNdS
```{r fig_gene_dNdS, fig.height = 8, fig.width = 12}


p_dndsMuts <- sigGenes |> 
  mutate(totMuts = n_syn+n_mis+n_spl+n_non+n_ind)|> 
  #Group non+spl
  mutate(`n_non+spl` = n_spl+n_non) |> select(-c(n_spl,n_non)) |> 
  pivot_longer(cols = starts_with("n_"), names_to = "class", names_prefix = "n_", values_to = "muts") |> 
  mutate(class = fct_relevel(class, rev(classOrder))) |>
  mutate(gene_name = fct_reorder(gene_name, totMuts, .desc = T)) |>
  ggplot(aes(x = gene_name, y = muts, fill = class)) + 
    geom_bar(position="stack", stat="identity") +
  scale_fill_manual(values = mutType_colours) +
  scale_y_continuous(expand = c(0,0)) +
  coord_flip() +
  ylab("Mutations") +
  theme_pubr() +
  theme(axis.title.y = element_blank(), legend.position=c(.82,.6), 
        strip.background = element_blank(), strip.text =element_text(size = 14))

p_dndsRatios <- sigGenes |>  
  mutate(totMuts = n_syn+n_mis+n_spl+n_non+n_ind) |> 
  pivot_longer(cols = starts_with("w"), names_to = "class", names_prefix = "w", values_to = "ratio") |> 
  filter(class != "spl_cv") |> 
  mutate(class = str_remove(class, "_cv")) |> 
  mutate(class = str_replace(class, "non", "non+spl")) |> 
  mutate(class = fct_relevel(class, classOrder)) |>
  mutate(gene_name = fct_reorder(gene_name, totMuts, .desc = T)) |>
  ggplot(aes(x = gene_name, y = ratio, fill = class)) + 
    geom_bar(position="dodge", stat="identity") +
  scale_fill_manual(values = mutType_colours) +
  scale_y_continuous(expand = c(0,0)) +
  coord_flip() +
  ylab("dN/dS Observed/Expected") + 
  xlab("Targeted") +
  theme_pubr() +
  theme(axis.title.y = element_blank(), axis.text.y = element_blank(),
        legend.position="none",
        strip.background = element_blank(), strip.text = element_text(colour = "white"), axis.ticks.y = element_blank()) 
p_dndsMuts + p_dndsRatios + plot_layout(widths = c(1, 2)) 

```

#### Sig Comp
```{r fig_sig_dNdS, fig.height = 6, fig.width = 9}

sigComp <- sig_dndsAll |> 
  select(gene_name, pglobal_cv, qglobal_cv) |> 
  mutate(pglobal_cv = if_else(pglobal_cv <1e-30 , 1e-30, pglobal_cv)) |>
  left_join(
    read_tsv(paste0(dataPath, "twinsUK_SupTable3_sig_dndsAll.tsv"), col_types = cols()) |> 
      select(gene_name, pTwinsUK = pglobal_cv, qTwinsUK = qglobal_cv), 
    by = join_by(gene_name)
  ) |> 
  drop_na() |> 
  mutate(pglobal_cv = if_else(pglobal_cv == 0, 1e-15, pglobal_cv)) |> 
  mutate(across(where(is.numeric), ~ ifelse(. == 0, 1e-15, .))) |> 
  # Label signficance across datasets
  mutate(dndsLabel = if_else(gene_name %in% sigTwinsUK$gene_name & gene_name %in% sigGenes$gene_name, "DD fathers & Controls", "NS")) |> 
  mutate(dndsLabel = if_else(gene_name %in% sigTwinsUK$gene_name & !(gene_name %in% sigGenes$gene_name), "Controls", dndsLabel)) |> 
  mutate(dndsLabel = if_else(!(gene_name %in% sigTwinsUK$gene_name) & gene_name %in% sigGenes$gene_name, "DD fathers", dndsLabel)) |> 
  mutate(gene_name = if_else(dndsLabel == "NS", "", gene_name)) |> 
  mutate(dndsLabel = factor(dndsLabel, levels = c("DD fathers & Controls","DD fathers","Controls","NS")))

# Compute approximate p-value cutoffs
last_below_q_global <- sigComp |> arrange(qglobal_cv) |> filter(qglobal_cv <= 0.01) |> slice(n()) |> pull(pglobal_cv)
first_above_q_global <- sigComp |> arrange(qglobal_cv) |> filter(qglobal_cv > 0.01) |> slice(1) |> pull(pglobal_cv)
p_cutoff_global <- (last_below_q_global + first_above_q_global) / 2

last_below_q_twinsUK <- sigComp |> arrange(qTwinsUK) |> filter(qTwinsUK <= 0.1) |> slice(n()) |> pull(pTwinsUK)
first_above_q_twinsUK <- sigComp |> arrange(qTwinsUK) |> filter(qTwinsUK > 0.1) |> slice(1) |> pull(pTwinsUK)
p_cutoff_twinsUK <- (last_below_q_twinsUK + first_above_q_twinsUK) / 2


# Create a dummy row to fake the asterisk legend entry
legend_key_df <- tibble(
  pglobal_cv = Inf,
  pTwinsUK = Inf,
)

# Plot
p_sigComp <- ggplot(sigComp, aes(x = -log(pglobal_cv), y = -log(pTwinsUK), fill = dndsLabel)) +
  # NS points first (lighter and underneath)
  geom_point(data = sigComp |> filter(dndsLabel == "NS"),
             aes(fill = dndsLabel), size = 3.5, pch = 21,
             color = "white", stroke = 0.25, alpha = 0.5) +
  # All other points on top
  geom_point(data = sigComp |> filter(dndsLabel != "NS"),
             aes(fill = dndsLabel), size = 3.5, pch = 21,
             color = "white", stroke = 0.25) +
  # FDR cutoff: only horizontal line appears in legend
  geom_vline(xintercept = -log(p_cutoff_global), linetype = "dashed", color = "red") +
  geom_hline(aes(yintercept = -log(p_cutoff_twinsUK), linetype = "FDR q cutoff"), color = "red") +
  # Labels
  geom_text_repel(
    aes(label = gene_name), 
    data = sigComp |> filter(!(is.na(dndsLabel))), 
    size = 3, color = "black", max.overlaps = Inf) +
  # Legend controls
  scale_fill_manual(
    name = "dN/dS Significance",
    values = c(dataset_colours),
    breaks = c("DD fathers & Controls", "DD fathers", "Controls", "NS", "* Hotspot significant")
  ) +
  scale_linetype_manual(
    name = NULL,  # remove title
    values = c("FDR q cutoff" = "dashed")
  ) +
  # Axis labels
  labs(x = "Gene -log(p) DD fathers", y = "Gene -log(p) Controls") +
  # Final theme
  theme_pubr() +
  theme(
    legend.position = c(0.85, 0.6),
    legend.title.align = 0)

p_sigComp


```

## e) Path burden
```{r vafSums}
# TwinsUK VAF sums (tukVafSums) are individual-level and controlled-access - see
# README. When twinsUKDataAvailable it's read in directly as in the manuscript
# (relabelling vafGroup "ClinDDlofCADD" -> "likelyDisease" to match the naming
# used below for the DD-father side); otherwise the age-trend fits/stats derived
# from it (data/twinsUK_vafFits.tsv, data/twinsUK_exomePanelCorrection_stats.tsv -
# both pure aggregate summaries) are used instead further down.
if (twinsUKDataAvailable) {
  tukVafSums <- read_tsv(paste0(dataPath, "tukGermanPanelVafSums.tsv"), col_types = cols()) |>
    mutate(source = "Controls") |>
    mutate(vafGroup = if_else(vafGroup == "ClinDDlofCADD", "likelyDisease", vafGroup))
} else {
  twinsUK_vafFits <- read_tsv(paste0(dataPath, "twinsUK_vafFits.tsv"), col_types = cols())
  twinsUK_exomePanelCorrection_stats <- read_tsv(paste0(dataPath, "twinsUK_exomePanelCorrection_stats.tsv"), col_types = cols())
}

vafSumsVars <- spermVariants |>
  bind_rows(spermVariants |> filter(likelyDisease) |> mutate(vafGroup = "likelyDisease")) |> 
  bind_rows(spermVariants |> filter(likelyDriver) |> mutate(vafGroup = "allPos")) |> 
  bind_rows(spermVariants |> filter(likelyDisease & likelyDriver) |> mutate(vafGroup = "posPath")) |>
  group_by(Sample_PD_ID, Age, vafGroup) |> 
  summarize(vafSum = sum(DUPLEX_VAF, na.rm = TRUE), countSum = sum(TIMES_CALLED), .groups = "drop") |> 
  complete(nesting(Sample_PD_ID, Age), vafGroup, fill = list(countSum=0, vafSum = 0)) |> 
  mutate(method = "obs") |>
  mutate(source = if_else(str_detect(Sample_PD_ID, "PD6"), "DD fathers", "Apert fathers"))
```

### Selection VAF
```{r geneContribC, fig.height=3, fig.width=8}

# TwinsUK's quasibinomial age-trend (fit live here when available, otherwise
# drawn from the pre-computed aggregate fit) is added as a line+ribbon layer
# rather than bind_rows-ing individual TwinsUK points into the plot data.
driverVafBase <- vafSumsVars |> filter(vafGroup == "allPos" & source != "Apert fathers")
if (twinsUKDataAvailable) driverVafBase <- bind_rows(driverVafBase, tukVafSums |> filter(vafGroup == "allPos"))

p_vafSumsGenesDriver <- driverVafBase |>
  ggplot(aes(x = Age, y = vafSum, colour = source, fill = source)) +
  geom_point(size = 2) +
  stat_smooth(method = "glm", formula = 'y ~ x', method.args = list(family = "quasibinomial"),
              se = TRUE, level = 0.95) +
  scale_colour_manual(values = dataset_colours) +
  scale_fill_manual(values = dataset_colours) +
    scale_y_continuous(labels = scales::percent, limits = c(0, NA)) +
  ylab("% of sperm with known driver") +
  xlab('Age (years)') +
  theme_pubr() +
  theme(legend.position = c(0.2,0.8), legend.title = element_blank(), strip.background = element_blank())

if (!twinsUKDataAvailable) {
  driverVafFitTUK <- predict_from_fit_table(twinsUK_vafFits, "vafSum_allPos_quasibinom", seq(min(driverVafBase$Age), max(driverVafBase$Age), by = 1))
  p_vafSumsGenesDriver <- p_vafSumsGenesDriver +
    geom_ribbon(data = driverVafFitTUK, aes(x = Age, ymin = ci_lo, ymax = ci_hi, fill = "Controls"), inherit.aes = FALSE, alpha = 0.25) +
    geom_line(data = driverVafFitTUK, aes(x = Age, y = fit, colour = "Controls"), inherit.aes = FALSE, linewidth = 1.1)
}

p_vafSumsGenesDriver

driverVafBaseApert <- vafSumsVars |> filter(vafGroup == "allPos")
if (twinsUKDataAvailable) driverVafBaseApert <- bind_rows(driverVafBaseApert, tukVafSums |> filter(vafGroup == "allPos"))

p_vafSumsGenesDriverApert <- driverVafBaseApert |>
  ggplot(aes(x = Age, y = vafSum, colour = source, fill = source)) +
  geom_point(size = 2) +
  stat_smooth(method = "glm", formula = 'y ~ x', method.args = list(family = "quasibinomial"),
              se = TRUE, level = 0.95) +
  scale_colour_manual(values = dataset_colours) +
  scale_fill_manual(values = dataset_colours) +
    scale_y_continuous(labels = scales::percent, limits = c(0, NA)) +
  ylab("% of sperm with known driver") +
  xlab('Age (years)') +
  theme_pubr() +
  theme(legend.position = c(0.2,0.8), legend.title = element_blank(), strip.background = element_blank())

if (!twinsUKDataAvailable) {
  driverVafFitTUKApert <- predict_from_fit_table(twinsUK_vafFits, "vafSum_allPos_quasibinom", seq(min(driverVafBaseApert$Age), max(driverVafBaseApert$Age), by = 1))
  p_vafSumsGenesDriverApert <- p_vafSumsGenesDriverApert +
    geom_ribbon(data = driverVafFitTUKApert, aes(x = Age, ymin = ci_lo, ymax = ci_hi, fill = "Controls"), inherit.aes = FALSE, alpha = 0.25) +
    geom_line(data = driverVafFitTUKApert, aes(x = Age, y = fit, colour = "Controls"), inherit.aes = FALSE, linewidth = 1.1)
}

p_vafSumsGenesDriverApert

fit <- glm(vafSum ~ Age, family = quasibinomial, data = vafSumsVars |> filter(vafGroup == "allPos" & source == "DD fathers"))
summary(fit)
# Get the predicted values (on the logit scale) and the standard errors
preds <- predict(fit, data.frame(Age = c(30, 60)), type = "link", se.fit = TRUE)
predicted_vals <- preds$fit

# Calculate the 95% confidence intervals on the logit scale
lower_ci <- predicted_vals - 1.96 * preds$se.fit
upper_ci <- predicted_vals + 1.96 * preds$se.fit

predsDriverRate <- tibble(
  Age = c(30, 60),
  Predicted_VAF = plogis(predicted_vals),
  Lower_CI = plogis(lower_ci),
  Upper_CI = plogis(upper_ci))


```

##### Figure 2
```{r fig2, fig.height=10, fig.width=13}

((p_globaldnds + p_vafSumsGenesDriver + plot_layout(widths = c(1, 4))) /
  (p_dndsMuts + p_dndsRatios + p_sigComp + plot_layout(widths = c(1, 1, 3)))) +
  plot_layout(heights = c(1, 2)) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(size = 14, face = "bold"))

pdf(paste0(outputPath, 'Fig2_selection.pdf'), height=10, width=13, onefile = F)
((p_globaldnds + p_vafSumsGenesDriver + plot_layout(widths = c(1, 4))) /
  (p_dndsMuts + p_dndsRatios + p_sigComp + plot_layout(widths = c(1, 1, 3)))) +
  plot_layout(heights = c(1, 2)) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(size = 14, face = "bold"))
dev.off()

```
### Disease VAF
#### Calibrate exome correction
This whole calibration (exome-to-panel capture fraction) compares TwinsUK's own
panel vs exome sequencing of the same individuals - there's no DD-father
equivalent (DD fathers weren't exome sequenced), so when TwinsUK data isn't
available there's nothing to plot as points; `exomePanelCorrection` (used
throughout the rest of this notebook to scale DD-father panel VAF sums to
exome-equivalent estimates) and the reported stats are read back from the
pre-computed aggregate summary instead.
```{r germanTwin}
if (twinsUKDataAvailable) {
  # Exome→panel capture fraction + stats to report (R², slope, P for age, correlation, n)

  ## 1) Point estimate: fraction of exome-wide likely-disease VAF captured by 220-gene panel
  fractionWithinPanel <- tukVafSums |>
    group_by(vafGroup) |>
    summarize(
      totalVAFs      = sum(vafSum, na.rm = TRUE),
      totalVAFsExome = sum(exome_vafSum, na.rm = TRUE),
      .groups = "drop"
    ) |>
    mutate(percentageInPanel = totalVAFs / totalVAFsExome)

  exomePanelCorrection <- fractionWithinPanel |>
    filter(vafGroup == "likelyDisease") |>
    pull(percentageInPanel)

  ## 2) Per-sample ratio + basic QC + age-association stats
  panel_exome_df <- tukVafSums |>
    filter(vafGroup == "likelyDisease") |>
    mutate(
      percentageInPanel = vafSum / exome_vafSum
    ) |>
    filter(is.finite(percentageInPanel), percentageInPanel > 0)

  n_samples <- nrow(panel_exome_df)
  n_sources <- dplyr::n_distinct(panel_exome_df$source)

  # Correlation of panel vs exome burden (VAF sums) across samples
  cor_test <- cor.test(panel_exome_df$vafSum, panel_exome_df$exome_vafSum, method = "pearson")

  # Age dependence of the fraction captured (key claim: no age trend)
  fit_age <- lm(percentageInPanel ~ Age, data = panel_exome_df)
  fit_age_sum <- summary(fit_age)
  beta_age <- coef(fit_age)["Age"]
  se_age   <- coef(fit_age_sum)["Age", "Std. Error"]
  p_age    <- coef(fit_age_sum)["Age", "Pr(>|t|)"]
  r2_age   <- fit_age_sum$r.squared

  # Summary distribution of the per-sample fraction (helps justify using a fixed fraction)
  frac_summ <- panel_exome_df |>
    summarize(
      mean_frac   = mean(percentageInPanel, na.rm = TRUE),
      median_frac = median(percentageInPanel, na.rm = TRUE),
      sd_frac     = sd(percentageInPanel, na.rm = TRUE),
      iqr_frac    = IQR(percentageInPanel, na.rm = TRUE),
      min_frac    = min(percentageInPanel, na.rm = TRUE),
      max_frac    = max(percentageInPanel, na.rm = TRUE)
    )

  list(
    n_samples = n_samples,
    n_sources = n_sources,
    exomePanelCorrection = exomePanelCorrection,
    frac_summary = frac_summ,
    panel_vs_exome_cor = list(
      estimate = unname(cor_test$estimate),
      conf_int = cor_test$conf.int,
      p_value  = cor_test$p.value
    ),
    age_effect_on_fraction = list(
      beta_per_year = beta_age,
      se = se_age,
      p_value = p_age,
      r_squared = r2_age
    )
  )

  ## 3) Plot
  pExomeToPanelFraction <- panel_exome_df |>
    ggplot(aes(x = Age, y = percentageInPanel, fill = source, colour = source)) +
    geom_point() +
    geom_smooth(method = "lm") +
    scale_colour_manual(values = dataset_colours) +
    scale_fill_manual(values = dataset_colours) +
    scale_y_continuous(labels = scales::percent, limits = c(0, NA)) +
    ylab("% of exome-wide likely disease\nvariants in 220-gene panel") +
    theme_pubr() +
    theme(
      legend.position = c(0.8, 0.8),
      legend.title = element_blank()
    )
} else {
  exomePanelCorrection <- twinsUK_exomePanelCorrection_stats$exomePanelCorrection
  print(twinsUK_exomePanelCorrection_stats)

  # No DD-father points exist for this comparison - show the TwinsUK fitted
  # trend only (Age grid + CI, no individual points).
  exomeFractionFitTUK <- predict_from_fit_table(twinsUK_vafFits, "percentageInPanel_exomeCorrection", age_grid_fine$Age)
  pExomeToPanelFraction <- ggplot(exomeFractionFitTUK, aes(x = Age, y = fit)) +
    geom_ribbon(aes(ymin = ci_lo, ymax = ci_hi), fill = dataset_colours[["Controls"]], alpha = 0.25) +
    geom_line(colour = dataset_colours[["Controls"]], linewidth = 1.1) +
    scale_y_continuous(labels = scales::percent, limits = c(0, NA)) +
    ylab("% of exome-wide likely disease\nvariants in 220-gene panel\n(TwinsUK fit only - see README)") +
    theme_pubr()
}

pExomeToPanelFraction

```

#### German vs TwinsUK
TwinsUK VAF sums are individual-level and controlled-access (see README). The
plots below use vafSumsVars (DD/Apert fathers) points directly; the TwinsUK
trend line comes from either the live tukVafSums fit (twinsUKDataAvailable) or
the pre-computed aggregate fit (data/twinsUK_vafFits.tsv) - never TwinsUK's
individual points. The superseded naive-scaling VafSumAdjusted43 draft (using a
flat 1/exomePanelCorrection multiplier instead of the age-adjusted TwinsUK
model) has been dropped - only the final, age-adjusted version is built below.
```{r germanTwinVaf}

likelyDiseaseBase <- vafSumsVars |> filter(source != "Apert fathers", vafGroup == "likelyDisease")
if (twinsUKDataAvailable) likelyDiseaseBase <- bind_rows(likelyDiseaseBase, tukVafSums |> filter(vafGroup == "likelyDisease"))

pVafSumLikelyDisease <- likelyDiseaseBase |>
  ggplot(aes(x = Age, y = vafSum, colour = source, fill = source)) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::percent, limits = c(0, NA)) +
  ylab("% of sperm with likely disease mutation\n(220 gene panel; no mosaics)") +
  xlab('Age (years)') +
  scale_colour_manual(values = dataset_colours) +
  scale_fill_manual(values = dataset_colours) +
  stat_smooth(method = "lm", formula = 'y ~ x') +
  theme_pubr() +
  theme(strip.background = element_blank(),
        legend.title = element_blank(),
        legend.position = c(0.2, 0.85),
        plot.title = element_text(hjust = 0.5, size = 10))

if (!twinsUKDataAvailable) {
  likelyDiseaseFitTUK <- predict_from_fit_table(twinsUK_vafFits, "vafSum_likelyDisease", seq(min(likelyDiseaseBase$Age), max(likelyDiseaseBase$Age), by = 1))
  pVafSumLikelyDisease <- pVafSumLikelyDisease +
    geom_ribbon(data = likelyDiseaseFitTUK, aes(x = Age, ymin = ci_lo, ymax = ci_hi, fill = "Controls"), inherit.aes = FALSE, alpha = 0.25) +
    geom_line(data = likelyDiseaseFitTUK, aes(x = Age, y = fit, colour = "Controls"), inherit.aes = FALSE, linewidth = 1.1)
}
pVafSumLikelyDisease

likelyDiseaseBaseApert <- vafSumsVars |> filter(vafGroup == "likelyDisease")
if (twinsUKDataAvailable) likelyDiseaseBaseApert <- bind_rows(likelyDiseaseBaseApert, tukVafSums |> filter(vafGroup == "likelyDisease"))

pVafSumLikelyDiseaseApert <- likelyDiseaseBaseApert |>
  ggplot(aes(x = Age, y = vafSum, colour = source, fill = source)) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::percent, limits = c(0, NA)) +
  ylab("% of sperm with likely disease mutation\n(220 gene panel; no mosaics)") +
  xlab('Age (years)') +
  scale_colour_manual(values = dataset_colours) +
  scale_fill_manual(values = dataset_colours) +
  stat_smooth(method = "lm", formula = 'y ~ x') +
  theme_pubr() +
  theme(strip.background = element_blank(),
        legend.title = element_blank(),
        legend.position = c(0.2, 0.85),
        plot.title = element_text(hjust = 0.5, size = 10))

if (!twinsUKDataAvailable) {
  likelyDiseaseFitTUKApert <- predict_from_fit_table(twinsUK_vafFits, "vafSum_likelyDisease", seq(min(likelyDiseaseBaseApert$Age), max(likelyDiseaseBaseApert$Age), by = 1))
  pVafSumLikelyDiseaseApert <- pVafSumLikelyDiseaseApert +
    geom_ribbon(data = likelyDiseaseFitTUKApert, aes(x = Age, ymin = ci_lo, ymax = ci_hi, fill = "Controls"), inherit.aes = FALSE, alpha = 0.25) +
    geom_line(data = likelyDiseaseFitTUKApert, aes(x = Age, y = fit, colour = "Controls"), inherit.aes = FALSE, linewidth = 1.1)
}
pVafSumLikelyDiseaseApert

# Fit linear model to twinsUK data (or use the pre-computed fit)
if (twinsUKDataAvailable) {
  twins_model <- lm(exome_vafSum ~ Age, data = tukVafSums |> filter(vafGroup == "likelyDisease"))
}

# Add age-based prediction for German data
german_adjusted <- vafSumsVars |>
  filter(vafGroup == "likelyDisease") |>
  mutate(twinsUK_pred = if (twinsUKDataAvailable) predict(twins_model, newdata = cur_data())
                        else predict_from_fit_table(twinsUK_vafFits, "exome_vafSum_likelyDisease", Age)$fit,
         adjustment_factor = twinsUK_pred * 0.565,
         exome_vafSum = vafSum + adjustment_factor,
         exome_countSum = round(countSum * (1/exomePanelCorrection)))

# VafSumAdjusted43 is DD fathers + Apert fathers only unless twinsUKDataAvailable
VafSumAdjusted43 <- german_adjusted
if (twinsUKDataAvailable) VafSumAdjusted43 <- bind_rows(VafSumAdjusted43, tukVafSums |> filter(vafGroup == "likelyDisease"))

pVafSumAdjusted43 <- VafSumAdjusted43 |>
  ggplot(aes(x = Age, y = exome_vafSum, colour = source, fill = source)) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::percent, limits = c(0, NA)) +
  ylab("% of sperm with likely disease mutation") +
  xlab('Age (years)') +
  scale_colour_manual(values = dataset_colours) +
  scale_fill_manual(values = dataset_colours) +
  stat_smooth(method = "lm", formula = 'y ~ x') +
  theme_pubr() +
  theme(strip.background = element_blank(),
        legend.title = element_blank(),
        legend.position = c(0.2, 0.85),
        plot.title = element_text(hjust = 0.5, size = 10))

if (!twinsUKDataAvailable) {
  exomeVafFitTUK <- predict_from_fit_table(twinsUK_vafFits, "exome_vafSum_likelyDisease", seq(min(VafSumAdjusted43$Age), max(VafSumAdjusted43$Age), by = 1))
  pVafSumAdjusted43 <- pVafSumAdjusted43 +
    geom_ribbon(data = exomeVafFitTUK, aes(x = Age, ymin = ci_lo, ymax = ci_hi, fill = "Controls"), inherit.aes = FALSE, alpha = 0.25) +
    geom_line(data = exomeVafFitTUK, aes(x = Age, y = fit, colour = "Controls"), inherit.aes = FALSE, linewidth = 1.1)
}
pVafSumAdjusted43

```

##### Figure 3 
```{r germanTwinFig3, fig.height=8, fig.width=11}

# --- 1) Prepare mosaic variant info ---
mosaicVars <- diagnosticVarsCalls |> 
  filter(varLabel != "") |> 
  left_join(spermSamplesFiltered |> select(Sample_PD_ID, FAM_ID, Age)) |> 
  select(Sample_PD_ID, Age, mosaic_vaf = duplexVAF) |> 
  distinct() |> 
  filter(mosaic_vaf > 0.005)

# --- 2) Age-based adjustment & combined frame ---
if (twinsUKDataAvailable) {
  twins_model <- lm(exome_vafSum ~ Age, data = tukVafSums |> filter(vafGroup == "likelyDisease"))
}

german_adjusted <- vafSumsVars |>
  filter(vafGroup == "likelyDisease") |>
  mutate(
    twinsUK_pred = if (twinsUKDataAvailable) predict(twins_model, newdata = cur_data())
                   else predict_from_fit_table(twinsUK_vafFits, "exome_vafSum_likelyDisease", Age)$fit,
    adjustment_factor = twinsUK_pred * (1 - exomePanelCorrection),
    exome_vafSum = vafSum + adjustment_factor
  )


# --- 3) Build mosaic points (base + mosaic totals) ---
originalPoints <- VafSumAdjusted43 |> 
  semi_join(mosaicVars, by = "Sample_PD_ID") |> 
  select(Sample_PD_ID, exome_vafSum, source)

mosaicPoints <- originalPoints |> 
  left_join(mosaicVars |> select(Sample_PD_ID, mosaic_vaf, Age), by = "Sample_PD_ID") |> 
  mutate(
    mosaic_total_vaf = exome_vafSum + mosaic_vaf,
    label = paste0(scales::percent(mosaic_vaf, accuracy = 0.1), " mosaic\n", scales::percent(exome_vafSum, accuracy = 0.1), " other"),
    source_display = case_when(
      source == "twinsUK" ~ "Controls",
      source == "germanTrio" ~ "DD fathers",
      TRUE ~ source
    )
  ) 

# --- 4) Split datasets for plotting & legend entries ---
mosaic_ids <- mosaicPoints |> distinct(Sample_PD_ID) |> pull()

# Non-mosaic data only (exclude the 4 mosaic fathers); recode for legend
base_non_mosaic0 <- VafSumAdjusted43 |>
  filter(!(Sample_PD_ID %in% mosaic_ids)) |>
  mutate(source_display = source)

# Mosaic fathers: base exome (without mosaic)
mosaic_without <- VafSumAdjusted43 |>
  semi_join(mosaicVars, by = "Sample_PD_ID") |>
  select(Sample_PD_ID, Age, exome_vafSum) |>
  mutate(source_display = "Mosaic DD fathers - other")

# Mosaic fathers: exome + mosaic (total)
mosaic_with <- mosaicPoints |>
  transmute(Sample_PD_ID, Age,
            exome_vafSum = mosaic_total_vaf,
            label,
            source_display = "Mosaic DD fathers - total")

# --- 5) Scales (extend the palette with two new legend items) ---
legend_order <- c("Controls",
                  "DD fathers",
                  "Apert fathers",
                  "Mosaic DD fathers - other",
                  "Mosaic DD fathers - total")

col_values  <- c(dataset_colours,
                 "Mosaic DD fathers - other" = "black",
                 "Mosaic DD fathers - total" = "black")

fill_values <- c(dataset_colours,
                 "Mosaic DD fathers - other" = "#ef8a62",
                 "Mosaic DD fathers - total" = "goldenrod2")

# Ensure legend order by setting factors (optional but robust)
base_non_mosaic <- base_non_mosaic0  |> mutate(source_display = factor(source_display, levels = legend_order)) |> filter(source != "Apert fathers")
base_non_mosaicApert <- base_non_mosaic0  |> mutate(source_display = factor(source_display, levels = legend_order))

mosaic_without  <- mosaic_without  |> mutate(source_display = factor(source_display, levels = legend_order))
mosaic_with     <- mosaic_with     |> mutate(source_display = factor(source_display, levels = legend_order)) 

normal_lbl <- mosaic_with |> dplyr::filter(!(Sample_PD_ID %in% c("PD63917c_tds0001", "PD63954c_tds0001", "PD63911c_tds0001", "PD68787b_tds0001", "PD68809b_tds0001")))
nudge_lbl  <- mosaic_with |> dplyr::filter(Sample_PD_ID %in% "PD63917c_tds0001")
right_lbl  <- mosaic_with |> dplyr::filter(Sample_PD_ID %in% c("PD63954c_tds0001", "PD63911c_tds0001"))
bottom_lbl  <- mosaic_with |> dplyr::filter(Sample_PD_ID %in% c("PD68787b_tds0001", "PD68809b_tds0001"))
# --- 6) Plot ---
pVafSumAdjusted43 <- ggplot() +
  # Non-mosaic controls & DD fathers
  geom_point(data = base_non_mosaic,
             aes(x = Age, y = exome_vafSum,
                 colour = source_display, fill = source_display),
             size = 2) +
  stat_smooth(data = base_non_mosaic,
              aes(x = Age, y = exome_vafSum, colour = source_display, fill = source_display),
              method = "lm", formula = y ~ x) +
  # Dotted up-ticks for mosaic fathers
  geom_segment(data = mosaicPoints,
               aes(x = Age, xend = Age,
                   y = exome_vafSum, yend = mosaic_total_vaf),
               linetype = "dotted", colour = "black",
               inherit.aes = FALSE, show.legend = FALSE) +
  # Mosaic fathers: base exome (without mosaic)
  geom_point(data = mosaic_without,
             aes(x = Age, y = exome_vafSum,
                 colour = source_display, fill = source_display),
             shape = 21, size = 2, stroke = 0.75) +
  # Mosaic fathers: exome + mosaic (total)
  geom_point(data = mosaic_with,
             aes(x = Age, y = exome_vafSum,
                 colour = source_display, fill = source_display),
             shape = 21, size = 2, stroke = 0.75) +
  # Labels for mosaic increment
  geom_label_repel(
    data = normal_lbl,
    aes(x = Age, y = exome_vafSum, label = label),
    fill = "white", colour = "black", size = 3,
    label.size = 0.25, inherit.aes = FALSE,
    hjust = 0.5, direction = "x", nudge_x = -4,
    min.segment.length = 0, show.legend = FALSE) +
  geom_label_repel(
    data = right_lbl,
    aes(x = Age, y = exome_vafSum, label = label),
    fill = "white", colour = "black", size = 3,
    label.size = 0.25, inherit.aes = FALSE,
    hjust = 0.5, direction = "x", nudge_x = 4,
    min.segment.length = 0, show.legend = FALSE) +
  geom_label_repel(
    data = nudge_lbl,
    aes(x = Age, y = exome_vafSum, label = label),
    fill = "white", colour = "black", size = 3,
    label.size = 0.25, inherit.aes = FALSE,
    hjust = 0.5, direction = "x", nudge_x = -12,
    min.segment.length = 0, show.legend = FALSE) +
  geom_label_repel(
    data = bottom_lbl,
    aes(x = Age, y = exome_vafSum, label = label),
    fill = "white", colour = "black", size = 3,
    label.size = 0.25, inherit.aes = FALSE,
    hjust = 0.5, direction = "y", 
    nudge_y = -1,   # or -1, -5, etc. depending on exome_vafSum scale
    min.segment.length = 0, show.legend = FALSE
  ) +
  # Axes & theme
  scale_y_continuous(labels = scales::percent, limits = c(0, NA)) +
  ylab("% of sperm with likely disease mutation") +
  xlab("Age (years)") +
  scale_colour_manual(values = col_values, breaks = legend_order, name = NULL) +
  scale_fill_manual(values = fill_values, breaks = legend_order, name = NULL) +
  theme_pubr() +
  theme(
    strip.background = element_blank(),
    legend.title = element_blank(),
    legend.position = c(0.7, 0.85),
    plot.title = element_text(hjust = 0.5, size = 10)
  )

if (!twinsUKDataAvailable) {
  exomeVafFitTUKFig3 <- predict_from_fit_table(twinsUK_vafFits, "exome_vafSum_likelyDisease", seq(min(base_non_mosaic$Age), max(base_non_mosaic$Age), by = 1))
  pVafSumAdjusted43 <- pVafSumAdjusted43 +
    geom_ribbon(data = exomeVafFitTUKFig3, aes(x = Age, ymin = ci_lo, ymax = ci_hi, fill = "Controls"), inherit.aes = FALSE, alpha = 0.25) +
    geom_line(data = exomeVafFitTUKFig3, aes(x = Age, y = fit, colour = "Controls"), inherit.aes = FALSE, linewidth = 1.1)
}

pVafSumAdjusted43

# --- 6) Plot ---
pVafSumAdjusted43Apert <- ggplot() +
  # Non-mosaic controls & DD fathers
  geom_point(data = base_non_mosaicApert,
             aes(x = Age, y = exome_vafSum,
                 colour = source_display, fill = source_display),
             size = 2) +
  stat_smooth(data = base_non_mosaicApert,
              aes(x = Age, y = exome_vafSum, colour = source_display, fill = source_display),
              method = "lm", formula = y ~ x) +
  # Dotted up-ticks for mosaic fathers
  geom_segment(data = mosaicPoints,
               aes(x = Age, xend = Age,
                   y = exome_vafSum, yend = mosaic_total_vaf),
               linetype = "dotted", colour = "black",
               inherit.aes = FALSE, show.legend = FALSE) +
  # Mosaic fathers: base exome (without mosaic)
  geom_point(data = mosaic_without,
             aes(x = Age, y = exome_vafSum,
                 colour = source_display, fill = source_display),
             shape = 21, size = 2, stroke = 0.75) +
  # Mosaic fathers: exome + mosaic (total)
  geom_point(data = mosaic_with,
             aes(x = Age, y = exome_vafSum,
                 colour = source_display, fill = source_display),
             shape = 21, size = 2, stroke = 0.75) +
  # Labels for mosaic increment
  geom_label_repel(
    data = normal_lbl,
    aes(x = Age, y = exome_vafSum, label = label),
    fill = "white", colour = "black", size = 3,
    label.size = 0.25, inherit.aes = FALSE,
    # hjust = 0.5, direction = "x", nudge_x = -4,
    min.segment.length = 0, show.legend = FALSE) +
  geom_label_repel(
    data = right_lbl,
    aes(x = Age, y = exome_vafSum, label = label),
    fill = "white", colour = "black", size = 3,
    label.size = 0.25, inherit.aes = FALSE,
    hjust = 0.5, direction = "x", nudge_x = 4,
    min.segment.length = 0, show.legend = FALSE) +
  geom_label_repel(
    data = nudge_lbl,
    aes(x = Age, y = exome_vafSum, label = label),
    fill = "white", colour = "black", size = 3,
    label.size = 0.25, inherit.aes = FALSE,
    hjust = 0.5, direction = "x", nudge_x = -12,
    min.segment.length = 0, show.legend = FALSE) +
  # Axes & theme
  scale_y_continuous(labels = scales::percent, limits = c(0, NA)) +
  ylab("% of sperm with likely disease mutation") +
  xlab("Age (years)") +
  scale_colour_manual(values = col_values, breaks = legend_order, name = NULL) +
  scale_fill_manual(values = fill_values, breaks = legend_order, name = NULL) +
  theme_pubr() +
  theme(
    strip.background = element_blank(),
    legend.title = element_blank(),
    legend.position = c(0.7, 0.85),
    plot.title = element_text(hjust = 0.5, size = 10)
  )

if (!twinsUKDataAvailable) {
  exomeVafFitTUKFig3Apert <- predict_from_fit_table(twinsUK_vafFits, "exome_vafSum_likelyDisease", seq(min(base_non_mosaicApert$Age), max(base_non_mosaicApert$Age), by = 1))
  pVafSumAdjusted43Apert <- pVafSumAdjusted43Apert +
    geom_ribbon(data = exomeVafFitTUKFig3Apert, aes(x = Age, ymin = ci_lo, ymax = ci_hi, fill = "Controls"), inherit.aes = FALSE, alpha = 0.25) +
    geom_line(data = exomeVafFitTUKFig3Apert, aes(x = Age, y = fit, colour = "Controls"), inherit.aes = FALSE, linewidth = 1.1)
}

pVafSumAdjusted43Apert

pdf(paste0(outputPath, 'Fig3_Mosaic.pdf'), height = 5, width = 8, onefile = F)
pVafSumAdjusted43
dev.off()

```

##### SupFig 6
```{r extfig2, fig.height=12, fig.width=16}

layoutFigApert <- "
AC
BC"

(snv_plotApert +  labs(tag = "A") + indel_plotApert +  labs(tag = "B") + ((pDomains +  labs(tag = "C"))/trinucDNM/trinucSperm/trinucApert + plot_layout(heights = c(1, 15, 15, 15))) +  plot_layout(design = layoutFigApert , widths = c(1,1.5)))/ (p_vafSumsGenesDriverApert +  labs(tag = "D") + pVafSumLikelyDiseaseApert +  labs(tag = "E")) + plot_layout(heights = c(2,1), nrow = 2) & theme(plot.tag = element_text(size = 14, face = "bold"))

pdf(paste0(outputPath, 'SupFig6_Apert.pdf'), height = 12, width = 16, onefile = F)
(snv_plotApert +  labs(tag = "A") + indel_plotApert +  labs(tag = "B") + ((pDomains +  labs(tag = "C"))/trinucDNM/trinucSperm/trinucApert + plot_layout(heights = c(1, 15, 15, 15))) +  plot_layout(design = layoutFigApert , widths = c(1,1.5)))/ (p_vafSumsGenesDriverApert +  labs(tag = "D") + pVafSumLikelyDiseaseApert+  labs(tag = "E")) + plot_layout(heights = c(2,1), nrow = 2) & theme(plot.tag = element_text(size = 14, face = "bold"))
dev.off()
```

## f) Early mosaic rates
### Published data sets
```{r mosaicSources, fig.width=10, fig.height=6}

mosaicVarsAnnot <- read_tsv(paste0(dataPath, "mosaicVarsAnnot.tsv"), col_types = cols())
mosaicVarsAnnot |> group_by(set, ID) |> summarise(count = n()) |> summarize(mean = mean(count)) 

mosaicVarsAnnot |> group_by(ID,set) |> summarise(count = n()) |> 
  ggplot(aes(x = set, y = count)) +
  geom_point() +
  theme_pubr()

mosaicVarsAnnot |> 
  ggplot(aes(x = set, y = VAF, colour = set)) +
  geom_boxplot() +
  geom_jitter() +
  theme_pubr()

mosaicVarsAnnot |>
  ggplot(aes(VAF, colour = set)) +
  geom_density() +
  theme_pubr()

```

#### Model expected 
```{r mosaicModelExpected, fig.width=10, fig.height=6}
# 1) Analytic expectation + Normal-approx CI
perID_expect <- mosaicVarsAnnot |>
  group_by(ID) |>
  summarise(
    n_variants = n(),
    expected = sum(VAF),
    variance = sum(VAF * (1 - VAF)),
    .groups = "drop"
  ) |>
  mutate(
    ci_low_norm  = pmax(0, qnorm(0.025, mean = expected, sd = sqrt(variance))),
    ci_high_norm = pmin(n_variants, qnorm(0.975, mean = expected, sd = sqrt(variance)))
  )

# 2) Bootstrap Poisson-binomial CI (recommended, per individual)
boot_ci_fun <- function(p, nsims = 20000L) {
  if (length(p) == 0L) return(c(NA_real_, NA_real_))
  mat <- matrix(runif(nsims * length(p)) < rep(p, each = nsims), nrow = nsims)
  totals <- rowSums(mat)
  as.numeric(quantile(totals, c(0.025, 0.975)))
}

set.seed(1)
perID_boot <- mosaicVarsAnnot |>
  group_by(ID) |>
  summarise(
    ci = list(boot_ci_fun(VAF, nsims = 20000L)),
    .groups = "drop"
  ) |>
  mutate(
    ci_low  = purrr::map_dbl(ci, 1),
    ci_high = purrr::map_dbl(ci, 2)
  ) |>
  select(-ci)

# Combine per-ID results
perID_results <- perID_expect |>
  left_join(perID_boot, by = "ID") |>
  select(ID, n_variants, expected,
         ci_low, ci_high,         # bootstrap CI
         ci_low_norm, ci_high_norm)  # normal-approx CI (for reference)

# Bootstrap helper for mean
boot_mean_ci <- function(x, nsims = 20000L) {
  means <- replicate(nsims, mean(sample(x, replace = TRUE)))
  as.numeric(quantile(means, c(0.025, 0.975)))
}

set.seed(1)
mean_expected_ci <- boot_mean_ci(perID_results$expected)
mean_variants_ci <- boot_mean_ci(perID_results$n_variants)

overall_individual <- tibble(
  ID = "Whole genome",
  n_variants = mean(perID_results$n_variants),
  ci_low_n_variants     = mean_variants_ci[1],
  ci_high_n_variants    = mean_variants_ci[2],
  expected_transmitted   = mean(perID_results$expected)*2,
  ci_low_transmitted  = mean_expected_ci[1]*2,
  ci_high_transmitted = mean_expected_ci[2]*2
)

# Final table with per-ID + average
perID_with_avg <- bind_rows(perID_results, overall_individual)

# Plot per-ID expected values
ggplot(perID_results, aes(x = reorder(ID, expected), y = expected)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0.2) +
  coord_flip() +
  labs(
    x = "Individual (ID)",
    y = "Expected transmitted variants (95% CI)"
  )

perID_with_avg


paternal_fraction=0.794
maternal_fraction=0.206
paternal_ss_rate=0.0104
maternal_ss_rate=0.0454
dnms_per_child=68.7
expected_fraction=paternal_fraction*paternal_ss_rate+maternal_fraction*maternal_ss_rate
dnms_per_child*expected_fraction
dnms_per_child*paternal_fraction*paternal_ss_rate
dnms_per_child*maternal_fraction*maternal_ss_rate
expected_fraction

#Risk in goriely paper
1/((0.23 + 8 + 6 + 12.1)/100/59)
```

#### Mosaic carrier estimate:
##### In cohort
```{r mosaic}
# Number of fathers
n_total <- 127

mosaicVAF <- diagnosticVarsCalls |> 
  filter(duplexVAF > 0.001 & duplexCalls > 1) |> 
  mutate(mosaicCount = 1)  |> 
  mutate(source = "DD fathers") |> 
  select(FAM_ID, source, mosaicVAF = duplexVAF, mosaicCount) |> 
  left_join(spermSamplesFiltered |> select(FAM_ID, Sample_PD_ID), by = join_by(FAM_ID))

# Create full VAF vector (6 mosaic VAFs + zeros for the rest)
all_vafs <- c(mosaicVAF$mosaicVAF,
              rep(0, n_total - length(mosaicVAF$mosaicVAF)))

# Point estimate
mean_vaf <- mean(all_vafs)
one_in_estimate <- 1 / mean_vaf

# Bootstrap CI
set.seed(1)
B <- 10000

boot_one_in <- replicate(B, {
  boot_mean <- mean(sample(all_vafs, replace = TRUE))
  1 / boot_mean
})

one_in_ci <- quantile(boot_one_in, c(0.025, 0.975))

list(
  one_in_estimate = one_in_estimate,
  one_in_CI = one_in_ci
)

```

##### After SCN1A exclusion
```{r scn1a_mosaic}
# Re-run the in-cohort mosaic estimate excluding SCN1A-ascertained sperm samples.
# mosaicVAF is keyed by Sample_PD_ID, so filtering is direct.
dd_all_ids   <- VafSumAdjusted43 |> filter(source == "DD fathers") |> pull(Sample_PD_ID)
dd_nonscn_ids <- setdiff(dd_all_ids, scn1a_sperm_ids)
run_mosaic_estimate <- function(mosaic_df, n_total, label) {
  all_vafs <- c(mosaic_df$mosaicVAF, rep(0, n_total - nrow(mosaic_df)))
  mean_vaf <- mean(all_vafs)
  set.seed(1)
  boot_one_in <- replicate(10000, {
    1 / mean(sample(all_vafs, replace = TRUE))
  })
  tibble(
    group         = label,
    n_total       = n_total,
    n_mosaic      = nrow(mosaic_df),
    one_in        = round(1 / mean_vaf),
    ci_low        = round(quantile(boot_one_in, 0.025)),
    ci_high       = round(quantile(boot_one_in, 0.975))
  )
}

mosaic_full   <- mosaicVAF
mosaic_nonscn <- mosaicVAF |> filter(!Sample_PD_ID %in% scn1a_sperm_ids)
n_nonscn      <- length(dd_nonscn_ids)

bind_rows(
  run_mosaic_estimate(mosaic_full,   n_total = 127,      label = "All DD fathers (manuscript)"),
  run_mosaic_estimate(mosaic_nonscn, n_total = n_nonscn, label = "Non-SCN1A DD fathers")
) |>
  mutate(result = paste0("1 in ", one_in, " (95% CI: 1 in ", ci_high, " – 1 in ", ci_low, ")")) |>
  select(group, n_total, n_mosaic, result) |>
  print()
```

## g) Relative contributions
### Cohort fractions
```{r germanAge, fig.width=6, fig.height=5}
# ------------------------------
# Colors
# ------------------------------
plot_colors <- c(
  "Rare (other)"   = "#1F77B4", # Blue (Middle)
  "Rare (known driver)"    = "#6A3D9A", # Green (Top)
  "Mosaic" = "#9E2A2B"  # Red (Overlay)
)
plot_colorsLegend <- c(
  "Pathogenic (early mosaic)"          = "#9E2A2B",
  "Pathogenic (known driver)"       = "#6A3D9A",
  "Pathogenic (other rare variant)"    = "#1F77B4"
)

# ------------------------------
# Input datasets
# ------------------------------
driverVAF <- vafSumsVars |>
  filter(vafGroup == "posPath")
if (twinsUKDataAvailable) driverVAF <- bind_rows(driverVAF, tukVafSums |> filter(vafGroup == "posPath"))
driverVAF <- driverVAF |>
  select(Sample_PD_ID, source, driverVAF = vafSum, driverCount = countSum)

mosaicVAF <- diagnosticVarsCalls |>
  filter(duplexVAF > 0.001 & duplexCalls > 1) |>
  mutate(mosaicCount = 1)  |>
  mutate(source = "DD fathers") |>
  select(FAM_ID, source, mosaicVAF = duplexVAF, mosaicCount) |>
  left_join(spermSamplesFiltered |> select(FAM_ID, Sample_PD_ID), by = join_by(FAM_ID))

# ------------------------------
# Cohort fractions with sample counts
# ------------------------------
cohortFraction <- VafSumAdjusted43 |>
  select(Sample_PD_ID, source, exome_vafSum, exomeCount = exome_countSum) |>
  left_join(driverVAF, by = join_by(Sample_PD_ID, source)) |>
  left_join(mosaicVAF, by = join_by(Sample_PD_ID, source)) |>
  mutate(otherVAF   = exome_vafSum - driverVAF,
         otherCount = exomeCount - driverCount) |>
  group_by(source) |>
  summarise(driverVAF   = sum(driverVAF,   na.rm = TRUE),
            driverCount = sum(driverCount, na.rm = TRUE),
            otherVAF    = sum(otherVAF,    na.rm = TRUE),
            otherCount  = sum(otherCount,  na.rm = TRUE),
            mosaicVAF   = sum(mosaicVAF,   na.rm = TRUE),
            mosaicCount = sum(mosaicCount, na.rm = TRUE),
            n_samples   = n_distinct(Sample_PD_ID),
            .groups = "drop") |>
  mutate(totalVAF = driverVAF + otherVAF + mosaicVAF) |>
  mutate(across(c(driverVAF, otherVAF, mosaicVAF), ~ .x / totalVAF)) |>
  pivot_longer(cols = c(driverVAF, otherVAF, mosaicVAF),
               names_to = "vafType",
               values_to = "vafProp") |>
  mutate(count = case_when(
    vafType == "driverVAF" ~ driverCount,
    vafType == "otherVAF"  ~ otherCount,
    vafType == "mosaicVAF" ~ mosaicCount
  )) |>
  mutate(vafType = recode(vafType,
                          driverVAF = "Rare (known driver)",
                          otherVAF  = "Rare (other)",
                          mosaicVAF = "Mosaic")) |>
  mutate(source_label = paste0(source, "\n", "(n=", n_samples, ")"))

# TwinsUK "Controls" is under controlled access - see README. The saved values
# (data/twinsUK_cohortFraction.tsv) are already a pure aggregate - proportions
# and counts summed across the whole TwinsUK cohort, no individual data - added
# back in only when the live data isn't available.
if (!twinsUKDataAvailable) {
  cohortFraction <- bind_rows(
    cohortFraction,
    read_tsv(paste0(dataPath, "twinsUK_cohortFraction.tsv"), col_types = cols()) |>
      mutate(source_label = paste0(source, "\n", "(n=", n_samples, ")"))
  )
}

cohortFraction <- cohortFraction |>
  mutate(source_label = fct_relevel(source_label,
                                    "DD fathers\n(n=127)",
                                    "Apert fathers\n(n=8)",
                                    "Controls\n(n=36)")) |>
  filter(vafProp != 0)

# ------------------------------
# Plot
# ------------------------------
p_cohortFraction <- ggplot(cohortFraction,
       aes(x = source_label, y = vafProp, fill = vafType)) +
  geom_col(color = "black") +
  geom_label(
    aes(label = paste0("vars = ", count), group = vafType),
    position = position_stack(vjust = 0.5),
    fill = "white", color = "black", size = 3, label.size = 0
  ) +
  scale_fill_manual(values = plot_colors) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = NULL, y = "Proportion of disease mutation VAF",
       fill = "Disease mutation class") +
  theme_pubr() +
  theme(
        legend.position = "none",
        axis.line.x = element_blank())+
  guides(fill = guide_legend(direction = "vertical"))

p_cohortFraction


```

#### Cohort fractions — DD fathers back-inferred to age of conception
```{r cohortFractionConception, fig.width=7, fig.height=5}
# ------------------------------
# Driver-specific TwinsUK age model (posPath vafGroup)
# Fitted on panel vafSum so it is on the same scale as driverVAF (also panel posPath).
# Used to derive per-sample driver scale factors independently of the ClinDDlofCADD model.
# ------------------------------
if (twinsUKDataAvailable) {
  twins_model_driver <- lm(vafSum ~ Age, data = tukVafSums |> filter(vafGroup == "posPath"))
}

# ------------------------------
# All DD father samples shifted by the cohort mean paternal age gap.
# ClinDDlofCADD rate scales overall panel burden; posPath rate scales driver VAFs.
# Mosaic VAF unchanged. exome_countSum is count-based and does not scale with age.
# ------------------------------
mean_age_gap <- mean(burdenComp$paternalAgeGap)

dd_at_conception <- german_adjusted |>
  filter(source == "DD fathers") |>
  mutate(
    Age_conception    = Age - mean_age_gap,
    twinsUK_pred_conc = if (twinsUKDataAvailable) predict(twins_model, newdata = data.frame(Age = Age_conception))
                        else predict_from_fit_table(twinsUK_vafFits, "exome_vafSum_likelyDisease", Age_conception)$fit,
    age_scale_factor  = twinsUK_pred_conc / twinsUK_pred,
    vafSum_conc       = vafSum * age_scale_factor,
    exome_vafSum      = vafSum_conc + twinsUK_pred_conc * 0.565,
    exome_countSum    = round(countSum * (1 / exomePanelCorrection)),
    source            = "DD fathers (at conception)"
  )

# Driver VAFs scaled using the posPath-specific model rate
driverVAF_conc <- driverVAF |>
  filter(source == "DD fathers") |>
  left_join(dd_at_conception |> select(Sample_PD_ID, Age, Age_conception), by = join_by(Sample_PD_ID)) |>
  mutate(
    driver_pred_obs     = if (twinsUKDataAvailable) predict(twins_model_driver, newdata = data.frame(Age = Age))
                          else predict_from_fit_table(twinsUK_vafFits, "vafSum_posPath", Age)$fit,
    driver_pred_conc    = if (twinsUKDataAvailable) predict(twins_model_driver, newdata = data.frame(Age = Age_conception))
                          else predict_from_fit_table(twinsUK_vafFits, "vafSum_posPath", Age_conception)$fit,
    driver_scale_factor = driver_pred_conc / driver_pred_obs,
    driverVAF           = driverVAF * driver_scale_factor,
    source              = "DD fathers (at conception)"
  )

# ------------------------------
# Cohort fractions including the at-conception group
# ------------------------------
cohortFractionConception <- bind_rows(VafSumAdjusted43, dd_at_conception) |>
  select(Sample_PD_ID, source, exome_vafSum, exomeCount = exome_countSum) |>
  left_join(bind_rows(driverVAF, driverVAF_conc), by = join_by(Sample_PD_ID, source)) |>
  left_join(mosaicVAF |> select(Sample_PD_ID, mosaicVAF, mosaicCount), by = join_by(Sample_PD_ID)) |>
  mutate(otherVAF   = exome_vafSum - driverVAF,
         otherCount = exomeCount - driverCount) |>
  group_by(source) |>
  summarise(driverVAF   = sum(driverVAF,   na.rm = TRUE),
            driverCount = sum(driverCount, na.rm = TRUE),
            otherVAF    = sum(otherVAF,    na.rm = TRUE),
            otherCount  = sum(otherCount,  na.rm = TRUE),
            mosaicVAF   = sum(mosaicVAF,   na.rm = TRUE),
            mosaicCount = sum(mosaicCount, na.rm = TRUE),
            n_samples   = n_distinct(Sample_PD_ID),
            .groups = "drop") |>
  mutate(totalVAF = driverVAF + otherVAF + mosaicVAF) |>
  mutate(across(c(driverVAF, otherVAF, mosaicVAF), ~ .x / totalVAF)) |>
  pivot_longer(cols = c(driverVAF, otherVAF, mosaicVAF),
               names_to = "vafType",
               values_to = "vafProp") |>
  mutate(vafType = recode(vafType,
                          driverVAF = "Rare (known driver)",
                          otherVAF  = "Rare (other)",
                          mosaicVAF = "Mosaic")) |>
  mutate(source = factor(source, levels = c("DD fathers", "DD fathers (at conception)",
                                            "Apert fathers", "Controls"))) |>
  mutate(source_label = case_when(
    source == "DD fathers" ~
      paste0("DD fathers\nat sampling\n(n=", n_samples, ")"),
    source == "DD fathers (at conception)" ~
      paste0("DD fathers\nat conception\n(−", round(mean_age_gap, 1), " yrs, n=", n_samples, ")"),
    TRUE ~ paste0(source, "\n(n=", n_samples, ")")
  )) |>
  mutate(source_label = fct_reorder(source_label, as.integer(source))) |>
  filter(vafProp != 0)

# ------------------------------
# Sperm burden composition: observed vs mean-age-gap back-inferred conception
# ------------------------------
p_vafFractionConception <- cohortFractionConception |>
  filter(source %in% c("DD fathers", "DD fathers (at conception)")) |>
  ggplot(aes(x = source_label, y = vafProp, fill = vafType)) +
  geom_col(color = "black") +
  geom_label(
    aes(label = paste0(round(vafProp * 100, 1), "%"), group = vafType),
    position = position_stack(vjust = 0.5),
    fill = "white", color = "black", size = 3, label.size = 0
  ) +
  scale_fill_manual(values = plot_colors) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = NULL, y = "Proportion of disease mutation VAF",
       fill = "Disease mutation class") +
  theme_pubr() +
  theme(legend.position = "right",
        axis.line.x = element_blank())

p_vafFractionConception
```

#### Variant composition at conception: non-SCN1A cohort (n=110)
```{r scn1a_composition_conception, fig.width=5, fig.height=5}
# Regenerate the at-conception composition using only the 110 non-SCN1A DD fathers,
# reusing twins_model, twins_model_driver and mean_age_gap defined above.

dd_at_conception_110 <- german_adjusted |>
  filter(source == "DD fathers", Sample_PD_ID %in% dd_nonscn_ids) |>
  mutate(
    Age_conception    = Age - mean_age_gap,
    twinsUK_pred_conc = if (twinsUKDataAvailable) predict(twins_model, newdata = data.frame(Age = Age_conception))
                        else predict_from_fit_table(twinsUK_vafFits, "exome_vafSum_likelyDisease", Age_conception)$fit,
    age_scale_factor  = twinsUK_pred_conc / twinsUK_pred,
    vafSum_conc       = vafSum * age_scale_factor,
    exome_vafSum      = vafSum_conc + twinsUK_pred_conc * 0.565,
    exome_countSum    = round(countSum * (1 / exomePanelCorrection)),
    source            = "DD fathers (at conception)"
  )

driverVAF_conc_110 <- driverVAF |>
  filter(source == "DD fathers", Sample_PD_ID %in% dd_nonscn_ids) |>
  left_join(dd_at_conception_110 |> select(Sample_PD_ID, Age, Age_conception),
            by = join_by(Sample_PD_ID)) |>
  mutate(
    driver_pred_obs     = if (twinsUKDataAvailable) predict(twins_model_driver, newdata = data.frame(Age = Age))
                          else predict_from_fit_table(twinsUK_vafFits, "vafSum_posPath", Age)$fit,
    driver_pred_conc    = if (twinsUKDataAvailable) predict(twins_model_driver, newdata = data.frame(Age = Age_conception))
                          else predict_from_fit_table(twinsUK_vafFits, "vafSum_posPath", Age_conception)$fit,
    driver_scale_factor = driver_pred_conc / driver_pred_obs,
    driverVAF           = driverVAF * driver_scale_factor,
    source              = "DD fathers (at conception)"
  )

# Sampling-time composition for n=110 (for direct comparison)
dd_sampling_110 <- VafSumAdjusted43 |>
  filter(source == "DD fathers", Sample_PD_ID %in% dd_nonscn_ids) |>
  select(Sample_PD_ID, source, exome_vafSum, exomeCount = exome_countSum) |>
  left_join(driverVAF |> filter(source == "DD fathers", Sample_PD_ID %in% dd_nonscn_ids),
            by = join_by(Sample_PD_ID, source)) |>
  left_join(mosaic_nonscn, by = join_by(Sample_PD_ID, source)) |>
  mutate(otherVAF = exome_vafSum - driverVAF, otherCount = exomeCount - driverCount)

conception_comp_110 <- bind_rows(
  dd_sampling_110 |>
    summarise(driverVAF   = sum(driverVAF,   na.rm = TRUE),
              driverCount = sum(driverCount, na.rm = TRUE),
              otherVAF    = sum(otherVAF,    na.rm = TRUE),
              otherCount  = sum(otherCount,  na.rm = TRUE),
              mosaicVAF   = sum(mosaicVAF,   na.rm = TRUE),
              mosaicCount = sum(mosaicCount, na.rm = TRUE),
              n_samples   = n_distinct(Sample_PD_ID)) |>
    mutate(group = paste0("At sampling\n(n=", n_samples, ")")),
  bind_rows(
    dd_at_conception_110 |>
      select(Sample_PD_ID, source, exome_vafSum, exomeCount = exome_countSum),
  ) |>
    left_join(driverVAF_conc_110, by = join_by(Sample_PD_ID, source)) |>
    left_join(mosaic_nonscn, by = join_by(Sample_PD_ID)) |>
    mutate(otherVAF = exome_vafSum - driverVAF, otherCount = exomeCount - driverCount) |>
    summarise(driverVAF   = sum(driverVAF,   na.rm = TRUE),
              driverCount = sum(driverCount, na.rm = TRUE),
              otherVAF    = sum(otherVAF,    na.rm = TRUE),
              otherCount  = sum(otherCount,  na.rm = TRUE),
              mosaicVAF   = sum(mosaicVAF,   na.rm = TRUE),
              mosaicCount = sum(mosaicCount, na.rm = TRUE),
              n_samples   = n_distinct(Sample_PD_ID)) |>
    mutate(group = paste0("At conception\n(−", round(mean_age_gap, 1), " yrs, n=", n_samples, ")"))
) |>
  mutate(totalVAF = driverVAF + otherVAF + mosaicVAF) |>
  mutate(across(c(driverVAF, otherVAF, mosaicVAF), ~ .x / totalVAF)) |>
  pivot_longer(cols = c(driverVAF, otherVAF, mosaicVAF),
               names_to = "vafType", values_to = "vafProp") |>
  mutate(count = case_when(
    vafType == "driverVAF" ~ driverCount,
    vafType == "otherVAF"  ~ otherCount,
    vafType == "mosaicVAF" ~ mosaicCount
  )) |>
  mutate(vafType = recode(vafType,
                          driverVAF = "Rare (known driver)",
                          otherVAF  = "Rare (other)",
                          mosaicVAF = "Mosaic"),
         group = fct_inorder(group)) |>
  filter(vafProp != 0)

# Print proportions for results text
cat("=== Composition at sampling and conception (non-SCN1A, n=110) ===\n")
conception_comp_110 |>
  select(group, vafType, vafProp) |>
  mutate(vafProp = round(vafProp * 100, 1)) |>
  print(n = Inf)

ggplot(conception_comp_110, aes(x = group, y = vafProp, fill = vafType)) +
  geom_col(colour = "black") +
  geom_label(
    aes(label = paste0(round(vafProp * 100, 1), "%"), group = vafType),
    position = position_stack(vjust = 0.5),
    fill = "white", colour = "black", size = 3, label.size = 0
  ) +
  scale_fill_manual(values = plot_colors) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = NULL, y = "Proportion of disease mutation VAF",
       fill = "Disease mutation class") +
  theme_pubr() +
  theme(legend.position = "right", axis.line.x = element_blank(),
        plot.title = element_text(size = 10))
```

##### SupFig 5
```{r scn1a_supfig5, fig.height=5, fig.width=17}

conception_comp_110_plot <- conception_comp_110 |>
  mutate(source_label = str_replace(group, "At sampling",   "DD fathers\nat sampling") |>
                        str_replace("At conception", "DD fathers\nat conception")) |>
  mutate(source_label = fct_inorder(source_label)) |>
  select(source_label, vafType, vafProp, count)

p_vafFractionConception_110 <- p_vafFractionConception %+% conception_comp_110_plot

supfig5_110 <- pVafSumLikelyDisease + pExomeToPanelFraction + p_vafFractionConception_110 +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(size = 14, face = "bold"))

supfig5_110

supfig5_110
pdf(paste0(outputPath, "SupFig5_Path_nonSCN1A.pdf"),
    height = 5, width = 17, onefile = FALSE)
supfig5_110
dev.off()
```


#### Rate increase with age
```{r rateAge, fig.width=5, fig.height=10}

# ------------------------------
# Fit age models in TwinsUK exomes (or use the pre-computed fits - both are
# already saved in twinsUK_vafFits.tsv as "exome_vafSum_posPath" and
# "exome_vafSum_likelyDisease")
# ------------------------------
if (twinsUKDataAvailable) {
  model_exome_posPath <- tukVafSums |>
    filter(vafGroup == "posPath") |>
    lm(exome_vafSum ~ Age, data = _)

  model_exome_clin <- tukVafSums |>
    filter(vafGroup == "likelyDisease") |>
    lm(exome_vafSum ~ Age, data = _)
}

# ------------------------------
# Compute stacked curves
# ------------------------------
Age_seq <- seq(0, 80, length.out = 100)
newdat  <- tibble(Age = Age_seq)

if (twinsUKDataAvailable) {
  pred_posPath <- predict(model_exome_posPath, newdata = newdat)
  pred_clin    <- predict(model_exome_clin,    newdata = newdat)
} else {
  pred_posPath <- predict_from_fit_table(twinsUK_vafFits, "exome_vafSum_posPath", Age_seq)$fit
  pred_clin    <- predict_from_fit_table(twinsUK_vafFits, "exome_vafSum_likelyDisease", Age_seq)$fit
}

# Other = ClinLofDDCadd minus posPath (on the prediction scale)
pred_other <- pmax(pred_clin - pred_posPath, 0)

base_area_data <- tibble(
  Age = Age_seq,
  `Rare (known driver)` = pred_posPath,
  `Rare (other)` = pred_other
)

make_area_data <- function(base, mosaic_value) {
  base |>
    mutate(`Mosaic` = mosaic_value) |>
    pivot_longer(
      cols = c(`Rare (known driver)`, `Rare (other)`, `Mosaic`),
      names_to = "Category",
      values_to = "Value"
    ) |>
    mutate(Category = fct_relevel(Category, "Rare (known driver)", "Rare (other)", "Mosaic"))
}

plot_data_nomosaic <- make_area_data(base_area_data, mosaic_value = 0)
plot_data_mosaic5  <- make_area_data(base_area_data, mosaic_value = 0.05)

# ------------------------------
# Plots
# ------------------------------
plot_area_nomosaic <- ggplot(plot_data_nomosaic, aes(x = Age, y = Value, fill = Category)) +
  geom_area(alpha = 0.8) +
  scale_fill_manual(values = plot_colors) +
  labs(y = "", x = "Age") +
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.10)) +
  xlim(20, 80) +
  theme_pubr() + guides(fill = "none")

plot_area_mosaic5 <- ggplot(plot_data_mosaic5, aes(x = Age, y = Value, fill = Category)) +
  geom_area(alpha = 0.8) +
  scale_fill_manual(values = plot_colors) +
  labs(y = "                                        Theoretical risk of disease mutation", x = "Age") +
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.10)) +
  xlim(20, 80) +
  theme_pubr() + guides(fill = "none")
# ------------------------------
# Combine plots
# ------------------------------
(plot_area_nomosaic / plot_area_mosaic5) +
  plot_layout(guides = "collect")


```

#### Graphic
```{r graphicPlot}

img <- readPNG(paste0(dataPath, "GraphicFig4a_purple.png"))

img_ratio <- dim(img)[1] / dim(img)[2]   # height / width

p_graphic4a <- ggplot() +
  annotation_custom(
    rasterGrob(
      img,
      width  = unit(1, "npc"),
      height = unit(1, "npc"),
      interpolate = FALSE
    ),
    -Inf, Inf, -Inf, Inf
  ) +
  coord_fixed(ratio = img_ratio, expand = FALSE, clip = "off") +
  theme_void() +
  theme(
    panel.background = element_rect(fill = "white", colour = NA),
    plot.margin = margin(0, 0, 0, 0)
  )
p_graphic4a 
```

#### Legend
```{r legendPlot}



df_legend <- data.frame(
  label = factor(names(plot_colorsLegend), levels = names(plot_colorsLegend)),
  x = seq_along(plot_colorsLegend),
  y = 1
)

legend_grob <- cowplot::get_legend(
  ggplot(df_legend, aes(x = x, y = y, fill = label)) +
    geom_tile(width = 0.9, height = 0.9) +
    scale_fill_manual(values = plot_colorsLegend, name = NULL) +
    guides(fill = guide_legend(
      nrow = 1,
      byrow = TRUE
    )) +
    theme_void() +
    theme(
      legend.text = element_text(size = 9),
      legend.key.size = unit(0.5, "cm")
    )
)

legend_plot <- patchwork::wrap_elements(panel = legend_grob)
```

### Variant composition: SCN1A exclusion
#### Fig 4d composition: all vs SCN1A-only vs non-SCN1A
```{r scn1a_composition, fig.width=6, fig.height=5}
# Rebuild the cohortFraction stacked bar (Fig 4d style) for three DD father subgroups.
# Uses the same VafSumAdjusted43 / driverVAF / mosaicVAF inputs as the main figure.



build_cohort_frac <- function(ids, label) {
  VafSumAdjusted43 |>
    filter(source == "DD fathers", Sample_PD_ID %in% ids) |>
    select(Sample_PD_ID, source, exome_vafSum, exomeCount = exome_countSum) |>
    left_join(driverVAF |> filter(source == "DD fathers"),
              by = join_by(Sample_PD_ID, source)) |>
    left_join(mosaicVAF, by = join_by(Sample_PD_ID, source)) |>
    mutate(otherVAF   = exome_vafSum - driverVAF,
           otherCount = exomeCount   - driverCount) |>
    summarise(driverVAF   = sum(driverVAF,   na.rm = TRUE),
              driverCount = sum(driverCount, na.rm = TRUE),
              otherVAF    = sum(otherVAF,    na.rm = TRUE),
              otherCount  = sum(otherCount,  na.rm = TRUE),
              mosaicVAF   = sum(mosaicVAF,   na.rm = TRUE),
              mosaicCount = sum(mosaicCount, na.rm = TRUE),
              n_samples   = n_distinct(Sample_PD_ID)) |>
    mutate(source_label = paste0(label, "\n(n=", n_samples, ")"),
           totalVAF     = driverVAF + otherVAF + mosaicVAF) |>
    mutate(across(c(driverVAF, otherVAF, mosaicVAF), ~ .x / totalVAF)) |>
    pivot_longer(cols = c(driverVAF, otherVAF, mosaicVAF),
                 names_to = "vafType", values_to = "vafProp") |>
    mutate(count = case_when(
      vafType == "driverVAF" ~ driverCount,
      vafType == "otherVAF"  ~ otherCount,
      vafType == "mosaicVAF" ~ mosaicCount
    )) |>
    mutate(vafType = recode(vafType,
                            driverVAF = "Rare (known driver)",
                            otherVAF  = "Rare (other)",
                            mosaicVAF = "Mosaic"))
}

comp_scn1a <- bind_rows(
  build_cohort_frac(dd_all_ids,    "DD fathers (all)"),
  build_cohort_frac(scn1a_sperm_ids, "SCN1A only"),
  build_cohort_frac(dd_nonscn_ids, "Non-SCN1A")
) |>
  mutate(source_label = fct_inorder(source_label)) |>
  filter(vafProp != 0)

ggplot(comp_scn1a, aes(x = source_label, y = vafProp, fill = vafType)) +
  geom_col(colour = "black") +
  geom_label(
    aes(label = paste0("vars = ", count), group = vafType),
    position = position_stack(vjust = 0.5),
    fill = "white", colour = "black", size = 3, label.size = 0
  ) +
  scale_fill_manual(values = plot_colors) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = NULL, y = "Proportion of disease mutation VAF",
       fill = "Disease mutation class") +
  theme_pubr() +
  theme(legend.position = "right", axis.line.x = element_blank())
```



#### Figure 4 
```{r scn1a_fig4d, fig.height=5, fig.width=13}
# Replace the DD fathers rows in cohortFraction with the non-SCN1A equivalent.
nonscn_rows <- comp_scn1a |>
  filter(str_detect(as.character(source_label), "Non-SCN1A")) |>
  mutate(source_label = paste0("DD fathers\n(n=", n_samples, ")")) |>
  select(source_label, vafType, vafProp, count, n_samples)

cohortFraction_110 <- cohortFraction |>
  filter(!str_detect(as.character(source_label), "DD fathers")) |>
  select(source_label, vafType, vafProp, count, n_samples) |>
  bind_rows(nonscn_rows) |>
  mutate(source_label = fct_relevel(as.character(source_label),
                                    "Apert fathers\n(n=8)",
                                    "Controls\n(n=36)",
                                    "DD fathers\n(n=110)"))

p_cohortFraction_110 <- p_cohortFraction %+% cohortFraction_110

layoutFig4 <- "
AAA
BCC
BDG
BEG
BFF
"


figure4_110 <-
  (legend_plot +
     p_graphic4a + labs(tag = "A") +
     plot_spacer() +
     plot_area_nomosaic + labs(tag = "B") +
     plot_area_mosaic5 + labs(tag = "C") +
     plot_spacer() +
     p_cohortFraction_110 + labs(tag = "D")) +
  plot_layout(
    design  = layoutFig4,
    heights = c(0.2, 0, 2, 2, 0.1),
    widths  = c(8, 2, 3.5)
  ) &
  theme(plot.tag = element_text(size = 14, face = "bold"))

figure4_110

pdf(paste0(outputPath, "Fig4_diseaseRisk_nonSCN1A.pdf"),
    height = 5, width = 13, onefile = FALSE)
figure4_110
dev.off()
```

