---
title: "germanTrio_metadata"
author: "Matt Neville"
date: "17/01/2022"
output: html_document
---

# 1. Prep
## a) Paths/Libraries
```{r parameters}
#Input files on the farm
path <- "<hpc_scratch>/"
#Local output
localPath <- "<local_working_dir>/germanTrio/output/"
#Meta data/sample lists
metaPath <- "<local_working_dir>/germanTrio/meta/"
#Manuscript files
manuscriptPath <- "<local_working_dir>/germanTrio/Manuscript/"
#dnds folder
dndsPath <- "<hpc_scratch>/sperm/germanTrio_data/postNextflow/dNdS_basePairCov/"
# TwinsUK data
TUKpath <-  "<local_working_dir>/twinsUK_SpermPaperFiles/"

library(egg)
library(dndscv)
library(scales)
library(readxl)
library(openxlsx)
library(ggrepel)
library(patchwork)
library(stringi)
library(ggforce)
library(ggpubr)
library(ggpattern)
library(corrplot)
library(lme4)
library(bootpredictlme4)
suppressPackageStartupMessages(library(tidyverse))

# Shortcut load metadata
germanTrioSperm <- read_tsv(paste0(metaPath, "germanTrioSperm.tsv"), col_types = cols())

```

## b) Metadata processing

### Sperm samples
```{r germanTrioMeta}
germanTrioSpermLong <- 
  read_xlsx(paste0(metaPath,"Tables/targNanoSeqSpermCanApps.xlsx"), sheet = 1) |> select(Sample_PD_ID = Sample) |> 
  left_join(germanTrioMeta |> filter(tissue == "SPERM") |> 
    mutate(Sample_PD_ID = paste0(PD_ID, "_tds0001")), by = join_by(Sample_PD_ID)) |> 
  mutate(sampleFilter = "FULL") |> 
  distinct()
write_tsv(germanTrioSpermLong, paste0(metaPath, "germanTrioSpermLong.tsv"))

OxSpermMetaBasic <- read_tsv(paste0(metaPath, "OxSpermMetaBasic.tsv"), col_types = cols()) |> 
  arrange(Age) |> 
  select(-tissue, -ApertMut, -SupplierID, -ssID) |> 
  mutate(
    sampleFilter = "FULL",
    FAM_ID = paste0("FAM_A", row_number()),
    Cohort = "Apert fathers"
  )

germanTrioSperm <- germanTrioSpermLong |> 
  mutate(Cohort = "DD fathers") |> 
  select(PD_ID, Age, FAM_ID, Cohort, Sample_PD_ID, sampleFilter) |> 
  distinct() |> 
  # Remove double samples where one got replaced
  filter(!PD_ID %in% c("PD63866c", "PD63911d", "PD63789c")) |> 
  bind_rows(OxSpermMetaBasic) |> filter(sampleFilter == "FULL")

write_tsv(germanTrioSperm, paste0(metaPath, "germanTrioSperm.tsv"))
write_tsv(germanTrioSperm |> select(Sample_PD_ID, Cohort), paste0(path, "sperm/germanTrio_data/postNextflow/samplesgermanTrio.txt"), col_names = F)
write_tsv(germanTrioSperm |> select(Sample_PD_ID, Cohort) |> mutate(Cohort = str_replace(Cohort, " ", ".")), paste0(path, "sperm/germanTrio_data/samplesCovList.txt"), col_names = F)

```

### Diagnostic vars
```{r annotateDiagVars}
# Vars for Oxford fathers
diagnosticVarsOx <- read_tsv(paste0(metaPath, "OxSpermMetaBasic.tsv"), col_types = cols()) |> 
  mutate(
    Cohort = "Apert fathers",
    FAM_ID = paste0("FAM_A", row_number()),
    VAR_ID = if_else(ApertMut == "S252W", "VAR_A52", "VAR_A53"),
    chr = "10",
    start = if_else(ApertMut == "S252W", 123279677, 123279674),
    ref = if_else(ApertMut == "S252W", "G", "G"),
    alt = if_else(ApertMut == "S252W", "C", "C"),
    gene = "FGFR2",
    phasing = "unknown"
  ) |> 
  select(chr:phasing, VAR_ID, FAM_ID, Cohort) 

# Vars for german fathers from phasing table. Also add missing var from above
diagnosticVarsPos <- read_xlsx(paste0(metaPath, "Tables/phasing_information_2025-05-15.xlsx"), range = "A1:Q186") |> 
  bind_rows(read_xlsx(paste0(metaPath, "Tables/disease-variants_cohort2.xlsx")) |> mutate(START = as.numeric(START))) |> 
  # Add WGS calling info
  left_join(read_xlsx(paste0(metaPath, "Tables/phasing_information_Jan2026-RS.xlsx")), by = join_by(VAR_ID, IND_ID)) |> 
  # Remove non causal vars
  mutate(numericVar = as.numeric(str_sub(VAR_ID, 5,7))) |> filter(numericVar < 144 | numericVar > 184) |> 
  mutate(FAM_ID = str_replace(IND_ID, "IND", "FAM")) |> 
  # Harmonize phasing info
  mutate(phasing = if_else(status %in% c("maternal", "paternal"), status, "unknown")) |> 
  mutate(phasing = if_else(phasing == "unknown" & !is.na(Inheritance_Sanger), tolower(Inheritance_Sanger), phasing)) |> 
  mutate(phasing = if_else(FAM_ID %in% c("FAM_205", "FAM_179", "FAM_177"), "paternal", phasing)) |> 
  mutate(Cohort = "DD fathers") |> 
  select(VAR_ID, FAM_ID, Cohort, chr = CHROM, start = START, ref= REF, alt = ALT, gene = GENE,  
         called_TrioWGS = Called, missing_TrioWGS = `Missing Reason`, phasing) |> 
  # Ammend Missing/incomplete mutations
  mutate(start = if_else(FAM_ID == "FAM_205", 166896020, start),
         ref = if_else(FAM_ID == "FAM_205", "G", ref),
         alt = if_else(FAM_ID == "FAM_205", "GATATTCC", alt)) |> 
  mutate(chr = if_else(VAR_ID == "VAR_186", "1", chr),
         start = if_else(VAR_ID == "VAR_186", 1724694, start),
         ref = if_else(VAR_ID == "VAR_186", "C", ref),
         alt = if_else(VAR_ID == "VAR_186", "T", alt)) |> 
  mutate(chr = if_else(VAR_ID == "VAR_189", "X", chr),
         start = if_else(VAR_ID == "VAR_189", 133549118, start),
         ref = if_else(VAR_ID == "VAR_189", "G", ref),
         alt = if_else(VAR_ID == "VAR_189", "C", alt)) |> 
  mutate(chr = if_else(VAR_ID == "VAR_194", "8", chr),
         start = if_else(VAR_ID == "VAR_194", 61765556, start),
         ref = if_else(VAR_ID == "VAR_194", "G", ref),
         alt = if_else(VAR_ID == "VAR_194", "A", alt)) |> 
  mutate(chr = if_else(VAR_ID == "VAR_201", "X", chr),
         start = if_else(VAR_ID == "VAR_201", 53265644, start),
         ref = if_else(VAR_ID == "VAR_201", "GGCCGCATCATACCTTTCTGCTTCTCCA", ref),
         alt = if_else(VAR_ID == "VAR_201", "G", alt)) |> 
  mutate(chr = if_else(VAR_ID == "VAR_215", "9", chr),
         start = if_else(VAR_ID == "VAR_215", 135776225, start),
         ref = if_else(VAR_ID == "VAR_215", "C", ref),
         alt = if_else(VAR_ID == "VAR_215", "G", alt)) |> 
  # Ammened indels with slightly different coordinates
  mutate(chr = if_else(VAR_ID == "VAR_058", "2", chr),
       start = if_else(VAR_ID == "VAR_058", 230661314, start),
       ref = if_else(VAR_ID == "VAR_058", "GA", ref),
       alt = if_else(VAR_ID == "VAR_058", "G", alt),
       called_TrioWGS = if_else(VAR_ID == "VAR_058", "Yes", called_TrioWGS),
       missing_TrioWGS = if_else(VAR_ID == "VAR_058", NA, missing_TrioWGS)) |> 
  mutate(chr = if_else(VAR_ID == "VAR_193", "5", chr),
     start = if_else(VAR_ID == "VAR_193", 176696808, start),
     ref = if_else(VAR_ID == "VAR_193", "G", ref),
     alt = if_else(VAR_ID == "VAR_193", "A", alt),
     missing_TrioWGS = if_else(VAR_ID == "VAR_193", NA, missing_TrioWGS)) |> 
  mutate(chr = if_else(VAR_ID == "VAR_224", "9", chr),
     start = if_else(VAR_ID == "VAR_224", 135782135, start),
     ref = if_else(VAR_ID == "VAR_224", "T", ref),
     alt = if_else(VAR_ID == "VAR_224", "TC", alt)) |> 
  mutate(chr = if_else(VAR_ID == "VAR_232", "17", chr),
     start = if_else(VAR_ID == "VAR_232", 29560047, start),
     ref = if_else(VAR_ID == "VAR_232", "CAA", ref),
     alt = if_else(VAR_ID == "VAR_232", "C", alt)) |> 
  mutate(chr = if_else(VAR_ID == "VAR_004", "9", chr),
     start = if_else(VAR_ID == "VAR_004", 130416029, start),
     ref = if_else(VAR_ID == "VAR_004", "GTCC", ref),
     alt = if_else(VAR_ID == "VAR_004", "G", alt)) |> 
  mutate(chr = if_else(VAR_ID == "VAR_008", "8", chr),
     start = if_else(VAR_ID == "VAR_008", 61769134, start),
     ref = if_else(VAR_ID == "VAR_008", "TG", ref),
     alt = if_else(VAR_ID == "VAR_008", "T", alt)) |> 
  mutate(chr = if_else(VAR_ID == "VAR_124", "3", chr),
     start = if_else(VAR_ID == "VAR_124", 41275128, start),
     ref = if_else(VAR_ID == "VAR_124", "TATA", ref),
     alt = if_else(VAR_ID == "VAR_124", "T", alt)) |> 
  mutate(chr = if_else(VAR_ID == "VAR_132", "X", chr),
     start = if_else(VAR_ID == "VAR_132", 73960933, start),
     ref = if_else(VAR_ID == "VAR_132", "G", ref),
     alt = if_else(VAR_ID == "VAR_132", "GT", alt)) |> 
  mutate(chr = if_else(VAR_ID == "VAR_198", "21", chr),
     start = if_else(VAR_ID == "VAR_198", 34927287, start),
     ref = if_else(VAR_ID == "VAR_198", "CAGTT", ref),
     alt = if_else(VAR_ID == "VAR_198", "C", alt)) |> 
  mutate(chr = if_else(VAR_ID == "VAR_217", "X", chr),
     start = if_else(VAR_ID == "VAR_217", 153296811, start),
     ref = if_else(VAR_ID == "VAR_217", "G", ref),
     alt = if_else(VAR_ID == "VAR_217", "C", alt)) |> 
  mutate(chr = if_else(VAR_ID == "VAR_228", "2", chr),
     start = if_else(VAR_ID == "VAR_228", 166893043, start),
     ref = if_else(VAR_ID == "VAR_228", "TAAAAACAAC", ref),
     alt = if_else(VAR_ID == "VAR_228", "T", alt)) |> 
  mutate(alt = if_else(VAR_ID == "VAR_115", "AC", alt)) |> 
  mutate(alt = if_else(VAR_ID == "VAR_139", "GA", alt)) |> 
  mutate(ref = if_else(VAR_ID == "VAR_197", "TC", ref)) |> 
  mutate(ref = if_else(VAR_ID == "VAR_206", "C", ref),
         alt = if_else(VAR_ID == "VAR_206", "T", alt)) |> 
  mutate(ref = if_else(VAR_ID == "VAR_220", "TG", ref),
         alt = if_else(VAR_ID == "VAR_220", "T", alt)) |> 
  mutate(ref = if_else(VAR_ID == "VAR_229", "TG", ref),
         alt = if_else(VAR_ID == "VAR_229", "T", alt)) |> 
  mutate(ref = if_else(VAR_ID == "VAR_230", "T", ref),
         alt = if_else(VAR_ID == "VAR_230", "G", alt)) |> 
  mutate(called_TrioWGS = if_else(VAR_ID == "VAR_223", NA, called_TrioWGS),
       missing_TrioWGS = if_else(VAR_ID == "VAR_223", NA, missing_TrioWGS)) |> 
  # Annotating large deletions
  mutate(chr = if_else(VAR_ID == "VAR_210", "11", chr),
         start = if_else(VAR_ID == "VAR_210", NA, start),
         ref = if_else(VAR_ID == "VAR_210", "LargeDeletion", ref),
         alt = if_else(VAR_ID == "VAR_210", "LargeDeletion", alt)) |> 
  mutate(ref = if_else(VAR_ID == "VAR_086", "LargeDeletion", ref),
       alt = if_else(VAR_ID == "VAR_086", "LargeDeletion", alt)) |> 
  mutate(ref = if_else(VAR_ID == "VAR_012", "LargeDeletion", ref),
     alt = if_else(VAR_ID == "VAR_012", "LargeDeletion", alt),
     missing_TrioWGS = if_else(VAR_ID == "VAR_012", "Too Large variant", missing_TrioWGS)) |> 
  # Low vaf
  mutate(start = if_else(VAR_ID == "VAR_191", 26156978, start),
    called_TrioWGS = if_else(VAR_ID == "VAR_191", "No", called_TrioWGS),
    missing_TrioWGS = if_else(VAR_ID == "VAR_191", "VAF <0.3 (0.20)", missing_TrioWGS)) |> 
  mutate(start = if_else(VAR_ID == "VAR_192", 135778036, start),
    called_TrioWGS = if_else(VAR_ID == "VAR_192", "No", called_TrioWGS),
    missing_TrioWGS = if_else(VAR_ID == "VAR_192", "VAF <0.3 (0.21)", missing_TrioWGS)) |> 
  mutate(start = if_else(VAR_ID == "VAR_200", 52159789, start),
    called_TrioWGS = if_else(VAR_ID == "VAR_200", "No", called_TrioWGS),
    missing_TrioWGS = if_else(VAR_ID == "VAR_200", "VAF <0.3 (0.25)", missing_TrioWGS)) |> 
  mutate(start = if_else(VAR_ID == "VAR_022", 71064729, start),
    called_TrioWGS = if_else(VAR_ID == "VAR_022", "No", called_TrioWGS),
    missing_TrioWGS = if_else(VAR_ID == "VAR_022", "VAF <0.3 (0.25)", missing_TrioWGS)) |> 
  # Encode missing more clearly
  mutate(missing_TrioWGS = str_replace(missing_TrioWGS, "FF010", "In unmatched normal panel")) |> 
  mutate(missing_TrioWGS = str_replace(missing_TrioWGS, "VUM", "In unmatched normal panel")) |> 
  mutate(missing_TrioWGS = str_replace(missing_TrioWGS, "FF016", "Failed Pindel indel validation")) |> 
  mutate(missing_TrioWGS = str_replace(missing_TrioWGS, "parental VAF > 0", "Paternal VAF > 0")) |> 
  mutate(missing_TrioWGS = str_replace(missing_TrioWGS, "ASRD < 0.87 \\(0.85\\)", "Low variant-read alignment score")) |>
  mutate(missing_TrioWGS = str_replace(missing_TrioWGS, "7:101882729;p.L1262Q", "No Pindel call but visible in manual inspection")) |>
  mutate(missing_TrioWGS = str_replace(missing_TrioWGS, "Too Large variant", "Too large but visible in manual inspection")) |>
  # Add oxford apert cases
  bind_rows(diagnosticVarsOx) 

write_tsv(diagnosticVarsPos, paste0(metaPath, "diagnosticVarsPos.txt"))
```

# 2. Post Canapps processing
### Pull samples Nexflow batch output
```{bash ContamSummaries, eval=FALSE}

bsub -J "canAppsCopy_3606" \
  -o <hpc_scratch>/sperm/exomes/canAppsOutput/canAppsCopy_3606.%J.out \
  -e <hpc_scratch>/sperm/exomes/canAppsOutput/canAppsCopy_3606.%J.err \
  -q normal \
  -R 'select[mem>=200] rusage[mem=200] span[hosts=1]' -M 200 \
  "bash -lc '
# Set destination directory
dest_base=\"<hpc_scratch>/sperm/exomes/canAppsOutput\"

# Set source directory depends where pulling
#src_base=\"<hpc_references>/nst_links/live/3472\"
#src_base=\"<hpc_references>/nst_links/live/3707\"
# Loop over each PD* directory in the source
for pd_dir in \"\$src_base\"/PD*/; do
  pd_id=\$(basename \"\$pd_dir\")
  dest_dir=\"\$dest_base/\$pd_id\"

  # Create corresponding PD folder in destination
  mkdir -p \"\$dest_dir\"

  # Copy only specific files
  cp \"\$pd_dir\"/*.sample.effi.tsv \"\$dest_dir\" 2>/dev/null
  cp \"\$pd_dir\"/*.summary.tar.gz \"\$dest_dir\" 2>/dev/null
  cp \"\$pd_dir\"/*.tnanoseq.discardedvariants.vcf.gz \"\$dest_dir\" 2>/dev/null

  # Unzip summary tarballs inside each PD folder
  for tarfile in \"\$dest_dir\"/*.summary.tar.gz; do
    [ -e \"\$tarfile\" ] || continue
    tar -xvzf \"\$tarfile\" -C \"\$dest_dir\"
  done
done
'"

```

### Efficiency
```{bash ContamSummaries, eval=FALSE}
cd <hpc_scratch>/sperm/germanTrio_data/canAppsOutput
rm effi.txt

commonPath="<hpc_scratch>/sperm/germanTrio_data/canAppsOutput"

for sample in $(cut -f1 ../samplesgermanTrio.txt); do
  echo $sample
  
  # Shows contaminant level.
  seqReads=$(head -n 2 "$commonPath/$sample/$sample.sample.effi.tsv" | tail -n 1 | awk '{print $2}')
  dupRate=$(head -n 6 "$commonPath/$sample/$sample.sample.effi.tsv" | tail -n 1 | awk '{print $2}')
  f_eff=$(head -n 12 "$commonPath/$sample/$sample.sample.effi.tsv" | tail -n 1 | awk '{print $2}')
  
  echo -e "${sample}\t${seqReads}\t${dupRate}\t${f_eff}" >> effi.txt
done


```

### Burdens
```{bash burdens, eval=FALSE}
cd <hpc_scratch>/sperm/germanTrio_data/canAppsOutput/
rm burdens.tsv
for sample in `cut -f1 ../samplesgermanTrio.txt | grep "PD6"`;
  do echo $sample
  sed '/muts/d' <hpc_references>/nst_links/live/3472/$sample/$sample.tnanoseq.mut_burden.tsv | sed "s/$/\t$sample/" >> burdens.tsv
done
for sample in `cut -f1 ../samplesgermanTrio.txt | grep "PD5"`;
  do echo $sample
  sed '/muts/d' <hpc_references>/nst_links/live/3707/$sample/$sample.tnanoseq.mut_burden.tsv | sed "s/$/\t$sample/" >> burdens.tsv
done
```

### Variant Calls
```{bash varCalls, eval=FALSE}
#Merge together all calls
cd <hpc_scratch>/sperm/germanTrio_data/canAppsOutput/
rm mergedCalls.vcf
for sample in `cut -f1 ../samplesgermanTrio.txt | grep "PD6"`;
 do echo $sample
  gunzip -c <hpc_references>/nst_links/live/3472/$sample/$sample.tnanoseq.vcf.gz | sed '/#/d' | sed "s/$/\t$sample/" >> mergedCalls.vcf
done
for sample in `cut -f1 ../samplesgermanTrio.txt | grep "PD5"`;
 do echo $sample
  gunzip -c <hpc_references>/nst_links/live/3707/$sample/$sample.tnanoseq.vcf.gz | sed '/#/d' | sed "s/$/\t$sample/" >> mergedCalls.vcf
done

### Discarded Calls
cd <hpc_scratch>/sperm/germanTrio_data/canAppsOutput/
rm discardedCalls.vcf
for sample in `cut -f1 ../samplesgermanTrio.txt | grep "PD6"`;
 do echo $sample
  gunzip -c <hpc_references>/nst_links/live/3472/$sample/$sample.tnanoseq.discardedvariants.vcf.gz | sed '/#/d' | sed "s/$/\t$sample/" >> discardedCalls.vcf
done
for sample in `cut -f1 ../samplesgermanTrio.txt | grep "PD5"`;
 do echo $sample
  gunzip -c <hpc_references>/nst_links/live/3707/$sample/$sample.tnanoseq.discardedvariants.vcf.gz | sed '/#/d' | sed "s/$/\t$sample/" >> discardedCalls.vcf
done
```

### Coverage script
Save as runCov.sh
```{bash cov, eval=FALSE}
#!/usr/bin/env bash
# runCov.sh
# Usage:
#   bash runCov.sh <GROUP_NAME> <LIVE_FOLDER> <PANEL_FILE>
# Example:
#   bash runCov.sh targeted1 3769 panelV1_37_noChr.500pad.merged.bed

set -euo pipefail

if [ "$#" -ne 3 ]; then
  echo "Usage: bash runCov.sh <GROUP_NAME> <LIVE_FOLDER> <PANEL_FILE>"
  exit 1
fi

GROUP="$1"
LIVE_FOLDER="$2"
PANEL_FILE="$3"

BASE_DIR="<hpc_scratch>/sperm/germanTrio_data/postNextflow/"
cd "$BASE_DIR"

COV_SRC_BASE="<hpc_references>/nst_links/live/${LIVE_FOLDER}"
PANEL="<hpc_scratch>/genomicRegions/${PANEL_FILE}"

mkdir -p cov/{jobLogs,tmp,targCut,inPanel,sampleStats,final}

SAMPLE_LIST="cov/${GROUP}_samples.txt"
OUT_STATS="cov/final/${GROUP}.statsCoverage.tsv"
OUT_BED="cov/final/${GROUP}.totalCov.bed"
OUT_GZ="cov/final/${GROUP}.totalCov.bed.gz"

# Build sample list from tukTargSperm.tsv (col1=sample, col2=group)
awk -F'\t' -v g="$GROUP" '$2==g{print $1}' samplesCovList.txt > "$SAMPLE_LIST"

N_SAMPLES="$(wc -l < "$SAMPLE_LIST" | tr -d ' ')"

echo "Group: $GROUP"
echo "Live folder: $LIVE_FOLDER"
echo "Panel: $PANEL_FILE"
echo "Sample count: $N_SAMPLES"

if [ "$N_SAMPLES" -eq 0 ]; then
  echo "ERROR: No samples found for group '$GROUP'"
  exit 1
fi

rm -f "$OUT_STATS" "$OUT_BED" "$OUT_GZ" "${OUT_GZ}.tbi" 2>/dev/null || true

submitted=()

# -------------------------
# Stage 1: per-sample jobs
# -------------------------
while read -r sample; do
  out_tsv="cov/sampleStats/${sample}.tsv"
  targ="cov/targCut/${sample}.cov.bed"
  panel_bed="cov/inPanel/${sample}.inPanelCov.bed"

  if [ -s "$out_tsv" ]; then
    echo "Skipping (already done): $sample"
    continue
  fi

  echo "Submitting: $sample"

  bsub -J "${GROUP}_cov_${sample}" \
       -o "cov/jobLogs/${GROUP}_cov.${sample}.out" \
       -e "cov/jobLogs/${GROUP}_cov.${sample}.err" \
       -q normal \
       -R 'select[mem>=100] rusage[mem=100] span[hosts=1]' -M 100 \
    "set -euo pipefail
     module load bedtools2-2.29.0/ || true

     in_gz='${COV_SRC_BASE}/${sample}/${sample}.tnanoseq.cov.bed.gz'
     targ='${targ}'
     panel_bed='${panel_bed}'
     out_tsv='${out_tsv}'
     tmpdir='cov/tmp'

     gunzip -c \"\$in_gz\" \
       | awk -F'[;\\t]' '\$6 > 10' \
       | sed 's/;/\\t/g' \
       | cut -f1-3,6 \
       | sort -T \"\$tmpdir\" -k1,1 -k2,2n \
       > \"\$targ\"

     bedtools intersect -a \"\$targ\" -b '${PANEL}' > \"\$panel_bed\"

     bpUnique=\$(wc -l < \"\$targ\")
     bpTotal=\$(awk -F '\\t' '{s+=\$4} END {print s+0}' \"\$targ\")
     bpMedian=\$(datamash -s median 4 < \"\$targ\")

     bpUniquePanel=\$(wc -l < \"\$panel_bed\")
     bpTotalPanel=\$(awk -F '\\t' '{s+=\$4} END {print s+0}' \"\$panel_bed\")
     bpMedianPanel=\$(datamash -s median 4 < \"\$panel_bed\")

     printf '%s\\t%s\\t%s\\t%s\\t%s\\t%s\\t%s\\n' \
       '${sample}' \"\$bpUnique\" \"\$bpTotal\" \"\$bpMedian\" \
       \"\$bpUniquePanel\" \"\$bpTotalPanel\" \"\$bpMedianPanel\" \
       > \"\$out_tsv\"
    "

  submitted+=("$sample")
done < "$SAMPLE_LIST"

deps=""
if [ "${#submitted[@]}" -gt 0 ]; then
  for s in "${submitted[@]}"; do
    deps="${deps} && done(${GROUP}_cov_${s})"
  done
  deps="${deps# && }"
fi

# -------------------------
# Stage 2: finalize job (always runs)
# -------------------------
bsub -J "${GROUP}_finalize" \
     -o "cov/jobLogs/${GROUP}_finalize.%J.out" \
     -e "cov/jobLogs/${GROUP}_finalize.%J.err" \
     -q normal \
     -R 'select[mem>=200] rusage[mem=200] span[hosts=1]' -M 200 \
     ${deps:+-w "$deps"} \
  "set -euo pipefail
   module load bedtools2-2.29.0/ || true
   module load htslib/ || true

   : > \"${OUT_STATS}\"
   files=\"\"

   while read -r sample; do
     cat cov/sampleStats/\${sample}.tsv >> \"${OUT_STATS}\"
     files=\"\$files cov/targCut/\${sample}.cov.bed\"
   done < \"${SAMPLE_LIST}\"

   bedtools unionbedg -i \$files \
     | awk 'BEGIN{OFS=\"\t\"}{s=0; for(i=4;i<=NF;i++) s+=\$i; print \$1,\$2,\$3,s}' \
     > \"${OUT_BED}\"

   bgzip -f -c \"${OUT_BED}\" > \"${OUT_GZ}\"
   tabix -f -p bed \"${OUT_GZ}\"
  "

echo
echo "Submitted pipeline for $GROUP"
echo "Final outputs:"
echo "  $OUT_STATS"
echo "  $OUT_GZ"
```

### Run Cov
```{bash covRun, eval=FALSE}
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/cov/
bash runCov.sh DD.fathers 3472 germlineV2_hg19.bed
bash runCov.sh Apert.fathers 3707 germlineV1_hg19.bed
```

### TriNuc Background
```{bash triNucBackground, eval=FALSE}
#Merge together trinucs
cd <hpc_scratch>/sperm/germanTrio_data/canAppsOutput/
rm trinuc.txt
for sample in `cut -f1 ../samplesgermanTrio.txt`;
 do echo $sample
  cut -f1,2 <hpc_scratch>/sperm/germanTrio_data/canAppsOutput/$sample/${sample}_vs_${sample}.summary/${sample}_vs_${sample}.trint_counts_and_ratio2genome.tsv | sed '/ratio/d' | sed "s/$/\t$sample/" >> trinuc.txt
done

```

### Neat crams
```{bash cov, eval=FALSE}
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/neatCrams

for sample in `cut -f1 ../../samplesgermanTrio.txt | grep "PD6"`;
  do echo $sample
  ln -s <hpc_references>/nst_links/live/3472/$sample/$sample.sample.neat.cram .
  ln -s <hpc_references>/nst_links/live/3472/$sample/$sample.sample.neat.cram.crai .
done
for sample in `cut -f1 ../../samplesgermanTrio.txt | grep "PD5"`;
  do echo $sample
  ln -s <hpc_references>/nst_links/live/3707/$sample/$sample.sample.neat.cram .
  ln -s <hpc_references>/nst_links/live/3707/$sample/$sample.sample.neat.cram.crai .
done
```

# 3. QC
## a) Check burdens/Masked SNP counts
```{r maskedCounts}

#First snvs/mnvs then bind indels as they have different column orders
allCallsTarg <- read_tsv(paste0(path, "sperm/germanTrio_data/canAppsOutput/mergedCalls.vcf"), col_names = c("chr", "start", "ref", "alt", "FILTER","info", "Sample_PD_ID"), col_types = 'cd-cc-ccc') |>
  filter(str_detect(info, "nv")) |>
  separate(info, sep = ";", into = c("TRI","TIMES_CALLED","TYPE","DUPLEX_VAF","BAM_VAF","BAM_VAF_BQ10",NA,NA,NA,NA,"DUPLEX_COV","BAM_MUT","BAM_COV","BAM_MUT_BQ10","BAM_COV_BQ10",NA,"QPOS",NA,NA,NA,NA,NA), convert = T) |> 
  mutate_at(vars(TRI:QPOS), ~str_remove(., ".*=")) |> 
  mutate_at(vars(TIMES_CALLED, DUPLEX_VAF:BAM_COV_BQ10), as.numeric) |>
  # Add indels
  bind_rows(read_tsv(paste0(path, "sperm/germanTrio_data/canAppsOutput/mergedCalls.vcf"), col_names = c("chr", "start", "ref", "alt", "FILTER","info", "Sample_PD_ID"), col_types = 'cd-cc-ccc') |> 
    filter(!str_detect(info, "nv")) |>
    separate(info, sep = ";", into = c("TYPE","TIMES_CALLED","DUPLEX_VAF","BAM_VAF","BAM_VAF_BQ10",NA,NA,NA,NA,NA,"DUPLEX_COV","BAM_MUT","BAM_COV","BAM_MUT_BQ10","BAM_COV_BQ10", NA, "QPOS"), convert = T) |> 
    mutate_at(vars(TYPE:QPOS), ~str_remove(., ".*=")) |> 
    mutate_at(vars(TIMES_CALLED:BAM_COV_BQ10), as.numeric)) |> 
  left_join(germanTrioSperm, by = "Sample_PD_ID") |> 
  mutate(QPOS = ifelse(str_detect(QPOS, ","),
                       sapply(str_split(QPOS, ","), function(x) median(as.numeric(x))),
                       as.numeric(QPOS))) |> 
  mutate(FILTER = str_replace(FILTER, "MASKED", "NanoSeqMask")) |> 
  suppressWarnings()
write_tsv(allCallsTarg,  paste0(path, "sperm/germanTrio_data/postNextflow/contamFilter/allCallsTarg.tsv"))

#Read in QC metrics

seqMetrics <-  read_tsv(paste0(path, "sperm/germanTrio_data/canAppsOutput/effi.txt"),  col_names = c("Sample_PD_ID", "seqReads", "dupRate", "f_eff"), col_types = 'cddd')


#Calculate burdens from variant counts so that DNVs aren't double counted (DNV/MNVs merged post burden calc)
adjBurdens <- germanTrioSperm |> filter(sampleFilter != "dnaBelow0.05") |> select(Sample_PD_ID) |> 
  # Add muts
  left_join(allCallsTarg |> filter(FILTER == "PASS") |> filter(TYPE %in% c("snv", "dnv", "mnv")) |> 
  group_by(Sample_PD_ID) |> dplyr::summarize(muts = sum(TIMES_CALLED)), by = "Sample_PD_ID") |>  
  # Add indel burdens
  left_join(allCallsTarg |> filter(FILTER == "PASS") |> filter(TYPE %in% c("del", "ins")) |> 
    group_by(Sample_PD_ID) |> dplyr::summarize(indels = sum(TIMES_CALLED)), by = "Sample_PD_ID") |> 
  mutate(muts = replace_na(muts, 0)) |> mutate(indels = replace_na(indels, 0)) 

#Compare and merge with burdens in pipeline
burdens <- read_tsv(paste0(path, "sperm/germanTrio_data/canAppsOutput/burdens.tsv"), col_names = c("group", "muts",	"total",	"burden",	"burden_lci",	"burden_uci", "Sample_PD_ID"), col_types = 'cdddddc') |> 
  filter(group == "observed") |> select(-c(muts, burden)) |> 
  left_join(adjBurdens, by = "Sample_PD_ID") |> 
  mutate(burden = muts/total) |> 
  mutate(indelBurden = indels/total) |> 
  select(Sample_PD_ID, muts, indels, total, burden, indelBurden) 

#Find counts of masked and passed variants
maskedCounts <- allCallsTarg |> 
  mutate(FILTER = if_else(FILTER == "PASS", FILTER, "MASKED")) |> 
  mutate(TYPE = if_else(TYPE %in% c("snv", "dnv", "mnv"), "snv", "indel")) |> 
  group_by(Sample_PD_ID, FILTER, TYPE) |> summarize(count = n(), .groups = "drop") |> 
  pivot_wider(id_cols = Sample_PD_ID, names_from = c(FILTER,TYPE), names_sep = "_", values_from = count, values_fill = 0) |> 
  left_join(germanTrioSperm, by = "Sample_PD_ID") |> 
  left_join(burdens, by = "Sample_PD_ID") |> 
  left_join(seqMetrics, by = "Sample_PD_ID") |> 
  mutate(maskedProportion = MASKED_snv/PASS_snv) |> 
  mutate(maskedProportionIndel = MASKED_indel/PASS_indel) |> 
  mutate(above5 = if_else(maskedProportion > 5, T,F)) |> 
  mutate(label = if_else((maskedProportion > 5 | burden > 1e-7), str_remove(FAM_ID, "FAM_"), "")) |> 
  mutate(concern = if_else((maskedProportion > 5 | burden > 1e-7), T, F))

```


## b) Filter variants
### Implement filters
```{r varContam, fig.width= 12, fig.height=12}

# List of noisy sites in discarded calls file to exclude in call set
# anyCounts <- read_tsv(paste0(path, "sperm/germanTrio_data/canAppsOutput/discardedCalls.vcf"), col_names = c("chr", "start", "ref", "alt", "FILTER","info", "Sample_PD_ID"), col_types = 'cd-cc-ccc') |>
#   group_by(chr,start) |> summarise(varCount = n(), .groups = "drop") |> 
  # filter(varCount > 2) |> 
  # bind_rows(read_tsv(paste0(path, "sperm/germanTrio_data/canAppsOutput/discardedCalls.vcf"), col_names = c("chr", "start", "ref", "alt", "FILTER","info", "Sample_PD_ID"), col_types = 'cd-cc-ccc') |> 
  # filter(!str_detect(info, ";N=1;")) |>
  # group_by(chr,start) |> summarise(varCount2 = n(), .groups = "drop")) |> 
  # distinct(chr,start) |> 
  # mutate(varCount = 1)
# write_tsv(anyCounts, paste0(path, "sperm/germanTrio_data/canAppsOutput/discardedCallsSimple.vcf"))
anyCounts <- read_tsv(paste0(path, "sperm/germanTrio_data/canAppsOutput/discardedCallsSimple.vcf"), col_types = cols())

# Allele freq threshold to use:
bamThreshold = 0.3
duplexThreshold = 0.3
minCoverage = 11

allCallsFilter <- allCallsTarg |> 
  mutate(class = if_else(TYPE == "snv", "snv", "indel")) |> 
  # VAF/coverage cutoffs
  mutate(FILTER = if_else(FILTER == "PASS" & (BAM_VAF >= bamThreshold | DUPLEX_VAF >= duplexThreshold | DUPLEX_COV <= minCoverage), "alleleFreqThreshold", FILTER, missing = "NanoSeqMask")) |> 
  # Cohort germline vars. Small effect in this cohort so only removing 2 SNP sites common in cohort (seen as germline vars in >2 indivs)
  left_join(anyCounts, by = c("chr", "start"))  |> mutate(varCount = replace_na(varCount, 0)) |> 
  mutate(FILTER = if_else(FILTER == "PASS" & (varCount > 0), "cohortMask", FILTER)) |> 
  # Manually filter recurrent artefacts
  mutate(FILTER = if_else(start == 34927486 & chr == "21", "manualMask", FILTER)) |> 
  mutate(FILTER = if_else(start == 13409946 & chr == "19", "manualMask", FILTER)) |> 
  mutate(FILTER = if_else(start == 1952721 & chr == "4", "manualMask", FILTER)) |> 
  mutate(FILTER = if_else(start == 102516324 & chr == "14", "manualMask", FILTER)) |> 
  mutate(FILTER = if_else(start == 86584359 & chr == "9", "manualMask", FILTER)) |> 
  mutate(FILTER = if_else(start == 72831388 & chr == "16", "manualMask", FILTER)) |> 
  mutate(FILTER = if_else(start == 48982774 & chr == "2", "manualMask", FILTER)) |> 
  mutate(FILTER = if_else(start == 72822586 & chr == "16", "manualMask", FILTER)) |> 
  mutate(FILTER = if_else(start == 133184690 & chr == "8", "manualMask", FILTER)) |> 
  mutate(FILTER = if_else(start == 148338 & chr == "16", "manualMask", FILTER)) |> 
  mutate(FILTER = if_else(start == 114724469 & chr == "10", "manualMask", FILTER)) |> 
  mutate(FILTER = if_else(start == 61734242 & chr == "8", "manualMask", FILTER)) |> 
  mutate(FILTER = if_else(start == 2039773 & chr == "9", "manualMask", FILTER)) |> 
  mutate(FILTER = if_else(start == 38946606 & chr == "19", "manualMask", FILTER)) |>
  mutate(FILTER = if_else(start == 23892380 & chr == "15", "manualMask", FILTER)) |>
  mutate(FILTER = if_else(start == 232372410 & chr == "2", "manualMask", FILTER)) |>
  mutate(FILTER = if_else(str_detect(as.character(start), "7299") & chr == "16", "manualMask", FILTER)) |> # Bad ZFHX3 region
  mutate(FILTER = if_else(start %in% c(120539414,120539315,120539316) & chr == "1", "manualMask", FILTER)) |> 
  group_by(chr, start) |> 
  mutate(FILTER = if_else(FILTER == "PASS" & any(FILTER != "PASS"), "manualMask", FILTER)) |> 
  # Diagnostic vars treat separately
  mutate(FILTER = if_else(start == 116420191 & chr == "12" & PD_ID == "PD63954c", "diagVar", FILTER)) |> 
  mutate(FILTER = if_else(start == 13135994 & chr == "19" & PD_ID == "PD63888c", "diagVar", FILTER)) |> 
  mutate(FILTER = if_else(start == 14280495 & chr == "5" & PD_ID == "PD63917c", "diagVar", FILTER)) |> 
  mutate(FILTER = if_else(start == 3155143 & chr == "6" & PD_ID == "PD63911c", "diagVar", FILTER)) |> 
  mutate(FILTER = if_else(start == 148526832 & chr == "7" & PD_ID == "PD68811b", "diagVar", FILTER)) |> 
  mutate(FILTER = if_else(start == 166896020 & chr == "2" & PD_ID == "PD68831b", "diagVar", FILTER)) |> 
  mutate(FILTER = if_else(start == 166909427 & chr == "2" & PD_ID == "PD68787b", "diagVar", FILTER)) |> 
  mutate(FILTER = if_else(start == 9928084 & chr == "16" & PD_ID == "PD68809b", "diagVar", FILTER)) |> 
  # CA issue in Apert samples
  mutate(FILTER = if_else(paste0(str_sub(TRI, 2,2), str_sub(TRI, 5,5)) %in% c("CA", "GT") & 
                        BAM_VAF > 0.002 & str_detect(Sample_PD_ID, "PD5") & FILTER == "PASS", "C>A_enzyme", FILTER)) |> 
  mutate(FILTER = str_remove(FILTER, ";shearwater")) |> 
  ungroup() 
 
table(allCallsFilter$FILTER)

# Filter to preset cutoff
filteredVars <- allCallsFilter |> 
  filter(FILTER == "PASS") |> 
  select(chr:alt,TRI, TYPE, TIMES_CALLED, DUPLEX_COV, DUPLEX_VAF, BAM_MUT, BAM_COV, BAM_VAF, Sample_PD_ID:sampleFilter)

write_tsv(filteredVars, paste0(path, "sperm/germanTrio_data/postNextflow/contamFilter/snpFilterVars.tsv"))

repetitiveSites <- filteredVars |> group_by(chr,start, TYPE) |> summarise(count = n(), .groups = "drop", mean = mean(TIMES_CALLED), median = median(TIMES_CALLED)) |> 
  arrange(-count) |> filter(count > 1) |> 
  left_join(read_tsv(paste0(localPath, "data/allCallsFilterVarsVEP.tsv"), col_types = cols()) |> 
              select(chr, start, TYPE, csq:ClinVar) |> distinct(), by = join_by(chr, start, TYPE)) 
```

#### Bam VAF Plot
```{r bamVAF, fig.height=7, fig.width=9}
vaf_colours <- c("NanoSeqMask" = "#fb8072",
                 "cohortMask" = "#fdb462",
                 "manualMask" = "cadetblue",
                 "dbsnp" = "#addd8e",
                 "shearwater" = "lightpink",
                 "NEI_IND" = "#91003f",
                 "alleleFreqThreshold" = "lightyellow",
                 "C>A_enzyme" = "darkorchid3",
                 "PASS" = "#80b1d3")

# Calculate the counts for each filter category
filter_counts_initial <- allCallsTarg |>
  count(FILTER)

filter_counts_cohort <- allCallsFilter |>
  count(FILTER)

# Create new labels with counts for the legend
vaf_labels_initial <- setNames(paste0(filter_counts_initial$FILTER, " (n=", filter_counts_initial$n, ")"), filter_counts_initial$FILTER)
vaf_labels_cohort <- setNames(paste0(filter_counts_cohort$FILTER, " (n=", filter_counts_cohort$n, ")"), filter_counts_cohort$FILTER)

# Updated plot for initialVAF
p_initialVAF <- allCallsTarg |> 
  mutate(TYPE = if_else(TYPE == "snv", "Substitutions", "Indels")) |> 
  mutate(TYPE = factor(TYPE, levels = c("Substitutions", "Indels"), ordered = TRUE)) |> 
  mutate(FILTER = if_else(FILTER == "dbsnp;shearwater", "dbsnp", FILTER)) |>
  mutate(BAM_VAF = if_else(BAM_VAF == 0, BAM_VAF_BQ10, BAM_VAF)) |> 
  mutate(BAM_VAF = if_else(BAM_VAF == 0, DUPLEX_VAF, BAM_VAF)) |> 
  mutate(FILTER = factor(FILTER, levels = names(vaf_colours), ordered = TRUE)) |> 
  ggplot(aes(BAM_VAF, after_stat(count), fill = FILTER)) +
  scale_fill_manual(values = vaf_colours, labels = vaf_labels_initial) +
  scale_x_log10(limits = c(6e-5, NA)) +
  geom_density(position = "stack") +
  ylab("Density") +
  guides(fill=guide_legend(title="Pipeline filters")) +
  facet_wrap(~TYPE, scales = "free_y", ncol = 2) +
  theme_pubr() +
  theme(legend.position = "right", strip.background = element_blank(), strip.text = element_text(size = 12))

# Updated plot for cohortMaskVAF
p_cohortMaskVAF <- allCallsFilter |> filter(FILTER != "diagVar") |> 
  mutate(TYPE = if_else(TYPE == "snv", "Substitutions", "Indels")) |> 
  mutate(TYPE = factor(TYPE, levels = c("Substitutions", "Indels"), ordered = TRUE)) |> 
  mutate(BAM_VAF = if_else(BAM_VAF == 0, BAM_VAF_BQ10, BAM_VAF)) |> 
  mutate(BAM_VAF = if_else(BAM_VAF == 0, DUPLEX_VAF, BAM_VAF)) |> 
  mutate(FILTER = factor(FILTER, levels = names(vaf_colours), ordered = TRUE)) |> 
  ggplot(aes(BAM_VAF, after_stat(count), fill = FILTER)) +
  scale_fill_manual(values = vaf_colours, labels = vaf_labels_cohort) +
  scale_x_log10(limits = c(6e-5, NA)) +
  geom_density(position = "stack") +
  guides(fill=guide_legend(title="Full filters")) +
  ylab("Density") +
  facet_wrap(~TYPE, scales = "free_y", ncol = 2) + 
  theme_pubr() +
  theme(legend.position = "right", strip.background = element_blank(), strip.text = element_text(size = 12))

# Combine the plots
p_initialVAF / p_cohortMaskVAF + plot_annotation(tag_levels = "A")

png(paste0(localPath, 'pQC_variantFilters.png'), height=10, width=12)
p_initialVAF/p_cohortMaskVAF + plot_annotation(tag_levels = "A")
dev.off()

```


### PostFilter Contam
```{r postFilterContam, fig.height=7, fig.width=7}

burdensPost <- burdens |> 
  left_join(filteredVars |> 
              filter(TYPE %in% c("snv", "dnv", "mnv")) |> 
              group_by(Sample_PD_ID) |> summarize(mutsPost = sum(TIMES_CALLED), mutsPostSingle = n()), by = "Sample_PD_ID") |> 
  left_join(filteredVars |> 
              filter(TYPE %in% c("del", "ins")) |> 
              group_by(Sample_PD_ID) |> summarize(indelsPost = sum(TIMES_CALLED), indelsPostSingle = n()) , by = "Sample_PD_ID") |> 
  left_join(filteredVars |> filter(BAM_VAF < 0.01 & DUPLEX_VAF < 0.1) |> 
              filter(TYPE %in% c("snv", "dnv", "mnv")) |> 
              group_by(Sample_PD_ID) |> summarize(mutsPost1percent = sum(TIMES_CALLED)), by = "Sample_PD_ID") |> 
  left_join(filteredVars |> filter(BAM_VAF < 0.01 & DUPLEX_VAF < 0.1) |> 
              filter(TYPE %in% c("del", "ins")) |> 
              group_by(Sample_PD_ID) |> summarize(indelsPost1percent = sum(TIMES_CALLED)) , by = "Sample_PD_ID") |> 
  mutate(across(c(indelsPost, indelsPostSingle, indelsPost1percent, mutsPost, mutsPostSingle, mutsPost1percent), ~ replace_na(.x, 0))) |> 
  mutate(burdenPost = mutsPost/total) |> 
  mutate(burdenPostSingle = mutsPostSingle/total) |> 
  mutate(burdenPost1percent = mutsPost1percent/total) |> 
  mutate(burdenIndelPost = indelsPost/total) |> 
  mutate(burdenIndelPostSingle = indelsPostSingle/total) |> 
  mutate(burdenIndelPost1percent = indelsPost1percent/total) 

maskedCountsPost <- allCallsFilter |>
  mutate(FILTER = if_else(FILTER == "PASS", FILTER, "MASKED")) |> 
  mutate(TYPE = if_else(TYPE %in% c("snv", "dnv", "mnv"), "snv", "indel")) |> 
  group_by(Sample_PD_ID, FILTER, TYPE) |> summarize(count = n(), .groups = "drop") |> 
  pivot_wider(id_cols = Sample_PD_ID, names_from = c(FILTER,TYPE), names_sep = "_", values_from = count, values_fill = 0) |> 
  left_join(germanTrioSperm, by = "Sample_PD_ID") |> 
  left_join(burdensPost, by = "Sample_PD_ID") |> 
  left_join(seqMetrics, by = "Sample_PD_ID") |> 
  mutate(maskedProportionPost = MASKED_snv/PASS_snv) |> 
  mutate(maskedProportionPostIndel = MASKED_indel/PASS_indel) |> 
  left_join(maskedCounts |> select(Sample_PD_ID, maskedProportion, maskedProportionIndel), by = "Sample_PD_ID") |> 
  left_join(filteredVars |> mutate(above1 = if_else(TIMES_CALLED > 1, 1, 0)) |> group_by(Sample_PD_ID) |> summarize(above1Vars = sum(above1))) |> 
  mutate(above1Prop = above1Vars/mutsPostSingle) |> select(-above1Vars) |> 
  mutate(above5 = if_else(maskedProportion > 5, T,F)) |> 
  mutate(label = if_else((maskedProportion > 5 | burden > 1e-7), str_remove(FAM_ID, "FAM_"), "")) |> 
  mutate(concern = if_else((maskedProportion > 5 |  burdenPost > 1e-7), T, F))

```

### Corrected burdens
Using shorted version of code from NanoSeq pipeline to correct burdens for trinucleotide composition (https://github.com/cancerit/NanoSeq/blob/develop/R/nanoseq_results_plotter.R)
```{r burdensCorrected}
#Write out filtered and corrected burdens
burdensCorrected0 <- burdensPost |> 
  select(Sample_PD_ID, total, muts = mutsPost,  burden = burdenPost, indels = indelsPost, indelBurden = burdenIndelPost, burdenIndelPost1percent)

# Background full genome trinuc composition
genome_counts <- tibble(trinuc = c("ACA", "ACC", "ACG", "ACT", "ATA", "ATC", "ATG", "ATT", "CCA", "CCC", "CCG", "CCT", "CTA", "CTC", "CTG", "CTT", "GCA", "GCC", "GCG", "GCT", "GTA", "GTC", "GTG", "GTT", "TCA", "TCC", "TCG", "TCT", "TTA", "TTC", "TTG", "TTT"),
                genomeCount = c(115415924, 66550070, 14381094, 92058521, 117976329, 76401029, 105094288, 142651503, 105547494, 75238490,
    15801067, 101628641, 73791042, 96335416, 115950255, 114180747, 82414099, 68090507, 13621251, 80004082,
    64915540, 54055728, 86012414, 83421918, 112085858, 88336615, 12630597, 126566213, 119020255, 112827451,
    108406418, 219915599), totalCount = 2861326455) |> 
  mutate(genomeFraction = genomeCount/totalCount)

# Trinucleotides covered per sample and ratio to full genome
tri_bg <- read_tsv(paste0(path, "sperm/germanTrio_data/canAppsOutput/trinuc.txt"), col_types = cols(), col_names = c("trinuc", "trinucCount", "Sample_PD_ID")) |> 
  left_join(genome_counts, by = join_by(trinuc)) |> left_join(burdensCorrected0 |> select(Sample_PD_ID, total), by = join_by(Sample_PD_ID)) |> 
  mutate(sampleFraction = trinucCount/total) |> 
  mutate(ratioToGenome = sampleFraction/genomeFraction) |> 
  select(Sample_PD_ID, trinuc, ratioToGenome)

# Observed Trinuc mutations adjusted by ratio of background to genome
burdensCorr <- filteredVars |> drop_na(TRI) |> filter(TRI != "NA") |> 
  group_by(TRI, Sample_PD_ID) |> summarize(count = sum(TIMES_CALLED), .groups = "drop") |> 
  mutate(trinuc = str_sub(TRI, 1, 3)) |> 
  left_join(tri_bg, by = join_by(Sample_PD_ID, trinuc)) |> 
  mutate(trint_onto_genome = count/ratioToGenome) |> 
  group_by(Sample_PD_ID) |> summarize(mutsCorr = sum(trint_onto_genome)) 

# Observed Trinuc mutations adjusted by ratio of background to genome
burdensCorr1percent <- filteredVars |> filter(BAM_VAF < 0.01 & DUPLEX_VAF < 0.1) |> drop_na(TRI) |> filter(TRI != "NA") |> 
  group_by(TRI, Sample_PD_ID) |> summarize(count = sum(TIMES_CALLED), .groups = "drop") |> 
  mutate(trinuc = str_sub(TRI, 1, 3)) |> 
  left_join(tri_bg, by = join_by(Sample_PD_ID, trinuc)) |> 
  mutate(trint_onto_genome = count/ratioToGenome) |> 
  group_by(Sample_PD_ID) |> summarize(mutsCorr1perc = sum(trint_onto_genome)) 

burdensCorrected <- burdensCorrected0 |> 
  left_join(burdensCorr1percent, by = join_by(Sample_PD_ID)) |> mutate(burdenCorr1percent = mutsCorr1perc/total) |> 
  left_join(burdensCorr, by = join_by(Sample_PD_ID)) |> mutate(burdenCorr = mutsCorr/total) 

```

#### SNP burdens
```{r contam, fig.height=7, fig.width=10}

pPostMaskedProp <- ggplot(maskedCountsPost, aes(maskedProportion, maskedProportionPost, label = label)) +
  geom_abline(slope=1, linetype = "dashed") +
  geom_point(size=3, pch=21, col='white', stroke=0.25, fill = "maroon") +
  geom_text_repel(colour = "grey60",
                  arrow = arrow(angle = 0, length = unit(0.02, "npc")),
                box.padding = 1, max.overlaps = Inf) +
  ylim(c(0,15)) +
  xlim(c(0,15)) +
  xlab("Pre Filter (masked SNPs / passed variants)") +
  ylab("Post Filter (masked SNPs / passed variants)") +
  theme_minimal()   +
  theme(legend.position = "none")
pPostBurdenProp <- ggplot(maskedCountsPost, aes(x = burden, y = burdenPost, label = label)) +
  geom_abline(slope=1, linetype = "dashed") +
  geom_point(size=3, pch=21, col='white', stroke=0.25, fill = "maroon") +
  geom_text_repel(colour = "grey60",
                  arrow = arrow(angle = 0, length = unit(0.02, "npc")),
                box.padding = 1, max.overlaps = Inf) +
  xlab("Pre Filter Burden") +
  ylab("Post Filter Burden") +
  theme_minimal()  +
  theme(legend.position = "none")

ggarrange(pPostMaskedProp, pPostBurdenProp)

allpostBurden <- maskedCountsPost |> 
                  pivot_longer(cols = c(burden, burdenPost), names_to = "burdenType", values_to = "burdens") 
pAllPostBurden <- allpostBurden |> 
  ggplot(aes(x =  Age, y = burdens, label = label)) + 
  geom_smooth(method='lm', alpha = 0.2) +
  geom_point(size=3, pch=21, col='white', stroke=0.25, fill = "maroon") +
  geom_text_repel(colour = "grey60",
                  arrow = arrow(angle = 0, length = unit(0.02, "npc")),
                box.padding = 1, max.overlaps = Inf) +
  xlab('Age (years)') + 
  ylab('Mutations per bp') + 
  theme_minimal() + 
  # ylim(0,1.25e-7) +
  facet_wrap(~burdenType, nrow = 2)
pAllPostBurden

((pPostMaskedProp / pPostBurdenProp  + plot_layout(guides = 'auto')) | pAllPostBurden) + plot_layout(guides = 'collect')

pdf(paste0(localPath, 'pQC_prePostFilterComp.pdf'), height=7, width=10, onefile = F)
((pPostMaskedProp / pPostBurdenProp  + plot_layout(guides = 'auto')) | pAllPostBurden) + plot_layout(guides = 'collect')
dev.off()

```

#### Indel burdens
```{r contam, fig.height=7, fig.width=10}

pPostMaskedPropIndel <- ggplot(maskedCountsPost, aes(maskedProportionIndel, maskedProportionPostIndel,  label = label)) +
  geom_abline(slope=1, linetype = "dashed") +
  geom_point(size=3, pch=21, col='white', stroke=0.25, fill = "maroon") +
  geom_text_repel(colour = "grey60",
                  arrow = arrow(angle = 0, length = unit(0.02, "npc")),
                box.padding = 1, max.overlaps = Inf) +
  ylim(c(0,15)) +
  xlim(c(0,15)) +
  xlab("Pre Filter (masked Indel / passed Indel)") +
  ylab("Post Filter (masked Indel / passed Indel)") +
  theme_minimal() +
  theme(legend.position = "none")
pPostBurdenPropIndel <- ggplot(maskedCountsPost, aes(x = indelBurden, y = burdenIndelPost,  label = label)) +
  geom_abline(slope=1, linetype = "dashed") +
  geom_point(size=3, pch=21, col='white', stroke=0.25, fill = "maroon") +
  geom_text_repel(colour = "grey60",
                  arrow = arrow(angle = 0, length = unit(0.02, "npc")),
                box.padding = 1, max.overlaps = Inf) +
  xlab("Pre Filter Indel Burden") +
  ylab("Post Filter Indel Burden") +
  theme_minimal() +
  theme(legend.position = "none")

ggarrange(pPostMaskedPropIndel, pPostBurdenPropIndel)

allpostBurdenIndel <- maskedCountsPost |> dplyr::rename(indelBurdenPost = burdenIndelPost) |> 
                  pivot_longer(cols = c(indelBurden, indelBurdenPost), names_to = "burdenType", values_to = "burdens") |> 
  filter(PD_ID != "PD50495b")
pAllPostBurdenIndel <- allpostBurdenIndel |> 
  ggplot(aes(x =  Age, y = burdens,  label = label)) + 
  geom_smooth(method='lm', alpha = 0.2) +
  geom_point(size=3, pch=21, col='white', stroke=0.25, fill = "maroon") +
  geom_text_repel(colour = "grey60",
                  arrow = arrow(angle = 0, length = unit(0.02, "npc")),
                box.padding = 1, max.overlaps = Inf) +
  xlab('Age (years)') + 
  ylab('Mutations per bp') + 
  theme_minimal() + 
  ylim(0,NA) +
  facet_wrap(~burdenType, nrow = 2)
pAllPostBurdenIndel

((pPostMaskedPropIndel / pPostBurdenPropIndel  + plot_layout(guides = 'auto')) | pAllPostBurdenIndel) + plot_layout(guides = 'collect')

pdf(paste0(localPath, 'pQC_prePostFilterCompIndel.pdf'), height=7, width=10, onefile = F)
((pPostMaskedPropIndel / pPostBurdenPropIndel  + plot_layout(guides = 'auto')) | pAllPostBurdenIndel) + plot_layout(guides = 'collect')
dev.off()

```

## c) Filter Samples
#### BAM cov summary
```{bash covSumBAM, eval=FALSE}
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow

mkdir -p cov/rawBamDepth/jobs
mkdir -p cov/rawBamDepth/out
mkdir -p cov/rawBamDepth/tmp

cd <hpc_scratch>/sperm/germanTrio_data/postNextflow
cat << 'EOF' > cov/rawBamDepth/calc_raw_depth.sh
#!/bin/bash
set -euo pipefail

sample=$1
bam=$2
TARGET_BED=$3
out=$4

if [ ! -s "$bam" ]; then
  echo -e "${sample}\tNA\tNA\tBAM_NOT_FOUND\t${bam}" > "$out"
  exit 0
fi

# Mean on-target depth (including zero-depth bases)
mean=$(samtools depth -a -b "$TARGET_BED" "$bam" \
  | awk '{sum+=$3; n+=1} END {if(n>0) printf("%.6f", sum/n); else printf("NA")}')

# Also record totals for QC
stats=$(samtools depth -a -b "$TARGET_BED" "$bam" \
  | awk '{sum+=$3; n+=1} END {printf("%d\t%d", sum, n)}')

# sample  meanDepth  depthSum  nBases  status  bamPath
echo -e "${sample}\t${mean}\t${stats}\tOK\t${bam}" > "$out"
EOF

chmod +x cov/rawBamDepth/calc_raw_depth.sh


cd <hpc_scratch>/sperm/germanTrio_data/postNextflow

TARGET_BED="<hpc_scratch>/genomicRegions/germlineV2_hg19.bed"

for sample in $(cut -f1 ../samplesgermanTrio.txt | grep "PD6"); do
  bam="<hpc_references>/nst_links/live/3472/${sample}/${sample}.sample.merged.bam"
  out="cov/rawBamDepth/out/${sample}.meanDepth.tsv"

  bsub -q normal \
    -o cov/rawBamDepth/jobs/${sample}.out \
    -e cov/rawBamDepth/jobs/${sample}.err \
    -R 'select[mem>=100] rusage[mem=100] span[hosts=1]' -M 100 \
    cov/rawBamDepth/calc_raw_depth.sh \
      "$sample" "$bam" "$TARGET_BED" "$out"
done

cd <hpc_scratch>/sperm/germanTrio_data/postNextflow
(
  echo -e "sample\tmeanDepth\tdepthSum\tnBases\tstatus\tbam"
  cat cov/rawBamDepth/out/*.meanDepth.tsv | sort -k1,1
) > cov/rawBamDepth/rawBamDepth_summary.tsv
```

### Investigate Samples
```{r covSummary}
#Coverage summary

covSummaryTarg <- read_tsv(paste0(path, "sperm/germanTrio_data/postNextflow/statsCoverage.tsv"), col_types = cols(), col_names = c("Sample_PD_ID", "bpUnique","bpTotal", "bpMedian", "bpUniquePanel","bpTotalPanel", "bpMedianPanel"))  |>
  left_join(germanTrioSperm, by = "Sample_PD_ID") |> 
  left_join(read_tsv(paste0(metaPath, "germanTrioSpermLong.tsv"), col_types = cols()) |> select(DNA_Amount_UG, Sample_PD_ID) |> 
              bind_rows(read_xls("<local_working_dir>/OxfordTubules/meta/dnaQC.xls") |> mutate(Sample_PD_ID = paste0(SAMPLE, "_tds0001")) |> 
  filter(Sample_PD_ID %in% OxSpermMetaBasic$Sample_PD_ID) |> select(Sample_PD_ID, DNA_Amount_UG = `TOTAL AMOUNT (UG)` )), by = join_by(Sample_PD_ID)) |>
  mutate(percentGenomeCov = bpUnique/3135000000 * 100)|> 
  mutate(avgDuplexCov = bpTotal/bpUnique) |> 
  mutate(percentGenomeCovPanel = bpUniquePanel/3135000000 * 100) |> 
  mutate(avgDuplexCovPanel = bpTotalPanel/bpUniquePanel) |> 
  left_join(maskedCountsPost, by = join_by(Sample_PD_ID, PD_ID, FAM_ID, Age, sampleFilter, Cohort)) |> 
  mutate(maskedProportionPost = replace_na(maskedProportionPost, 0)) |> 
  # Has masked prop > 5
  mutate(sampleFilter = if_else(maskedProportion > 5, "snpContam", sampleFilter)) |> 
  # 024 is clear somatic contam. Has many high VAF variants, outlier burden and both a DMNT3A variant at 10% VAF (!) and a TET2 variant
  # 068 has normal snv burden but outlier indel burden & 3 common blood driver indels (2*TET2,ASXL1)
  # A2 is probable somatic contam. Has outlier burden and 2 duplex calls of V563M in DNMT3A, a known driver: https://eucentral.oncokb.org/gene/DNMT3A/V563M
  # 172 has TET2 & DNMT3A LOFs, 192 has high vaf TET2 lof, 211 has 3 high vaf CHIP drivers, 200 has high vaf SEMG1 + TET2, 227 has several high vaf CHIP
  mutate(sampleFilter = if_else(FAM_ID %in% c("FAM_068", "FAM_A2", "FAM_172", "FAM_192","FAM_211", "FAM_200", "FAM_227"), "somaticContam", sampleFilter)) |> 
  mutate(sampleFilter = if_else(sampleFilter == "FULL" & (bpMedianPanel <= 200 | is.na(bpMedianPanel)), "lowCov", sampleFilter)) |>
  mutate(label = if_else(sampleFilter != "FULL" | burdenPost > 1e-7, FAM_ID, ""))


write_tsv(covSummaryTarg, paste0(localPath, "data/covSummaryTarg.tsv"))

p_covToInput <- ggplot(covSummaryTarg, aes(x = DNA_Amount_UG, y = avgDuplexCovPanel, colour = sampleFilter)) +
  geom_point(alpha = 0.8, size = 3) +
  theme_bw() +
  xlab("Input DNA (ug)") +
  ylab("Average sperm sequenced (dx)") +
  xlim(0,NA) +
  ylim(0,NA)
p_covToInput 

p_burdenPerYear <- ggplot(covSummaryTarg, aes(x = avgDuplexCovPanel, y = burdenPost1percent/Age, colour = sampleFilter, label = label)) +
  geom_point(alpha = 0.8, size = 3) +
  geom_text_repel() +
  theme_bw() +
  xlim(0,NA) +
  ylim(0,NA)
p_burdenPerYear



p_indelsPerYear <- ggplot(covSummaryTarg, aes(x = avgDuplexCovPanel, y = burdenIndelPost1percent/Age, colour = sampleFilter, label = label)) +
  geom_point(alpha = 0.8, size = 3) +
  geom_text_repel() +
  theme_bw() +
  xlim(0,NA) +
  ylim(0,NA)
p_indelsPerYear

yo = covSummaryTarg |> mutate(yo = burdenIndelPost1percent/Age) |> filter(sampleFilter == "FULL") 

p_MaskedburdenPerYear <- ggplot(covSummaryTarg, aes(x = maskedProportion, y = burdenPost1percent/Age, colour = sampleFilter, label = label)) +
  geom_point(alpha = 0.8, size = 3) +
  geom_text_repel() +
  theme_bw() +
  xlim(0,NA) +
  ylim(0,NA)
p_MaskedburdenPerYear

p_covToBurden <- ggplot(covSummaryTarg, aes(x = Age, y = burdenPost1percent, colour = sampleFilter)) +
  geom_point(alpha = 0.8, size = 3) +
  theme_bw() +
  xlim(0,NA) +
  ylim(0,NA)
p_covToBurden 

```

## d) Passed QC
Samples/variants allowed to progress to main analyses
```{r passQC}
targSamplesPassingQC <- covSummaryTarg |> 
  filter(sampleFilter == "FULL")  |> 
  pull(Sample_PD_ID) 

germanTrioSamplesFiltered <- germanTrioSperm |> 
  filter(Sample_PD_ID %in% targSamplesPassingQC) |> 
  select(PD_ID, FAM_ID, Age, Sample_PD_ID)

#Write out filtered metadata locally
write_tsv(germanTrioSamplesFiltered, paste0(metaPath,"germanTrioSamplesFiltered.tsv"))

#Write out filtered sample list on farm
write_tsv(germanTrioSamplesFiltered |> select(Sample_PD_ID, PD_ID), paste0(path,"sperm/germanTrio_data/germanTrioSamplesFiltered.txt"), col_names = F)
write_tsv(germanTrioSamplesFiltered |> 
  mutate(dataset = if_else(str_detect(Sample_PD_ID, "PD6"), "DD.fathers", "Apert.fathers")) |> 
    select(Sample_PD_ID, dataset), paste0(path,"sperm/germanTrio_data/postNextflow/samplesCovFiltered.txt"), col_names = F)

#Write out filtered burdens
germanTrioBurdens <- burdensCorrected |> filter(Sample_PD_ID %in% germanTrioSamplesFiltered$Sample_PD_ID)
write_tsv(germanTrioBurdens, paste0(localPath, "data/germanTrioBurdens.tsv"))

#Write out filtered variant list
write_tsv(filteredVars |> filter(Sample_PD_ID %in% germanTrioSamplesFiltered$Sample_PD_ID), paste0(localPath, "data/filteredVars.tsv"))

write_tsv(allCallsFilter |> select(-sampleFilter) |> left_join(covSummaryTarg |> select(Sample_PD_ID, sampleFilter)), paste0(localPath, "data/allCallsFilterVars.tsv"))

# Write out for sigs with IDs
sigInputTjoint <- read_tsv(paste0(localPath, "data/filteredVars.tsv"), col_types = cols()) |> 
  mutate(blank = ".") |> 
  # ID useful for joining indels back up
  mutate(id = paste(chr,start,ref,alt,Sample_PD_ID, sep = "_")) |> 
  select(chr, start, id, ref, alt, blank2 = blank, blank3 = blank, blank4 = blank) |> 
  arrange(chr, as.numeric(start)) 
write_tsv(sigInputTjoint, paste0(path, "sperm/germanTrio_data/postNextflow/vep/sigInput.vcf"), col_names = F) 
```

#### SupFig 1
```{r supFig1, fig.height=7, fig.width=9}

# Create new labels with counts for the legend
filter_counts_DD <- allCallsFilter |> mutate(FILTER = if_else(FILTER == "diagVar", "PASS", FILTER)) |> 
  filter(Cohort == "DD fathers" & PD_ID %in% germanTrioSamplesFiltered$PD_ID) |> 
  count(FILTER)
vaf_labels_DD <- setNames(paste0(filter_counts_DD$FILTER, " (n=", filter_counts_DD$n, ")"), filter_counts_DD$FILTER)

p_varFiltersDD <- allCallsFilter |> mutate(FILTER = if_else(FILTER == "diagVar", "PASS", FILTER)) |> 
  filter(Cohort == "DD fathers" & PD_ID %in% germanTrioSamplesFiltered$PD_ID) |> 
  mutate(TYPE = if_else(TYPE == "snv", "Substitutions", "Indels")) |> 
  mutate(TYPE = factor(TYPE, levels = c("Substitutions", "Indels"), ordered = TRUE)) |> 
  mutate(BAM_VAF = if_else(BAM_VAF == 0, BAM_VAF_BQ10, BAM_VAF)) |> 
  mutate(BAM_VAF = if_else(BAM_VAF == 0, DUPLEX_VAF, BAM_VAF)) |> 
  mutate(FILTER = factor(FILTER, levels = names(vaf_colours), ordered = TRUE)) |> 
  ggplot(aes(BAM_VAF, after_stat(count), fill = FILTER)) +
  scale_fill_manual(values = vaf_colours, labels = vaf_labels_DD) +
  scale_x_log10(limits = c(6e-5, 1)) +
  geom_density(position = "stack") +
  guides(fill=guide_legend(title="DD father var filters")) +
  ylab("Density") +
  facet_wrap(~TYPE, scales = "free_y", ncol = 2) + 
  theme_pubr() +
  theme(legend.position = "right", strip.background = element_blank(), strip.text = element_text(size = 12))

# Create new labels with counts for the legend
filter_counts_Apert <- allCallsFilter |> filter(Cohort == "Apert fathers" & PD_ID %in% germanTrioSamplesFiltered$PD_ID) |> 
  count(FILTER)
vaf_labels_Apert <- setNames(paste0(filter_counts_Apert$FILTER, " (n=", filter_counts_Apert$n, ")"), filter_counts_Apert$FILTER)

p_varFiltersApert <- allCallsFilter |> filter(Cohort == "Apert fathers" & PD_ID %in% germanTrioSamplesFiltered$PD_ID) |> 
  mutate(TYPE = if_else(TYPE == "snv", "Substitutions", "Indels")) |> 
  mutate(TYPE = factor(TYPE, levels = c("Substitutions", "Indels"), ordered = TRUE)) |> 
  mutate(BAM_VAF = if_else(BAM_VAF == 0, BAM_VAF_BQ10, BAM_VAF)) |> 
  mutate(BAM_VAF = if_else(BAM_VAF == 0, DUPLEX_VAF, BAM_VAF)) |> 
  mutate(FILTER = factor(FILTER, levels = names(vaf_colours), ordered = TRUE)) |> 
  ggplot(aes(BAM_VAF, after_stat(count), fill = FILTER)) +
  scale_fill_manual(values = vaf_colours, labels = vaf_labels_Apert) +
  scale_x_log10(limits = c(6e-5, 1)) +
  geom_density(position = "stack") +
  guides(fill=guide_legend(title="Apert father var filters")) +
  ylab("Density") +
  facet_wrap(~TYPE, scales = "free_y", ncol = 2) + 
  theme_pubr() +
  theme(legend.position = "right", strip.background = element_blank(), strip.text = element_text(size = 12))

# Combine the plots
p_varFiltersDD / p_varFiltersApert + plot_annotation(tag_levels = "A")  &
  theme(plot.tag = element_text(size = 14, face = "bold"))


png(paste0(manuscriptPath, 'Figures/SupFig1_varFilters.png'), height=7, width=9, res = 300, units = "in")
p_varFiltersDD / p_varFiltersApert + plot_annotation(tag_levels = "A")  &
  theme(plot.tag = element_text(size = 14, face = "bold"))
dev.off()

pdf(paste0(manuscriptPath, 'Figures/SupFig1_varFilters.pdf'), height=7, width=9, onefile = F)
p_varFiltersDD / p_varFiltersApert + plot_annotation(tag_levels = "A")  &
  theme(plot.tag = element_text(size = 14, face = "bold"))
dev.off()

```
### Finalize coverage 
### Coverage script
Save as runCovFiltered.sh
```{bash cov, eval=FALSE}
#!/usr/bin/env bash
# runCovFiltered.sh
# Rebuild a filtered total coverage file from samplesCovFiltered.txt and subtract discarded sites.
# Usage:
#   bash runCovFiltered.sh <GROUP_NAME>
# Example:
#   bash runCovFiltered.sh targeted1

set -euo pipefail

if [ "$#" -ne 1 ]; then
  echo "Usage: bash runCovFiltered.sh <GROUP_NAME>"
  exit 1
fi

GROUP="$1"

BASE_DIR="<hpc_scratch>/sperm/germanTrio_data/postNextflow/"
cd "$BASE_DIR"

mkdir -p cov/{jobLogs,final}

SAMPLE_LIST="cov/${GROUP}_samples.filtered.txt"
OUT_BED="cov/final/filtered.${GROUP}.totalCov.bed"
OUT_GZ="cov/final/filtered.${GROUP}.totalCov.bed.gz"

DISCARD="<hpc_scratch>/sperm/germanTrio_data/canAppsOutput/discardedCallsSimple.vcf"

# Build filtered sample list from samplesCovFiltered.txt (col1=sample, col2=group)
awk -F'\t' -v g="$GROUP" '$2==g{print $1}' samplesCovFiltered.txt > "$SAMPLE_LIST"

N_SAMPLES="$(wc -l < "$SAMPLE_LIST" | tr -d ' ')"
echo "Group: $GROUP"
echo "Sample count: $N_SAMPLES"
echo "Sample list: $SAMPLE_LIST"

if [ "$N_SAMPLES" -eq 0 ]; then
  echo "ERROR: No samples found for group '$GROUP' in samplesCovFiltered.txt"
  exit 1
fi

# Always overwrite final outputs
rm -f "$OUT_BED" "$OUT_GZ" "${OUT_GZ}.tbi" 2>/dev/null || true

bsub -J "${GROUP}_finalize_filtered" \
  -o "cov/jobLogs/${GROUP}_finalize_filtered.%J.out" \
  -e "cov/jobLogs/${GROUP}_finalize_filtered.%J.err" \
  -q normal \
  -R 'select[mem>=200] rusage[mem=200] span[hosts=1]' -M 200 \
"set -euo pipefail
 module load bedtools2-2.29.0/ || true
 module load htslib/ || true

 TMP=\$(mktemp -d)

 files=\"\"
 while read -r sample; do
   f=\"cov/targCut/\${sample}.cov.bed\"
   if [ ! -s \"\$f\" ]; then
     echo \"ERROR: missing required per-sample file: \$f\" >&2
     rm -rf \"\$TMP\"
     exit 1
   fi
   files=\"\$files \$f\"
 done < \"${SAMPLE_LIST}\"

 # Combine total coverage across samples
 bedtools unionbedg -i \$files \
   | awk 'BEGIN{OFS=\"\t\"}{s=0; for(i=4;i<=NF;i++) s+=\$i; print \$1,\$2,\$3,s}' \
   > \"\$TMP/combined.bed\"

 # Build discard BED (chr, pos-1, pos) from discardedCallsSimple.vcf (skip header row, ignore 3rd col)
 awk 'BEGIN{OFS=\"\t\"}
      NR==1{next}
      {pos=\$2; print \$1, pos-1, pos}' \"${DISCARD}\" \
   > \"\$TMP/discard.bed\"

 # Remove discarded positions while keeping coverage column
 bedtools intersect -v -a \"\$TMP/combined.bed\" -b \"\$TMP/discard.bed\" > \"${OUT_BED}\"

 bgzip -f -c \"${OUT_BED}\" > \"${OUT_GZ}\"
 tabix -f -p bed \"${OUT_GZ}\"

 rm -rf \"\$TMP\"
"

echo "Submitted finalize-only rebuild for $GROUP (with discarded sites removed)"
echo "Output:"
echo "  $OUT_GZ (+ .tbi)"
```

#### Run Cov
```{bash covRun, eval=FALSE}
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/cov/
bash runCovFiltered.sh DD.fathers
bash runCovFiltered.sh Apert.fathers
```

# 4. Analysis Files
## a) Diagnostic variants
### Force call bams
#### Positions
```{r callPositions}
# Write out snps and indels separately
diagnosticSNVs <- diagnosticVarsPos |> drop_na(start, alt) |> filter(ref != "LargeDeletion") |> 
  filter(ref %in% c("A", "C", "T", "G") & alt %in% c("A", "C", "T", "G")) |> 
  select(chr, start) |> 
  arrange(chr, start) |> 
  distinct()
write_tsv(diagnosticSNVs, paste0(path, "sperm/germanTrio_data/postNextflow/diagnosticVars/diagnosticSites.txt"), col_names = F)

diagnosticIndels <- diagnosticVarsPos |> drop_na(start, alt) |> filter(ref != "LargeDeletion") |> 
  filter(!ref %in% c("A", "C", "T", "G") | !alt %in% c("A", "C", "T", "G")) |> 
  select(chr, start, ref, alt) |> 
  arrange(chr, start) |> 
  distinct()
write_tsv(diagnosticIndels, paste0(path, "sperm/germanTrio_data/postNextflow/diagnosticVars/mpileup/diagnosticIndels.bed"), col_names = F)

```

#### SNV Bam calls
```{bash germline, eval=FALSE}

# Force check diag sites in all indivs at all diag positions
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/diagnosticVars
mkdir -p mpileup/
mkdir -p jobLogs/
rm jobLogs/*
for sample in `cut -f1 ../../samplesgermanTrio.txt`;
  do echo $sample
  bsub -q normal -e jobLogs/germ.$sample.err -o jobLogs/germ.$sample.out -R 'select[mem>=1000] rusage[mem=1000]' -M1000 "module load bcftools-1.9/; bcftools mpileup -f <hpc_references>/pipeline_ref/Human/GRCh37d5/genome.fa -d 15000 -q 30 -a AD,ADF,ADR --ff UNMAP,SECONDARY,QCFAIL,DUP -R diagnosticSites.txt ../neatCrams/$sample.sample.neat.cram > mpileup/$sample.vcf"
done

#Group results for all
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/diagnosticVars
rm diagBamCalls.tsv
for sample in `cut -f1 ../../samplesgermanTrio.txt`;
  do echo $sample
  grep -v "#" mpileup/$sample.vcf | sed "s/$/\t$sample/" >> diagBamCalls.tsv
done

```

#### Indel Calls bam2R
runBam2r_indelCov.R file
```{r collapse, eval=FALSE}
suppressPackageStartupMessages({
  library(deepSNV)
})
suppressPackageStartupMessages({
  library(tidyverse)
})

args = commandArgs(trailingOnly = TRUE)
sample = args[1] #neat NanoSeq BAM

if (length(args) != 1) {
  cat("Wrong number of arguments\n")
  quit(save = "no", status = 1)
}

path = "<hpc_scratch>/sperm/germanTrio_data/"
dedup_bam <- paste0(path, "postNextflow/neatCrams/", sample, ".sample.neat.cram")

i = 1
indels_final <- read_tsv(paste0(path, "postNextflow/diagnosticVars/mpileup/diagnosticIndels.bed"), col_names = c("chr", "pos", "ref", "alt"), col_types = cols()) |> 
  mutate(TYPE = if_else(str_length(ref) > str_length(alt), "del", "ins")) |> 
  # mutate(start = pos - 1) |> 
  # mutate(end = pos + str_length(ref) + 1) |> 
  as.data.frame() |> 
  drop_na()

for (i in c(1:nrow(indels_final))) {
  print(paste("Running indel", i, "\ns"))
    kk = bam2R(dedup_bam, indels_final[i, "chr"], indels_final[i, "pos"], indels_final[i, "pos"], q = 20, mask = 3844, mq = 30)
    total_cov = sum(kk[1, c("A", "C", "G", "T", "a", "c", "g", "t","DEL","INS","del","ins")], na.rm = T)
    if (indels_final[i, "TYPE"] == "del") {
      total_mut = sum(kk[, c("DEL", "del")])
    } else {
      #ins
      total_mut = sum(kk[, c("INS", "ins")])
    }
    indels_final[i, "BAM_MUT"] = total_mut
    indels_final[i, "BAM_COV"] = total_cov
    indels_final[i, "BAM_VAF"] = total_mut / total_cov
}

write_tsv(indels_final, paste0(path, "postNextflow/diagnosticVars/bam2r/", sample, ".bamIndelCov.tsv"))

```

##### Submit bam2R
```{bash bam2r, eval=FALSE}
# Force check diag sites in all indivs at all diag positions
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/diagnosticVars/bam2r
mkdir -p jobLogs/
rm jobLogs/bam2r*

for sample in $(cut -f1 ../../../samplesgermanTrio.txt); do
  echo $sample
  bsub -q normal -e jobLogs/bam2r.$sample.err -o jobLogs/bam2r.$sample.out -R "select[mem>=2000] rusage[mem=2000]" -M2000 \
    "module purge; \
    export REF_CACHE=/lustre/scratch126/core/sciops_repository/cram_cache/%2s/%2s/%s; \
    export REF_PATH=/lustre/scratch125/core/sciops_repository/cram_cache/%2s/%2s/%s:/lustre/scratch126/core/sciops_repository/cram_cache/%2s/%2s/%s:URL=http://refcache.dnapipelines.sanger.ac.uk:8000/%s; \
    module load HGI/softpack/groups/hgi/ubeR; \
    Rscript runBam2r_indelCov.R $sample.sample.neat"
done

#Group results for all
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/diagnosticVars
rm diagBamIndelCalls.tsv
for sample in `cut -f1 ../../samplesgermanTrio.txt`;
  do echo $sample
  grep -v "chr" bam2r/$sample.sample.neat.bamIndelCov.tsv | sed "s/$/\t$sample/" >> diagBamIndelCalls.tsv
done
```

#### Process calls
```{r bamCalls}
bamDiagnosticVars <- read_tsv(paste0(path, "sperm/germanTrio_data/postNextflow/diagnosticVars/diagBamCalls.tsv"), col_types = 'ci-cc----cc', col_names = c("chr", "pos", "ref", "alt", "FORMAT","Sample_PD_ID")) |> 
  distinct() |> 
  separate(FORMAT, into = c("PL", "ADF", "ADR", "AD"), sep = ":") |> 
  select(-c(PL, ADF, ADR)) |> 
  mutate(alt = str_replace(alt, ",<\\*>", "")) |> mutate(alt = str_replace(alt, "<\\*>", ".")) |> 
  separate(AD, into = c("ReadsREF", "ReadsAlt1", "ReadsAlt2", "ReadsAlt3"), sep = ",", fill = "right") |> 
  mutate(across(starts_with("Reads"), ~as.numeric(replace_na(., "0")))) |> 
  mutate(altReads = ReadsAlt1 + ReadsAlt2 + ReadsAlt3) |> 
  mutate(BAM_COV = ReadsREF + altReads) |> 
  # Split by allele
  mutate(alt = strsplit(as.character(alt), ",")) |> 
  unnest(alt) |>
  group_by(chr, pos, ref, Sample_PD_ID, BAM_COV) |>
  mutate(
    altReads = case_when(
      row_number() == 1 ~ ReadsAlt1,
      row_number() == 2 ~ ReadsAlt2,
      row_number() == 3 ~ ReadsAlt3,
      TRUE ~ NA_real_
    )) |>
  ungroup() |> 
  select(-c(ReadsAlt1,ReadsAlt2,ReadsAlt3, ReadsREF)) |> 
  # Calculate VAF of correct base call
  left_join(diagnosticVarsPos |> select(chr, pos = start, altVar = alt) |> distinct(), by = join_by(chr, pos)) |> 
  mutate(BAMcalls = if_else(altVar == alt, altReads, 0)) |> 
  mutate(BAM_VAF = BAMcalls/BAM_COV) |> 
  group_by(chr,pos,ref,altVar,Sample_PD_ID) |> summarize(BAMcalls = max(BAMcalls), BAM_COV = max(BAM_COV), BAM_VAF = max(BAM_VAF), .groups = "drop") |> 
  rename(alt = altVar) |> 
  filter(str_length(ref) == 1 & str_length(alt) == 1) |> 
  # Add indel variants
  bind_rows(read_tsv(paste0(path, "sperm/germanTrio_data/postNextflow/diagnosticVars/diagBamIndelCalls.tsv"), col_types = 'cicc-ii-c', col_names = c("chr", "pos", "ref", "alt", "BAMcalls", "BAM_COV", "Sample_PD_ID")) |> 
              mutate(BAM_VAF = BAMcalls/BAM_COV)) |> 
  rename(start = pos) |> 
  left_join(diagnosticVarsPos |> select(chr,start,ref,alt,VAR_ID) |> distinct(), by = join_by(chr, start, ref, alt)) |> 
  drop_na(VAR_ID) |> distinct()


cohortSumm <- bamDiagnosticVars |> 
  left_join(diagnosticVarsPos |> left_join(germanTrioSperm |> select(Sample_PD_ID, FAM_ID), by = join_by(FAM_ID)) |> 
              select(Sample_PD_ID, VAR_matched = VAR_ID) |> 
              distinct(), 
            by = join_by(Sample_PD_ID), relationship = "many-to-many") |> 
  mutate(group = if_else(VAR_matched == VAR_ID, "Target", "Cohort")) |> 
  select(-VAR_matched) |> 
  filter(group == "Cohort") |> 
  group_by(chr, start, ref, alt, VAR_ID, group) |> 
  summarize(BAMcalls = sum(BAMcalls), BAM_COV = sum(BAM_COV), .groups = "drop")

indivBAMCov <- bamDiagnosticVars |> 
  left_join(diagnosticVarsPos |> left_join(germanTrioSperm |> select(Sample_PD_ID, FAM_ID), by = join_by(FAM_ID)) |> 
              select(Sample_PD_ID, VAR_matched = VAR_ID) |> 
              distinct(), 
            by = join_by(Sample_PD_ID), relationship = "many-to-many") |> 
  mutate(group = if_else(VAR_matched == VAR_ID, "Target", "Cohort")) |> 
  select(-VAR_matched) |> 
  filter(group == "Target") |> 
  bind_rows(cohortSumm) |> 
  mutate(BAM_VAF = BAMcalls / BAM_COV)

```

### Duplex
#### Diagnostic Vars
```{r collapse}
diagnosticVarsPosExpandBed <- read_tsv(paste0(metaPath, "diagnosticVarsPos.txt"), col_types = cols()) |> 
  mutate(startBed = start - 1) |> 
  mutate(endBed = start + str_length(ref) - 1) |> 
  select(chr, startBed, endBed) |> 
  arrange(chr, startBed) |> drop_na()
write_tsv(diagnosticVarsPosExpandBed, paste0(path, "sperm/germanTrio_data/postNextflow/diagnosticVars/duplexCov/diagnosticVarsPosExpand.bed"), col_names = F)

```

##### Cov bed intersect
```{bash collapse, eval=FALSE}
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/diagnosticVars/
mkdir -p duplexCov/
mkdir -p duplexCov/jobLogs/

cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/diagnosticVars/duplexCov/
sort -k1,1 -k2,2n diagnosticVarsPosExpand.bed > diagnosticVarsPosExpand.sorted.bed
rm jobLogs/*
for sample in `cut -f1 ../../../samplesgermanTrio.txt`;
  do echo $sample
  bsub -o jobLogs/intersect.$sample.out -e jobLogs/intersect.$sample.err -q normal -R 'select[mem>=1000] rusage[mem=1000] span[hosts=1]' -M 1000 "module load bedtools2-2.29.0/; bedtools intersect -a ../../cov/source/$sample.cov.bed -b diagnosticVarsPosExpand.sorted.bed -wa > $sample.diag.cov.bed"
done

# Group results
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/diagnosticVars/duplexCov/
rm allDiagCov.bed
for sample in `cut -f1 ../../../samplesgermanTrio.txt`;
  do echo $sample
  sed "s/$/\t$sample/" $sample.diag.cov.bed >> allDiagCov.bed
done

```

##### Process duplex Cov
```{r collapse}
# Process cov
diagnosticVarsCov <- read_tsv(paste0(metaPath, "diagnosticVarsPos.txt"), col_types = cols()) |> 
  drop_na(start, ref) |> 
  mutate(length = str_length(ref))|> 
  rowwise() |> 
  do({
    copy_count <- seq_len(.$length)
    data.frame(
      chr = rep(.$chr, .$length),
      start = rep(.$start, .$length),
      length = rep(.$length, .$length),
      VAR_ID = rep(.$VAR_ID, .$length), 
      pos = .$start + copy_count - 1
    )
  }) |> 
  ungroup() |> distinct() |> 
  # Add coverage
  left_join(read_tsv(paste0(path, "sperm/germanTrio_data/postNextflow/diagnosticVars/duplexCov/allDiagCov.bed"), col_types = 'c-dcc', col_names = c("chr", "pos", "DUPLEX_COV", "Sample_PD_ID")) |> distinct() |> 
  mutate(DUPLEX_COV = as.numeric(str_sub(DUPLEX_COV, 7, -1))), relationship = "many-to-many", by = join_by(pos, chr)) |> 
  mutate(DUPLEX_COV = replace_na(DUPLEX_COV, 0)) |> 
  group_by(chr, start,length, VAR_ID, Sample_PD_ID) |> summarize(DUPLEX_COV = mean(DUPLEX_COV), .groups = "drop") |> 
  filter(VAR_ID %in% (diagnosticVarsPos |> filter(FAM_ID %in% germanTrioSamplesFiltered$FAM_ID) |> pull(VAR_ID))) |> 
  select(-length) |> 
  left_join(germanTrioSperm |> select(Sample_PD_ID, FAM_ID), by = join_by(Sample_PD_ID))

cohortVarCov <- diagnosticVarsCov |> 
  group_by(chr, start, VAR_ID) |> summarize(totalCov = sum(DUPLEX_COV), meanCov = mean(DUPLEX_COV), count = n(), .groups = "drop") |> 
  mutate(count = if_else(totalCov == 0, 0, count)) 

# Anchor on diagnosticVarsPos (one row per FAM_ID x VAR_ID, always complete) rather
# than on diagnosticVarsCov. A variant with 0 reads across every sample has no row
# at all in allDiagCov.bed, so diagnosticVarsCov loses the Sample_PD_ID/FAM_ID for
# it (collapses to a single NA-keyed row) - starting the join from diagnosticVarsCov
# then silently dropped that variant/family entirely instead of showing 0 coverage.
# A family with no sperm sample at all (Sample_PD_ID unknown, i.e. patSpermSequenced
# == "No") is kept as NA throughout - only "sequenced but 0 reads" becomes 0. VAF is
# NA whenever coverage is 0 (or unsequenced), since it's undefined, not 0.
indivVarCov <- diagnosticVarsPos |>
  select(FAM_ID, VAR_ID, ref, alt, chr, start) |>
  distinct() |>
  left_join(germanTrioSperm |> select(Sample_PD_ID, FAM_ID), by = join_by(FAM_ID), relationship = "many-to-many") |>
  left_join(diagnosticVarsCov |> select(chr, start, VAR_ID, Sample_PD_ID, manualCov = DUPLEX_COV), by = join_by(chr, start, VAR_ID, Sample_PD_ID)) |>
  mutate(manualCov = if_else(is.na(Sample_PD_ID), NA_real_, replace_na(manualCov, 0))) |>
  left_join(read_tsv(paste0(localPath, "data/germanTrioAllTargVars.tsv"), col_types = cols()) |> select(chr, start, ref, alt, FAM_ID, TIMES_CALLED, DUPLEX_VAF), by = join_by(chr, start, ref, alt, FAM_ID)) |>
  mutate(TIMES_CALLED = if_else(is.na(Sample_PD_ID), NA_real_, replace_na(TIMES_CALLED, 0))) |>
  mutate(DUPLEX_VAF = replace_na(DUPLEX_VAF, 0)) |>
  mutate(DUPLEX_VAF = if_else(is.na(manualCov) | manualCov == 0, NA_real_, DUPLEX_VAF)) |>
  full_join(cohortVarCov, by = join_by(chr, start, VAR_ID)) |> drop_na(FAM_ID) |>
  mutate(totalCov = replace_na(totalCov, 0)) |>
  left_join(read_tsv(paste0(localPath, "data/germanTrioAllTargVars.tsv"), col_types = cols()) |> group_by(chr, start, ref, alt) |> summarize(cohortCalls = sum(TIMES_CALLED), .groups = "drop"), by = join_by(chr, start, ref, alt)) |>
  mutate(cohortCalls = replace_na(cohortCalls, 0)) |>
  mutate(nonTargetCalls = cohortCalls - coalesce(TIMES_CALLED, 0)) |>
  mutate(nonTargetVAF = if_else(totalCov == 0, NA_real_, nonTargetCalls/totalCov))

diagnosticVarsCalls <- indivVarCov |> select(VAR_ID, FAM_ID, chr,start,ref,alt, duplexCalls = nonTargetCalls, duplexCov = totalCov, duplexVAF = nonTargetVAF) |> mutate(group = "Cohort", FAM_ID = "Cohort") |>
  bind_rows(indivVarCov |> select(VAR_ID, FAM_ID, chr,start,ref,alt, Sample_PD_ID, duplexCalls = TIMES_CALLED, duplexCov = manualCov, duplexVAF = DUPLEX_VAF) |> mutate(group = "Target")) |>
  left_join(indivBAMCov |> left_join(germanTrioSperm |> select(Sample_PD_ID, FAM_ID), by = join_by(Sample_PD_ID)) |> select(-Sample_PD_ID), by = join_by(VAR_ID, FAM_ID, chr, start, ref, alt, group)) |>
  # Same 0-vs-NA rule as duplex above, but tracked independently: BAM and duplex
  # coverage come from separate pipelines, so a variant can have 0 duplex reads yet
  # real BAM coverage (or vice versa) - never infer one from the other.
  mutate(BAMcalls = if_else(group == "Target" & is.na(Sample_PD_ID), NA_real_, replace_na(BAMcalls, 0))) |>
  mutate(BAM_COV = if_else(group == "Target" & is.na(Sample_PD_ID), NA_real_, replace_na(BAM_COV, 0))) |>
  mutate(BAM_VAF = if_else(is.na(BAM_COV) | BAM_COV == 0, NA_real_, BAMcalls / BAM_COV)) |>
  left_join(diagnosticVarsPos |> select(VAR_ID, gene) |> distinct()) |>
  arrange(VAR_ID) |> distinct() |>
  mutate(varLabel = gene) |>
  mutate(varLabel = if_else(coalesce(duplexCalls, 0) > 0, varLabel, "")) |>
  select(-Sample_PD_ID) |>
  mutate(Cohort = if_else(str_detect(FAM_ID, "FAM_A"),  "Apert fathers", "DD fathers"))

write_tsv(diagnosticVarsCalls, paste0(localPath, "data/diagnosticVarsCalls.tsv"))
```


#### Other vars
```{r collapse}

trioDNM_Vars <- read_tsv(paste0(metaPath, "trioDNM_Vars.tsv"), col_types = cols())

famsIncluded <- germanTrioMeta |> filter(PD_ID %in% trioDNM_Vars$PD_ID)
spermIndivs <- germanTrioSamplesFiltered |> filter(FAM_ID %in% famsIncluded$FAM_ID)

trioPosExpandBed <- trioDNM_Vars |> 
  left_join(germanTrioSamplesFiltered |> select(Sample_PD_ID, FAM_ID), by = join_by(FAM_ID)) |> 
  drop_na(Sample_PD_ID) |> # Only take variants in retained sperm samples
  mutate(startBed = start - 1) |> 
  mutate(endBed = start + str_length(ref) - 1) |> 
  select(chr, startBed, endBed, Sample_PD_ID) |> 
  arrange(chr, startBed) |> drop_na()
write_tsv(trioPosExpandBed, paste0(path, "sperm/germanTrio_data/postNextflow/diagnosticVars/duplexCovAll/trioPosExpandBed.bed"), col_names = F)

```

##### Cov bed intersect
```{bash collapse, eval=FALSE}
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/diagnosticVars/
mkdir -p duplexCovAll/
mkdir -p duplexCovAll/jobLogs/

cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/diagnosticVars/duplexCovAll/
sort -k1,1 -k2,2n trioPosExpandBed.bed > trioPosExpandBed.sorted.bed
rm jobLogs/*
for sample in `cut -f1 ../../../germanTrioSamplesFiltered.txt`;
  do echo $sample
  bsub -o jobLogs/intersect.$sample.out -e jobLogs/intersect.$sample.err -q normal -R 'select[mem>=100] rusage[mem=100] span[hosts=1]' -M 100 "module load bedtools2-2.29.0/; grep '$sample' trioPosExpandBed.sorted.bed \
  | bedtools intersect -a ../../cov/source/$sample.cov.bed -b - -wa \
  > $sample.dnms.cov.bed"
done

# Group results
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/diagnosticVars/duplexCovAll/
rm allDNMsCov.bed
for sample in `cut -f1 ../../../germanTrioSamplesFiltered.txt`;
  do echo $sample
  sed "s/$/\t$sample/" $sample.dnms.cov.bed >> allDNMsCov.bed
done

```

##### Process duplex Cov
```{r collapse}
# Process cov
allDNMsCov0 <- trioDNM_Vars |> 
  left_join(germanTrioSamplesFiltered |> select(Sample_PD_ID, FAM_ID), by = join_by(FAM_ID)) |> 
  drop_na(Sample_PD_ID) |> # Only take variants in retained sperm samples
  mutate(length = str_length(ref)) |> 
  rowwise() |> 
  do({
    copy_count <- seq_len(.$length)
    data.frame(
      chr = rep(.$chr, .$length),
      start = rep(.$start, .$length),
      length = rep(.$length, .$length),
      PD_ID = rep(.$PD_ID, .$length), 
      Sample_PD_ID = rep(.$Sample_PD_ID, .$length), 
      pos = .$start + copy_count - 1
    )
  }) |> 
  ungroup() |> distinct()|> 
  # Add coverage
  left_join(read_tsv(paste0(path, "sperm/germanTrio_data/postNextflow/diagnosticVars/duplexCovAll/allDNMsCov.bed"), col_types = 'c-dcc', col_names = c("chr", "pos", "DUPLEX_COV", "Sample_PD_ID")) |> distinct() |> 
  mutate(DUPLEX_COV = as.numeric(str_sub(DUPLEX_COV, 7, -1))), by = join_by(pos, chr, Sample_PD_ID)) |> 
  mutate(DUPLEX_COV = replace_na(DUPLEX_COV, 0)) |> 
  group_by(chr, start,length, Sample_PD_ID, PD_ID) |> summarize(DUPLEX_COV = mean(DUPLEX_COV), .groups = "drop") |> 
  select(-length) |> left_join(diagnosticVars |> select(chr, start, FAM_ID, VAR_ID) |> mutate(diagVar = T)) |> 
  mutate(diagVar = replace_na(diagVar, F)) |> 
  filter(!diagVar)
  # Check if actually a diagnostic Var

table(allDNMsCov0 |> mutate(anyCov = if_else(DUPLEX_COV>0, T, F)) |> pull(anyCov))

allDNMsCov <- allDNMsCov0 |> filter(DUPLEX_COV>0) |> 
  left_join(germanTrioTargVars |> select(chr, start, Sample_PD_ID, TIMES_CALLED, DUPLEX_COV))

write_tsv(allDNMsCov, paste0(localPath, "data/allDNMsCov.tsv"))
```

## b) VEP 
### Setup files
#### ClinVar + CADD
```{bash CADD, eval=FALSE}
cd <hpc_shared>/External_databases/Variants/ClinVar/GRCh37/
wget https://ftp.ncbi.nlm.nih.gov/pub/clinvar/vcf_GRCh37/clinvar_20250729.vcf.gz
wget https://ftp.ncbi.nlm.nih.gov/pub/clinvar/vcf_GRCh37/clinvar_20250729.vcf.gz.tbi
wget https://kircherlab.bihealth.org/download/CADD/v1.7/GRCh37/whole_genome_SNVs.tsv.gz.tbi
echo "wget https://kircherlab.bihealth.org/download/CADD/v1.7/GRCh37/whole_genome_SNVs.tsv.gz" | bsub -o cadd.out -e cadd.err -q normal -R 'select[mem>=100] rusage[mem=100] span[hosts=1]' -M100 


# Converting cadd to vcf for custom annotation bc I can't get plugin to work
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/vep
nano caddToVCF.sh
#!/bin/bash

# Paths
SNV_FILE="<hpc_shared>/External_databases/Variants/CADD/GRCh37/whole_genome_SNVs.tsv.gz"
COVERED_BED="<hpc_scratch>/genomicRegions/germlineV1+V2_merged_hg19.bed"

module load bedtools2-2.29.0/

# Process: Convert to BED format, intersect with covered regions, convert to VCF, compress
gunzip -c "$SNV_FILE" \
  | tail -n +3 \
  | awk -F"\t" 'BEGIN {OFS="\t"} {print $1, $2 - 1, $2, $1, $2, ".", $3, $4, ".", ".", "CADD="$6}' \
  | bedtools intersect -a stdin -b "$COVERED_BED" -wa \
  | cut -f4- \
  | bgzip > cadd0.vcf.gz
tabix -p vcf cadd0.vcf.gz

chmod +x caddToVCF.sh
bsub -o jobLogs/cadd.out -e jobLogs/cadd.err -q long -R 'select[mem>=1000] rusage[mem=1000] span[hosts=1]' -M1000 ./caddToVCF.sh

```

#### Methylation 
```{bash methProcess, eval=FALSE}
cd <hpc_scratch>/genomicRegions/methylation

nano cov3Convert.sh
#!/bin/bash
#Select min coverage 3, take average methylation across samples, replace coverage count with mean methylation %
module load bedtools2-2.29.0/

cat ENCFF715DMX.bed ENCFF638QVP.bed | awk '$10>2' | sort -k1,1 -k2,2n | bedtools merge -d -1 -c 4,11,6 -o distinct,mean,distinct > liftover/testis_cov3_38.bed

#LiftOver to hg19
module load liftOverReport/2.2.1

liftOver liftover/testis_cov3_38.bed liftover/hg38ToHg19.over.chain.gz testis_cov3_19.bed liftover/testis_cov3_38_unmapped.bed

# Remove "chr" prefix from chromosome field
sed 's/^chr//' testis_cov3_19.bed > testis_cov3_19noCHR.bed

# Extract columns 1-3 and 5, sort, bgzip, and index
cut -f1-3,5 testis_cov3_19noCHR.bed \
  | sort -k1,1 -k2,2n \
  | bedtools merge -d -1 -c 4 -o mean \
  | bgzip -c > testisMeth19.bed.gz

tabix -p bed testisMeth19.bed.gz

gunzip testisMeth19.bed.gz > testisMeth19.germanTrioCov.bed

#Farm job submissions
bsub -o out.%J -e err.%J -q normal -n 1 -R 'select[mem>=1000] rusage[mem=1000]' -M1000 ./cov3Convert.sh

```

#### TriNuc Annotation
```{bash trinucCov, eval=FALSE}
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/cov
nano triNucBed.sh

#!/bin/bash
set -euo pipefail

cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/cov
SAMPLE_LIST=../../germanTrioSamplesFiltered.txt

rm -f triNucLong.txt
cut -f1 "$SAMPLE_LIST" | while read sample; do
    cut -f1 -d ";" source/${sample}.cov.bed | cut -f1,3,4 >> triNucLong.txt
done

sort -T tmp -k1,1 -k2,2n triNucLong.txt | uniq > triNucUniq.txt
awk 'BEGIN {FS=OFS="\t"} {print $1, $2-1, $2, $3}' triNucUniq.txt | bgzip -c > triNucUniq.bed.gz
tabix -p bed triNucUniq.bed.gz

cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/cov
bsub -o jobLogs/triNuc.%J.out -e jobLogs/triNuc.%J.err -q normal -R 'select[mem>=10000] rusage[mem=10000] span[hosts=1]' -M 10000 ./triNucBed.sh
```

#### Trio DNMs
```{bash trioDNMs, eval=FALSE}

for f in \
  <hpc_scratch>/GermanTrio_analyses_Project3500/Caveman/trio_cgpvaf_snp/F*/*_denovo_mutations_appended.vcf \
  <hpc_scratch>/GermanTrio_analyses_Project3500/Caveman/trio_cgpvaf_indels/F*/*_denovo_indel_mutations_appended.vcf
do
  id=$(basename "$f" | sed -E 's/_denovo(_indel)?_mutations_appended\.vcf$//')
  grep -v '^#' "$f" | awk -v id="$id" '{print $0 "\t" id}'
done > <hpc_scratch>/sperm/germanTrio_data/dnmData/trioDNMs.tsv

```

#### VAFs
```{bash trioDNMs, eval=FALSE}

cd <hpc_scratch>/sperm/germanTrio_data/dnmData

# Merge SNP VAFs
for f in <hpc_scratch>/GermanTrio_analyses_Project3500/Caveman/trio_cgpvaf_snp/F*/*_denovo_mutations.txt; do
    awk 'BEGIN{OFS="\t"} NR==1{split($27,a,"_"); id=a[1]} NR>1{print $1,$2,$3,$4,id,$27}' "$f"
done > merged_denovo_mutationsVAFs.txt

# Merge indel VAFs and append
for f in <hpc_scratch>/GermanTrio_analyses_Project3500/Caveman/trio_cgpvaf_indels/F*/*_denovo_indel_mutations.txt; do
    awk 'BEGIN{OFS="\t"} NR==1{split($21,a,"_"); id=a[1]} NR>1{print $1,$2,$3,$4,id,$21}' "$f"
done >> merged_denovo_mutationsVAFs.txt

# Add header
sed -i '1i chr\tstart\tref\talt\tPD_ID\tVAF' merged_denovo_mutationsVAFs.txt

```

#### Mosaic data sets
```{r mosaicSources}

# Convert evrony set to 37
evronySet <- read_excel(paste0(localPath, "externalData/mosaicVarsEvrony2025.xlsx")) |> 
    filter(`Sample type` == "Sperm" ) |>
      mutate(ID = paste0(ID, Timepoint)) |> 
    # Filter to callable range 1-30%
    filter(VAF >= 0.01 & VAF < 0.3) |> 
    mutate(set = "E25") |> 
    select(CHROM:ALT, ID, VAF, set) |> 
  mutate(varID = paste(CHROM,POS, sep = "_"))
write_tsv(evronySet |> mutate(bedStart = POS -1) |> select(CHROM, bedStart, POS, varID), paste0(localPath, "externalData/evronyVars38.bed"), col_names = F)

#Read in liftover output
evronySet37 <- evronySet |> select(-c(CHROM, POS)) |> 
  left_join(read_tsv(paste0(localPath, "externalData/evronyVars37.bed"), col_names = c('CHROM', 'POS', 'varID'),  col_types = 'c-ic') |> 
  mutate(CHROM = str_remove(CHROM, "chr")) |> distinct(), by = join_by(varID)) |> 
  select(-varID)

mosaicVars <- read_excel(paste0(localPath, "externalData/mosaicVarsGleeson2021.xlsx"), col_types = c("guess", "text", rep("guess", 60))) |> 
  filter(SET_SPERM_MOSAIC == 1) |> 
  # Filter to callable range 1-30%
  filter(MAF_SPERM_A >= 0.01 & MAF_SPERM_A < 0.3) |> 
  # Filter to cov ~300x samples
  filter(str_detect(INDIVIDUAL, "ID")) |> 
  mutate(set = "G21") |> 
  select(CHROM:ALT, ID = INDIVIDUAL, VAF = MAF_SPERM_A, set) |> 
  bind_rows(read_excel(paste0(localPath, "externalData/mosaicVarsGleeson2022.xlsx"), col_types = c("guess","guess", "text", rep("guess", 33))) |> 
    filter(str_detect(ID, "Sperm") & SD_SPERM_AF > 0) |> 
    # Filter to callable range 1-30%
    filter(SD_SPERM_AF >= 0.01 & SD_SPERM_AF < 0.3) |> 
    mutate(set = "G22") |> 
    select(CHROM:ALT, ID = FAMILY_ID, VAF = SD_SPERM_AF, set))|> 
  bind_rows(evronySet37) |> 
  rename(chr = CHROM, start = POS, ref = REF, alt = ALT)

write_tsv(mosaicVars, paste0(localPath, "externalData/mosaicVars0.tsv"))
```

### Run Variants
#### Format calls for VEP
```{r formatCalls}
# Variant calls
vepInputTjoint <- read_tsv(paste0(localPath, "data/allCallsFilterVars.tsv"), col_types = cols()) |> 
  mutate(blank = ".") |> 
  # ID useful for joining indels back up
  mutate(id = paste(chr,start,ref,alt, sep = "_")) |> 
  select(chr, start, id, ref, alt, blank2 = blank, blank3 = blank, blank4 = blank) |> 
  arrange(chr, as.numeric(start)) |> 
  distinct()
write_tsv(vepInputTjoint, paste0(path, "sperm/germanTrio_data/postNextflow/vep/allCallsFilterVars.vcf"), col_names = F) 

# Discarded calls - Commented to prevented accidental run
# discardedVepInput <- anyCounts <- read_tsv(paste0(path, "sperm/germanTrio_data/canAppsOutput/discardedCalls.vcf"), col_names = c("chr", "start", "ref", "alt", "FILTER","info", "Sample_PD_ID"), col_types = 'cd-cc-ccc') |> 
#   mutate(blank = ".") |> 
#   # ID useful for joining indels back up
#   mutate(id = paste(chr,start,ref,alt, sep = "_")) |> 
#   select(chr, start, id, ref, alt, blank2 = blank, blank3 = blank, blank4 = blank) |> 
#   arrange(chr, as.numeric(start)) |> 
#   distinct()
# write_tsv(discardedVepInput, paste0(path, "sperm/germanTrio_data/postNextflow/vep/discardedCallsVars.vcf"), col_names = F) 

# Diagnostic variants
vepInputdiagnosticVars <- read_tsv(paste0(metaPath, "diagnosticVarsPos.txt"), col_types = cols()) |> 
  mutate(blank = ".") |> 
  # ID useful for joining indels back up
  mutate(id = paste(chr,start,ref,alt, sep = "_")) |> 
  select(chr, start, id, ref, alt, blank2 = blank, blank3 = blank, blank4 = blank) |> 
  arrange(chr, as.numeric(start)) |> 
  distinct()
write_tsv(vepInputdiagnosticVars, paste0(path, "sperm/germanTrio_data/postNextflow/vep/diagnosticVars.vcf"), col_names = F) 

# Trio DNMs
vepInputTrioDNMs <- read_tsv(paste0(path, "sperm/germanTrio_data/dnmData/trioDNMs.tsv"), col_types = cols(), col_names = c("chr","start","blank","ref","alt", "PD_ID")) |> 
  filter(PD_ID != "PD67073b") |> # Accidentally sent sister's blood and not index
  filter(PD_ID != "PD64157a") |> # Removed for QC
  # ID useful for joining indels back up
  mutate(id = paste(chr,start,ref,alt, sep = "_")) |> 
  select(chr, start, id, ref, alt, blank2 = blank, blank3 = blank, blank4 = blank) |> 
  arrange(chr, as.numeric(start)) |> 
  distinct()
write_tsv(vepInputTrioDNMs, paste0(path, "sperm/germanTrio_data/postNextflow/vep/trioDNM_Vars.vcf"), col_names = F) 

# MosaicVars
vepInputMosaicVars <- read_tsv(paste0(localPath, "externalData/mosaicVars0.tsv"), col_types = cols()) |> 
  mutate(blank = ".") |> 
  # ID useful for joining indels back up
  mutate(id = paste(chr,start,ref,alt, sep = "_")) |> 
  select(chr, start, id, ref, alt, blank2 = blank, blank3 = blank, blank4 = blank) |> 
  arrange(chr, as.numeric(start)) |> 
  distinct()
write_tsv(vepInputMosaicVars, paste0(path, "sperm/germanTrio_data/postNextflow/vep/MosaicVars_Vars.vcf"), col_names = F) 

```

#### Script
```{bash maf, eval=FALSE}

cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/vep
module load conda/24.9.2
conda activate <hpc_software>/conda_envs/envs/py3.6
OUTPATH="<hpc_scratch>/sperm/germanTrio_data/postNextflow/vep"
export PATH="<hpc_software>/ensembl-vep//htslib:/software/jdk-11.0.2/bin/:$PATH"
export PERL5LIB="<hpc_software>/ensembl-vep/:<hpc_home>/perl5/lib/perl5:<hpc_home>/perl5/lib/perl5/site_perl:<hpc_software>/000_SOFTWARE/lib/perl5:$PERL5LIB"

rm $OUTPATH/out*_vep.*
ls *Vars.vcf | while read SAMPLE ; do echo "<hpc_software>/ensembl-vep/vep -i $SAMPLE -o $OUTPATH/out_$SAMPLE.tsv --tab \
--dir <hpc_software>/000_SOFTWARE/vep/ --species homo_sapiens --cache --offline --force --refseq --no_stats --fork 4 --assembly GRCh37 \
--fasta <hpc_software>/000_SOFTWARE/vep/homo_sapiens/Homo_sapiens.GRCh37.75.dna.primary_assembly.fa \
--symbol --af_gnomad --pick --show_ref_allele \
--custom <hpc_shared>/External_databases/Variants/ClinVar/GRCh37/clinvar_20250729.vcf.gz,ClinVar,vcf,exact,0,CLNDN,CLNSIG,CLNSIGCONF \
--custom <hpc_scratch>/genomicRegions/methylation/testisMeth19.bed.gz,methylation,bed,exact,0 \
--custom ../cov1/triNucUniq.bed.gz,trinuc,bed,exact,0 \
--custom <hpc_scratch>/sperm/germanTrio_data/postNextflow/vep/cadd0.vcf.gz,CADD,vcf,exact,0,CADD" | bsub -e <hpc_scratch>/sperm/germanTrio_data/postNextflow/vep/out.${SAMPLE%.vcf}_vep.err -o <hpc_scratch>/sperm/germanTrio_data/postNextflow/vep/out.${SAMPLE%.vcf}_vep.out -J ${SAMPLE%.vcf}_vep -n4 -q normal -R 'select[mem>=2000] rusage[mem=2000] span[hosts=1]' -M2000  -env "all" ; done 

```

#### Script trios for Joe
```{bash maf, eval=FALSE}

cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/vep
module load conda/24.9.2
conda activate <hpc_software>/conda_envs/envs/py3.6
OUTPATH="<hpc_scratch>/sperm/germanTrio_data/postNextflow/vep"
export PATH="<hpc_software>/ensembl-vep//htslib:/software/jdk-11.0.2/bin/:$PATH"
export PERL5LIB="<hpc_software>/ensembl-vep/:<hpc_home>/perl5/lib/perl5:<hpc_home>/perl5/lib/perl5/site_perl:<hpc_software>/000_SOFTWARE/lib/perl5:$PERL5LIB"

rm $OUTPATH/out*_vep.clinAnnot.*
ls trioDNM_Vars.vcf | while read SAMPLE ; do echo "<hpc_software>/ensembl-vep/vep \
  -i $SAMPLE \
  -o $OUTPATH/out_$SAMPLE.clinAnnot.tsv \
  --tab \
  --dir <hpc_software>/000_SOFTWARE/vep/ \
  --species homo_sapiens \
  --cache --offline --force --refseq --no_stats --fork 4 \
  --assembly GRCh37 \
  --fasta <hpc_software>/000_SOFTWARE/vep/homo_sapiens/Homo_sapiens.GRCh37.75.dna.primary_assembly.fa \
  --symbol --af_gnomad --pick --show_ref_allele \
  --check_existing \
  --sift b --polyphen b \
  --hgvs \
  --plugin Phenotypes \
  --fields 'Uploaded_variation,Location,Allele,REF_ALLELE,Gene,SYMBOL,Feature,Feature_type,Consequence,IMPACT,EXON,INTRON,\
HGVSc,HGVSp,SIFT,PolyPhen,Existing_variation,VARIANT_CLASS,gnomADe_AF,gnomADg_AF,\
ClinVar_CLNDN,ClinVar_CLNSIG,ClinVar_CLNSIGCONF,\
methylation,trinuc,CADD_CADD,Phenotypes' \
  --custom <hpc_shared>/External_databases/Variants/ClinVar/GRCh37/clinvar_20250729.vcf.gz,ClinVar,vcf,exact,0,CLNDN,CLNSIG,CLNSIGCONF \
  --custom <hpc_scratch>/genomicRegions/methylation/testisMeth19.bed.gz,methylation,bed,exact,0 \
  --custom ../cov1/triNucUniq.bed.gz,trinuc,bed,exact,0 \
  --custom <hpc_scratch>/sperm/germanTrio_data/postNextflow/vep/cadd0.vcf.gz,CADD,vcf,exact,0,CADD" | bsub -e <hpc_scratch>/sperm/germanTrio_data/postNextflow/vep/out.${SAMPLE%.vcf}_vep.clinAnnot.err -o <hpc_scratch>/sperm/germanTrio_data/postNextflow/vep/out.${SAMPLE%.vcf}_vep.clinAnnot.out -J ${SAMPLE%.vcf}_vep -n4 -q normal -R 'select[mem>=2000] rusage[mem=2000] span[hosts=1]' -M2000  -env "all" ; done 

```

#### Annot objects
```{r annotObjects}
# DDD monoallelic lof genes from 2025
dddLOF <- read_csv("<local_working_dir>/G2P_DD_2025-09-17.csv", col_types = cols()) |> 
  dplyr::rename(gene = `gene symbol`)|> 
  filter(`allelic requirement` != "biallelic_autosomal" & `confidence` %in% c("strong","definitive")) |> 
  filter(`molecular mechanism` == "loss of function") |> distinct(gene) |> pull(gene)
write_tsv(tibble(gene = dddLOF), paste0(localPath, "data/dddLOF.txt"))
# Likely drivers is combo of TwinsUK genes & hotspots + those identified in this project
lofGenesGermanTrio <- read_tsv(paste0(localPath, "data/sigGenes.tsv"), col_types = cols()) |> 
    mutate(germlineMechanism = if_else((ptrunc_cv < 0.1 | pind_cv < 0.1) & ((n_non + n_spl + n_ind) > 1) | gene_name == "PUF60", "LOF", "-")) |> 
  filter(germlineMechanism == "LOF") |> pull(gene_name) 
lofGenesGermanTUK <- bind_rows(read_csv(paste0(TUKpath, "data/simpleAnnotGenes.csv"), col_types = cols())) |> 
  filter(germlineMechanism == "LOF") |> filter(testSig == "Exome-wide") |> 
 pull(gene) 
lofSelectionGenes <- unique(c(lofGenesGermanTrio, lofGenesGermanTUK))
write_tsv(tibble(gene = lofSelectionGenes), paste0(localPath, "data/lofSelectionGenesShare.tsv"))

hotspots0 <- read_tsv(paste0(localPath, "data/siteHotspots_dndsBP_sig.tsv"), col_types = cols()) |> 
  bind_rows(read_tsv(paste0(TUKpath, "data/siteHotspots_dnds_sig.tsv"),col_types = cols())) 

hotspots <- hotspots0 |> 
  mutate(var = paste(chr,pos,ref,mut, sep = "_")) |> 
  distinct(var)
write_tsv(hotspots0 |> select(chr:impact)|> distinct(gene), paste0(localPath, "data/hotspotsSelectionShare.tsv"))

allSelectionGenes <- unique(c(hotspots0$gene, lofSelectionGenes))
write_tsv(tibble(gene = allSelectionGenes), paste0(localPath, "data/allSelectionGenesTUK_DDfathers.txt"))
```

#### Process Variants
```{r processVEP}
vepOutput <- read_tsv(paste0(path, "sperm/germanTrio_data/postNextflow/vep/out_allCallsFilterVars.vcf.tsv"), col_types = cols(), comment = "##") |> 
  select(id = `#Uploaded_variation`, csq = Consequence, IMPACT, gene = SYMBOL, gnomADe_AF, PHEN = ClinVar_CLNDN, CLNSIG = ClinVar_CLNSIG, CLNSIGCONF = ClinVar_CLNSIGCONF, cadd = CADD_CADD, trinuc, meth = methylation) |> 
  distinct()

allCallsFilterVarsVEP <- read_tsv(paste0(localPath, "data/allCallsFilterVars.tsv"), col_types = cols()) |> 
  left_join(vepInputTjoint |> select(chr:alt) |> distinct() |> 
              left_join(vepOutput, by = join_by(id)) |> select(-id), by = join_by(chr, start, ref, alt)) |> 
  mutate(csq = str_remove(csq, ",.*")) |> 
  mutate(csq = str_remove(csq, "_variant")) |> 
  mutate(csqSimple = if_else(csq %in% c("3_prime_UTR", "5_prime_UTR", "downstream_gene", "intergenic", "intron", "non_coding_transcript_exon", "splice_region", "upstream_gene"), "non-coding", csq)) |> 
  mutate(csqSimple = if_else(csq %in% c("stop_gained", "splice_acceptor", "splice_donor"), "nonsense+splice", csqSimple)) |> 
  mutate(csqSimple = if_else(csq %in% c("frameshift", "inframe_deletion", "inframe_insertion"), "coding-indel", csqSimple)) |> 
  mutate(cadd = as.numeric(na_if(cadd, "-"))) |> mutate(gnomADe_AF = as.numeric(na_if(gnomADe_AF, "-"))) |>
  # Criteria for labeling variants as ClinVar path or not:
  mutate(ClinVar = if_else(CLNSIG %in% c("Pathogenic/Likely_pathogenic", "Pathogenic", "Likely_pathogenic") | 
                             CLNSIG == "Conflicting_classifications_of_pathogenicity" & 
                              ((str_detect(CLNSIGCONF, "Pathogenic") | str_detect(CLNSIGCONF, "Likely_pathogenic")) & !str_detect(CLNSIGCONF, "enign")
                                 ), T, F)) |> 
  # Annotate likely disease/driver variants
  mutate(likelyDisease = if_else(ClinVar | (gene %in% dddLOF & IMPACT == "HIGH") |
         (gene %in% dddLOF & coalesce(cadd, -Inf) >= 30 & gnomADe_AF < 5e-6), TRUE, FALSE)) |> 
  mutate(var = paste(chr,start,ref,alt, sep = "_")) |> 
  mutate(likelyDriver = if_else((gene %in% lofSelectionGenes & (IMPACT == "HIGH" | coalesce(cadd, -Inf) >= 30)) |
         (var %in% hotspots$var), TRUE, FALSE)) |> 
  # Group meth
  mutate(meth = str_remove(meth, ",.*")) |>
  mutate(meth = as.numeric(str_replace(meth, "-", "-1"))) |>
  mutate(isCpGtoT = if_else((substr(TRI,2,5) == "CG>T" & alt == "T" & ref == "C") | (substr(TRI,2,5) == "CG>T" & alt == "A" & ref == "G"), T,F)) |>
  mutate(methBin = if_else(isCpGtoT & meth > 40, "High", "Mid")) |> mutate(methBin = if_else(isCpGtoT & meth >= 0 & meth <=10, "Low", methBin)) |>
  mutate(meth = if_else(isCpGtoT, meth, 0)) |>
  select(-c(QPOS,  isCpGtoT, var)) |> 
  distinct()


filterVarsVEP <- allCallsFilterVarsVEP |> 
  filter(FILTER == "PASS" & sampleFilter == "FULL") |> select(-FILTER, sampleFilter)

write_tsv(allCallsFilterVarsVEP, paste0(localPath, "data/allCallsFilterVarsVEP.tsv"))
write_tsv(allCallsFilterVarsVEP |> filter(sampleFilter == "FULL"), paste0(localPath, "data/germanTrioAllTargVars.tsv"))
write_tsv(filterVarsVEP, paste0(localPath, "data/germanTrioTargVars.tsv"))

diagVarsVEP <- allCallsFilterVarsVEP |> 
  filter(FILTER == "diagVar") |> select(-FILTER, sampleFilter)
write_tsv(diagVarsVEP, paste0(localPath, "data/diagVarsVEP.tsv"))

# Take a look at excluded variants - any of interest?
excluded <- allCallsFilterVarsVEP |> filter(!FILTER %in% c("PASS", "diagVar")) |> filter(likelyDisease|likelyDriver)

```

#### Process Discarded
```{r processVEP}
# vepOutputDiscarded <- read_tsv(paste0(path, "sperm/germanTrio_data/postNextflow/vep/out_discardedCallsVars.vcf.tsv"), col_types = cols(), comment = "##") |> 
#   select(id = `#Uploaded_variation`, csq = Consequence, IMPACT, gene = SYMBOL, gnomADe_AF, PHEN = ClinVar_CLNDN, CLNSIG = ClinVar_CLNSIG, CLNSIGCONF = ClinVar_CLNSIGCONF, cadd = CADD_CADD, trinuc, meth = methylation) |> 
#   distinct()
# 
# discardedVarsVEP <- read_tsv(paste0(path, "sperm/germanTrio_data/canAppsOutput/discardedCalls.vcf"), col_names = c("chr", "start", "ref", "alt", "FILTER","info", "Sample_PD_ID"), col_types = 'cd-cc-ccc') |> 
#   left_join(discardedVepInput |> select(chr:alt) |> distinct() |> 
#               left_join(vepOutputDiscarded, by = join_by(id)) |> select(-id), by = join_by(chr, start, ref, alt)) |> 
#   mutate(csq = str_remove(csq, ",.*")) |> 
#   mutate(csq = str_remove(csq, "_variant")) |> 
#   mutate(csqSimple = if_else(csq %in% c("3_prime_UTR", "5_prime_UTR", "downstream_gene", "intergenic", "intron", "non_coding_transcript_exon", "splice_region", "upstream_gene"), "non-coding", csq)) |> 
#   mutate(csqSimple = if_else(csq %in% c("stop_gained", "splice_acceptor", "splice_donor"), "nonsense+splice", csqSimple)) |> 
#   mutate(csqSimple = if_else(csq %in% c("frameshift", "inframe_deletion", "inframe_insertion"), "coding-indel", csqSimple)) |> 
#   mutate(cadd = as.numeric(na_if(cadd, "-"))) |> mutate(gnomADe_AF = as.numeric(na_if(gnomADe_AF, "-"))) |>
#   # Criteria for labeling variants as ClinVar path or not:
#   mutate(ClinVar = if_else(CLNSIG %in% c("Pathogenic/Likely_pathogenic", "Pathogenic", "Likely_pathogenic") | 
#                              CLNSIG == "Conflicting_classifications_of_pathogenicity" & 
#                               ((str_detect(CLNSIGCONF, "Pathogenic") | str_detect(CLNSIGCONF, "Likely_pathogenic")) & !str_detect(CLNSIGCONF, "enign")
#                                  ), T, F)) |> 
#   # Annotate likely disease/driver variants
#   mutate(likelyDisease = if_else(ClinVar | (gene %in% dddLOF & IMPACT == "HIGH") |
#          (gene %in% dddLOF & coalesce(cadd, -Inf) >= 30 & gnomADe_AF < 5e-6), TRUE, FALSE)) |> 
#   mutate(var = paste(chr,start,ref,alt, sep = "_")) |> 
#   mutate(likelyDriver = if_else((gene %in% lofSelectionGenes & (IMPACT == "HIGH" | coalesce(cadd, -Inf) >= 30)) |
#          (var %in% hotspots$var), TRUE, FALSE)) |> 
#   # Group meth
#   mutate(meth = str_remove(meth, ",.*")) |>
#   mutate(meth = as.numeric(str_replace(meth, "-", "-1"))) |>
#   distinct()
# 
# write_tsv(discardedVarsVEP, paste0(localPath, "data/discardedVarsVEP.tsv"))
```

#### Process Diagnostic
```{r processVEP}
# Annotation run in #### Process Variants step later on then read back in
vepOutputDiag <- read_tsv(paste0(path, "sperm/germanTrio_data/postNextflow/vep/out_diagnosticVars.vcf.tsv"), col_types = cols(), comment = "##") |> 
  select(id = `#Uploaded_variation`, csq = Consequence, IMPACT, gene = SYMBOL, gnomADe_AF, PHEN = ClinVar_CLNDN, CLNSIG = ClinVar_CLNSIG, CLNSIGCONF = ClinVar_CLNSIGCONF, cadd = CADD_CADD) |> mutate(cadd = as.numeric(cadd)) |> 
  distinct()

diagnosticVars <- diagnosticVarsPos |> rename(gene0 = gene) |> 
  left_join(vepInputdiagnosticVars |> select(chr:alt) |> distinct() |> 
              left_join(vepOutputDiag, by = join_by(id)) |> select(-id), by = join_by(chr, start, ref, alt)) |> 
  mutate(csq = str_remove(csq, ",.*")) |> 
  mutate(csq = str_remove(csq, "_variant")) |> 
  mutate(gnomADe_AF = as.numeric(na_if(gnomADe_AF, "-"))) |>
    # Criteria for labeling variants as ClinVar path or not:
  mutate(ClinVar = if_else(CLNSIG %in% c("Pathogenic/Likely_pathogenic", "Pathogenic", "Likely_pathogenic") | 
                             CLNSIG == "Conflicting_classifications_of_pathogenicity" & 
                              ((str_detect(CLNSIGCONF, "Pathogenic") | str_detect(CLNSIGCONF, "Likely_pathogenic")) & !str_detect(CLNSIGCONF, "enign")
                                 ), T, F)) |> 
  # Annotate likely disease/driver variants
  mutate(likelyDisease = if_else(ClinVar | (gene %in% dddLOF & IMPACT == "HIGH") |
         (gene %in% dddLOF & coalesce(cadd, -Inf) >= 30 & gnomADe_AF < 5e-6), TRUE, FALSE)) |> 
  mutate(var = paste(chr,start,ref,alt, sep = "_")) |> 
  mutate(likelyDriver = if_else((gene %in% lofSelectionGenes & (IMPACT == "HIGH" | coalesce(cadd, -Inf) >= 30)) |
         (var %in% hotspots$var) | start == 	
123279674, TRUE, FALSE)) |> 
  select(-c( missing_TrioWGS, var)) |> 
  arrange(VAR_ID) |> 
  left_join(trioBurdens |> select(FAM_ID, PD_ID) |> distinct() )|> mutate(gnomADe_AF = as.character(gnomADe_AF))|> mutate(cadd = as.character(cadd))

# KIAA2022 annotated as gene name for 2 vars by VEP but current approved name is NEXMIF as annotated in original file
write_tsv(diagnosticVars, paste0(metaPath, "germanTrio_diagnosticVars.tsv"))
```

#### Process trioDNMs
```{r processVEP}
# Annotation run in #### Process Variants step later on then read back in
vepOutputTrio <- read_tsv(paste0(path, "sperm/germanTrio_data/postNextflow/vep/out_trioDNM_Vars.vcf.tsv"), col_types = cols(), comment = "##") |> 
  select(id = `#Uploaded_variation`, csq = Consequence, IMPACT, gene = SYMBOL, gnomADe_AF, PHEN = ClinVar_CLNDN, CLNSIG = ClinVar_CLNSIG, CLNSIGCONF = ClinVar_CLNSIGCONF, cadd = CADD_CADD) |> 
  distinct() |> 
  mutate(diagnosticVar = if_else(id %in% vepInputdiagnosticVars$id, T, F))

trioDNM_Vars <- read_tsv(paste0(path, "sperm/germanTrio_data/dnmData/trioDNMs.tsv"), col_types = 'ci-cc------c', col_names = c("chr","start","ref","alt", "PD_ID")) |> 
  left_join(SupTable3_Trios |> select(PD_ID = PD_ID_Proband, FAM_ID), by = join_by(PD_ID)) |> 
    filter(PD_ID != "PD67073b") |> # Accidentally sent sister's blood and not index
  filter(PD_ID != "PD64157a") |> # Removed for QC
  left_join(vepInputTrioDNMs |> select(chr:alt) |> distinct() |> 
              left_join(vepOutputTrio, by = join_by(id)) |> select(-id), by = join_by(chr, start, ref, alt)) |> 
  mutate(csq = str_remove(csq, ",.*")) |> 
  mutate(csq = str_remove(csq, "_variant")) |> 
  mutate(csqSimple = if_else(csq %in% c("3_prime_UTR", "5_prime_UTR", "downstream_gene", "intergenic", "intron", "non_coding_transcript_exon", "splice_region", "upstream_gene"), "non-coding", csq)) |> 
  mutate(csqSimple = if_else(csq %in% c("stop_gained", "splice_acceptor", "splice_donor"), "nonsense+splice", csqSimple)) |> 
  mutate(csqSimple = if_else(csq %in% c("frameshift", "inframe_deletion", "inframe_insertion"), "coding-indel", csqSimple)) |> 
  mutate(cadd = as.numeric(na_if(cadd, "-"))) |> mutate(gnomADe_AF = as.numeric(na_if(gnomADe_AF, "-"))) |>
  # Criteria for labeling variants as ClinVar path or not:
  mutate(ClinVar = if_else(CLNSIG %in% c("Pathogenic/Likely_pathogenic", "Pathogenic", "Likely_pathogenic")| 
                             CLNSIG == "Conflicting_classifications_of_pathogenicity" & 
                              ((str_detect(CLNSIGCONF, "Pathogenic") | str_detect(CLNSIGCONF, "Likely_pathogenic")) & !str_detect(CLNSIGCONF, "enign")
                                 ), T, F)) |> 
    # Annotate likely disease/driver variants
  mutate(likelyDisease = if_else(ClinVar | (gene %in% dddLOF & IMPACT == "HIGH") |
         (gene %in% dddLOF & coalesce(cadd, -Inf) >= 30 & (is.na(gnomADe_AF) | gnomADe_AF < 5e-6)),
TRUE, FALSE)) |>
  mutate(var = paste(chr,start,ref,alt, sep = "_")) |> 
  mutate(likelyDriver = if_else((gene %in% lofSelectionGenes & (IMPACT == "HIGH" | coalesce(cadd, -Inf) >= 30)) |
         (var %in% hotspots$var), TRUE, FALSE)) |> 
  select(-c(var, ClinVar)) |> 
  mutate(Cohort = "DD fathers") |> 
  distinct() |> 
  #Add VAFs
  left_join(read_tsv(paste0(path, "sperm/germanTrio_data/dnmData/merged_denovo_mutationsVAFs.txt"), col_types = cols()), by = join_by(chr, start, ref, alt, PD_ID)) |> 
  filter(VAF >= 0.3) |> select(-VAF)


write_tsv(trioDNM_Vars, paste0(metaPath, "trioDNM_Vars.tsv"))
```

##### Process trioDNMs Clinical
```{r processVEP}
# Annotation run in #### Process Variants step later on then read back in
vepOutputTrioClin <- read_tsv(paste0(path, "sperm/germanTrio_data/postNextflow/vep/out_trioDNM_Vars.vcf.clinAnnot.tsv"), col_types = cols(), comment = "##") |> 
  mutate( HGVSp_Short = str_remove(HGVSp, "^[^:]+:")) |> 
  mutate(HGVSp_Short = str_replace_all(HGVSp_Short,
    c("Ala"="A","Arg"="R","Asn"="N","Asp"="D","Cys"="C","Gln"="Q","Glu"="E",
      "Gly"="G","His"="H","Ile"="I","Leu"="L","Lys"="K","Met"="M","Phe"="F",
      "Pro"="P","Ser"="S","Thr"="T","Trp"="W","Tyr"="Y","Val"="V","Ter"="*"))) |> 
  select(id = `#Uploaded_variation`, csq = Consequence, IMPACT, HGVSc, HGVSp, HGVSp_Short, gene = SYMBOL, gnomADe_AF, Existing_variation, PHEN = ClinVar_CLNDN, CLNSIG = ClinVar_CLNSIG, CLNSIGCONF = ClinVar_CLNSIGCONF, cadd = CADD_CADD, SIFT, PolyPhen, REF_ALLELE, Allele) |> 
  distinct() |> 
  mutate(diagnosticVar = if_else(id %in% vepInputdiagnosticVars$id, T, F))

trioDNM_VarsClin <- read_tsv(paste0(path, "sperm/germanTrio_data/dnmData/trioDNMs.tsv"), col_types = 'ci-cc------c', col_names = c("chr","start","ref","alt", "PD_ID")) |> 
    filter(PD_ID != "PD67073b") |> # Accidentally sent sister's blood and not index
  filter(PD_ID != "PD64157a") |> # Removed for QC
  left_join(vepInputTrioDNMs |> select(chr:alt) |> distinct() |> 
              left_join(vepOutputTrioClin, by = join_by(id)) |> select(-id), by = join_by(chr, start, ref, alt)) |> 
  mutate(csq = str_remove(csq, ",.*")) |> 
  mutate(csq = str_remove(csq, "_variant")) |> 
  mutate(csqSimple = if_else(csq %in% c("3_prime_UTR", "5_prime_UTR", "downstream_gene", "intergenic", "intron", "non_coding_transcript_exon", "splice_region", "upstream_gene"), "non-coding", csq)) |> 
  mutate(csqSimple = if_else(csq %in% c("stop_gained", "splice_acceptor", "splice_donor"), "nonsense+splice", csqSimple)) |> 
  mutate(csqSimple = if_else(csq %in% c("frameshift", "inframe_deletion", "inframe_insertion"), "coding-indel", csqSimple)) |> 
  filter(IMPACT %in% c("HIGH", "MODERATE") | diagnosticVar)

write_tsv(trioDNM_VarsClin, paste0(metaPath, "trioDNM_VarsClin.tsv"))

forClinicalAnnot <- trioDNM_VarsClin |> 
  left_join(read_csv(paste0(metaPath, "Tables/JoeClinicalReview2.0.csv"), col_types = cols()) |> select(chr = Chromosome, start = Start_Position, REF_ALLELE = Reference_Allele, Allele = ALT, Group:Comments), by = join_by(chr, start, REF_ALLELE, Allele)) |> 
  mutate(Group = if_else(diagnosticVar, "Diagnostic variant", Group)) |> 
  select(Group:Comments, chr:PolyPhen) |> 
  bind_rows(diagnosticVars |> filter(FAM_ID %in% trioBurdens$FAM_ID) |> filter((FAM_ID %in% (diagnosticVarsPos |> filter(called_TrioWGS == "No") |> pull(FAM_ID)))) |> mutate(Group = "Diagnostic variant (Uncalled)") |> select(-c(FAM_ID, VAR_ID, Cohort, gene0, called_TrioWGS, phasing, ClinVar, likelyDisease, likelyDriver))) |> 
  arrange(Group)


table(forClinicalAnnot$Group)

write_csv(forClinicalAnnot, paste0(metaPath, "forClinicalAnnotJoeC_3.1.csv"))
```

#### Process mosaicVars
```{r processVEP}
# Annotation run in #### Process Variants step later on then read back in
vepOutputMosaicVars <- read_tsv(paste0(path, "sperm/germanTrio_data/postNextflow/vep/out_MosaicVars_Vars.vcf.tsv"), col_types = cols(), comment = "##") |> 
  select(id = `#Uploaded_variation`, csq = Consequence, IMPACT, gene = SYMBOL, gnomADe_AF, PHEN = ClinVar_CLNDN, CLNSIG = ClinVar_CLNSIG, CLNSIGCONF = ClinVar_CLNSIGCONF, cadd = CADD_CADD) |> 
  distinct()

mosaicVarsAnnot <- read_tsv(paste0(localPath, "externalData/mosaicVars0.tsv"), col_types = cols()) |> 
  left_join(vepInputMosaicVars |> select(chr:alt) |> distinct() |> 
              left_join(vepOutputMosaicVars, by = join_by(id)) |> select(-id), by = join_by(chr, start, ref, alt)) |> 
  mutate(csq = str_remove(csq, ",.*")) |> 
  mutate(csq = str_remove(csq, "_variant")) |> 
  mutate(csqSimple = if_else(csq %in% c("3_prime_UTR", "5_prime_UTR", "downstream_gene", "intergenic", "intron", "non_coding_transcript_exon", "splice_region", "upstream_gene"), "non-coding", csq)) |> 
  mutate(csqSimple = if_else(csq %in% c("stop_gained", "splice_acceptor", "splice_donor"), "nonsense+splice", csqSimple)) |> 
  mutate(csqSimple = if_else(csq %in% c("frameshift", "inframe_deletion", "inframe_insertion"), "coding-indel", csqSimple)) |> 
  mutate(cadd = as.numeric(na_if(cadd, "-"))) |> mutate(gnomADe_AF = as.numeric(na_if(gnomADe_AF, "-"))) |>
  # Criteria for labeling variants as ClinVar path or not:
  mutate(ClinVar = if_else(CLNSIG %in% c("Pathogenic/Likely_pathogenic", "Pathogenic", "Likely_pathogenic") | 
                             CLNSIG == "Conflicting_classifications_of_pathogenicity" & 
                              ((str_detect(CLNSIGCONF, "Pathogenic") | str_detect(CLNSIGCONF, "Likely_pathogenic")) & !str_detect(CLNSIGCONF, "enign")
                                 ), T, F)) |> 
    # Annotate likely disease/driver variants
  mutate(likelyDisease = if_else(ClinVar | (gene %in% dddLOF & IMPACT == "HIGH") |
         (gene %in% dddLOF & coalesce(cadd, -Inf) >= 30 & gnomADe_AF < 5e-6), TRUE, FALSE)) |> 
  mutate(var = paste(chr,start,ref,alt, sep = "_")) |> 
  mutate(likelyDriver = if_else((gene %in% lofSelectionGenes & (IMPACT == "HIGH" | coalesce(cadd, -Inf) >= 30)) |
         (var %in% hotspots$var), TRUE, FALSE)) |> 
  distinct()

write_tsv(mosaicVarsAnnot, paste0(localPath, "externalData/mosaicVarsAnnot.tsv"))

```

### Run Covered
#### Coverage in cohort
```{bash covFull, eval=FALSE}
#Make summary files with coverage for all indivs at all positions
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/cov
nano getTotalCov.sh

#!/bin/bash

# Set working directory
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/cov/targCut

# Load bedtools
module load bedtools2-2.29.0/

# Full path to the filtered sample list
SAMPLE_LIST=../../../germanTrioSamplesFiltered.txt

### -----------------------------
### PART 1: PD6-only sample set
### -----------------------------

# Get PD6 IDs only (column 1)
PD6_IDS=$(cut -f1 "$SAMPLE_LIST" | grep "^PD6")

# Generate file names and names for bedtools
files_PD6=$(echo "$PD6_IDS" | awk '{print $0 ".cov.bed"}' | tr '\n' ' ')
names_PD6=$(echo "$PD6_IDS" | tr '\n' ' ')

# Run unionbedg for PD6 samples
bedtools unionbedg -i $files_PD6 -header -names $names_PD6 > ../summaryCov.bed

# Sum coverage and output
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/cov/
awk 'OFS="\t" {sum=0; for (col=4; col<=NF; col++) sum += $col; print $1, $2, $3, sum;}' summaryCov.bed | tail -n +2 > totalCov.bed
bgzip -c totalCov.bed > totalCov.bed.gz
tabix -p bed totalCov.bed.gz

bedtools merge -i totalCov.bed > totalCovMerged.bed

# Return to bed directory for second run
cd targCut

### -----------------------------
### PART 2: All samples (Ox version)
### -----------------------------

# Get all IDs (column 1)
ALL_IDS=$(cut -f1 "$SAMPLE_LIST")

# Generate file names and names for bedtools
files_all=$(echo "$ALL_IDS" | awk '{print $0 ".cov.bed"}' | tr '\n' ' ')
names_all=$(echo "$ALL_IDS" | tr '\n' ' ')

# Run unionbedg for all samples
bedtools unionbedg -i $files_all -header -names $names_all > ../summaryCovOx.bed

# Sum coverage and output
cd ..
awk 'OFS="\t" {sum=0; for (col=4; col<=NF; col++) sum += $col; print $1, $2, $3, sum;}' summaryCovOx.bed | tail -n +2 > totalCovOx.bed
bgzip -c totalCovOx.bed > totalCovOx.bed.gz
tabix -p bed totalCovOx.bed.gz



cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/cov
bsub -q long -o jobLogs/sumCov.%J.out -e jobLogs/sumCov.%J.err -R 'select[mem>=10000] rusage[mem=10000] span[hosts=1]' -M 10000 ./getTotalCov.sh

```

#### Full Covered vcf
```{bash CADDcov, eval=FALSE}
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/vep
nano getOpps.sh

#!/bin/bash
set -euo pipefail
# Function to generate VCF from a trinucleotide input file
generate_vcf() {
    local infile=$1
    local outfile=$2

    exec 3> "$outfile"

    while read -r chr pos tri; do
        ref="${tri:1:1}"
        for alt in A C G T; do
            if [ "$alt" != "$ref" ]; then
                echo -e "$chr\t$pos\t.\t$ref\t$alt\t.\t.\t." >&3
            fi
        done
    done < "$infile"

    exec 3>&-

    bgzip -f "$outfile"
    tabix -f "$outfile.gz"
}

# Run for PD6-only set
generate_vcf ../cov/triNucUniq.txt coveredOpps.vcf
# Run for full Ox set
generate_vcf ../cov/triNucUniqOx.txt coveredOppsOx.vcf


cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/vep
bsub -o jobLogs/getOpps.out -e jobLogs/getOpps.err -q normal -R 'select[mem>=100] rusage[mem=100] span[hosts=1]' -M100 ./getOpps.sh



```

#### Running VEP on all covered base pairs to be able to generate background expectations of mutation burden
```{bash maf, eval=FALSE}
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/vep
mkdir -p  splitChrVEP
module load conda/24.9.2
conda activate <hpc_software>/conda_envs/envs/py3.6
OUTPATH="<hpc_scratch>/sperm/germanTrio_data/postNextflow/vep"
export PATH="<hpc_software>/ensembl-vep//htslib:/software/jdk-11.0.2/bin/:$PATH"
export PERL5LIB="<hpc_software>/ensembl-vep/:<hpc_home>/perl5/lib/perl5:<hpc_home>/perl5/lib/perl5/site_perl:<hpc_software>/000_SOFTWARE/lib/perl5:$PERL5LIB"

#!/bin/bash
CHRs=(1 2 3 4 5 6 7 8 9 10 11 12 13 14 15 16 17 18 19 20 21 22 X Y)

for chr in ${CHRs[@]}
do
echo "tabix -h coveredOpps.vcf.gz ${chr} | <hpc_software>/ensembl-vep/vep --format vcf -o splitChrVEP/coveredVEP_${chr}.tsv --tab \
--dir <hpc_software>/000_SOFTWARE/vep/ --species homo_sapiens --cache --offline --force --refseq --no_stats --fork 4 --assembly GRCh37 \
--fasta <hpc_software>/000_SOFTWARE/vep/homo_sapiens/Homo_sapiens.GRCh37.75.dna.primary_assembly.fa \
--symbol --af_gnomad --pick --show_ref_allele \
--custom <hpc_shared>/External_databases/Variants/ClinVar/GRCh37/clinvar_20250729.vcf.gz,ClinVar,vcf,exact,0,CLNDN,CLNSIG,CLNSIGCONF \
--custom ../cov/totalCov.bed.gz,totalCov,bed,exact,0 \
--custom ../cov/triNucUniq.bed.gz,trinuc,bed,exact,0 \
--custom <hpc_scratch>/genomicRegions/methylation/testisMeth19.bed.gz,methylation,bed,exact,0 \
--custom <hpc_scratch>/sperm/germanTrio_data/postNextflow/vep/cadd0.vcf.gz,CADD,vcf,exact,0,CADD" | bsub -e jobLogs/split.${chr}.vep.err -o jobLogs/split.${chr}.vep.out -J ${chr}_vep -n4 -q normal -R 'select[mem>=1000] rusage[mem=1000] span[hosts=1]' -M1000  -env "all"
done

for chr in ${CHRs[@]}
do
echo "tabix -h coveredOppsOx.vcf.gz ${chr} | <hpc_software>/ensembl-vep/vep --format vcf -o splitChrVEP/Ox/coveredVEP_Ox_${chr}.tsv --tab \
--dir <hpc_software>/000_SOFTWARE/vep/ --species homo_sapiens --cache --offline --force --refseq --no_stats --fork 4 --assembly GRCh37 \
--fasta <hpc_software>/000_SOFTWARE/vep/homo_sapiens/Homo_sapiens.GRCh37.75.dna.primary_assembly.fa \
--symbol --af_gnomad --pick --show_ref_allele \
--custom <hpc_shared>/External_databases/Variants/ClinVar/GRCh37/clinvar_20250729.vcf.gz,ClinVar,vcf,exact,0,CLNDN,CLNSIG,CLNSIGCONF \
--custom ../cov/totalCovOx.bed.gz,totalCov,bed,exact,0 \
--custom ../cov/triNucUniqOx.bed.gz,trinuc,bed,exact,0 \
--custom <hpc_scratch>/genomicRegions/methylation/testisMeth19.bed.gz,methylation,bed,exact,0 \
--custom <hpc_scratch>/sperm/germanTrio_data/postNextflow/vep/cadd0.vcf.gz,CADD,vcf,exact,0,CADD" | bsub -e jobLogs/splitOx.${chr}.vep.err -o jobLogs/splitOx.${chr}.vep.out -J ${chr}_Oxvep -n4 -q normal -R 'select[mem>=1000] rusage[mem=1000] span[hosts=1]' -M1000  -env "all"
done

cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/vep
rm jobLogs/split.*
./vepCovered.sh
```

#### Merge and simplify
```{bash maf, eval=FALSE}
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/vep
nano preFilterVep.sh

#!/bin/bash
cat splitChrVEP/*.tsv | grep -v '#' | awk -F'\t' -v OFS='\t' '{split($2, arr, ":"); gsub(/,.*/, "", $7); gsub(/_variant/, "", $7); print arr[1], arr[2], $14, $3, $7, $15, $19, $28, $41, $42, $43, $44, $45, $48}' > vepCoveredProcessed.tsv
# Get unique gene + type
cut -f 1,2,6,7 vepCoveredProcessed.tsv | uniq > vepCoveredGeneImpact.tsv

cat splitChrVEP/Ox/*.tsv | grep -v '#' | awk -F'\t' -v OFS='\t' '{split($2, arr, ":"); gsub(/,.*/, "", $7); gsub(/_variant/, "", $7); print arr[1], arr[2], $14, $3, $7, $15, $19, $28, $41, $42, $43, $44, $45, $48}' > vepCoveredProcessedOx.tsv
# Get unique gene + type
cut -f 1,2,6,7 vepCoveredProcessedOx.tsv | uniq > vepCoveredGeneImpactOx.tsv


cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/vep
bsub -e jobLogs/preFilter.vep.err -o jobLogs/preFilter.vep.out -J preFilter_vep -q normal -R 'select[mem>=1000] rusage[mem=1000] span[hosts=1]' -M1000 ./preFilterVep.sh

```

### Create regions
#### Setup
```{bash setup_regions, eval=FALSE}
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/cov/
mkdir -p mutRegion/
mkdir -p mutRegion/regions/
mkdir -p mutRegion/muts/
mkdir -p mutRegion/twinsUK/
mkdir -p mutRegion/twinsUK/muts/
```

#### Write regions R
```{r regions}
# Coverage
covRegions <- read_tsv(paste0(path, "sperm/germanTrio_data/postNextflow/vep/vepCoveredGeneImpact.tsv"), col_types = 'cicc', col_names = c("chr", "pos", "IMPACT", "gene")) |> 
  mutate(bedStart = pos - 1)
table(covRegions$IMPACT)
table(covRegions$gene)

sigGenesTUK <- read_tsv(paste0("<local_working_dir>/twinsUK_SpermPaperFiles/", "data/sigGenes.tsv"), col_types = cols()) 
exclGenes <- c((sigGenesTUK |> pull(gene_name)), "ASXL1", "TET2", "SEMG1", "ANKRD11", "TSPYL2", "DNMT3A")


write_tsv(covRegions |> filter(IMPACT == "MODIFIER") |> select(chr, bedStart, pos), 
          file = paste0(path, "sperm/germanTrio_data/postNextflow/cov/mutRegion/regions/nonCoding.bed"), col_names = F)
write_tsv(covRegions |> filter(IMPACT != "MODIFIER") |> select(chr, bedStart, pos), 
          file = paste0(path, "sperm/germanTrio_data/postNextflow/cov/mutRegion/regions/coding.bed"), col_names = F)
write_tsv(covRegions |> filter(!(gene %in% exclGenes & IMPACT != "MODIFIER")) |> select(chr, bedStart, pos), 
          file = paste0(path, "sperm/germanTrio_data/postNextflow/cov/mutRegion/regions/exclDrivers.bed"), col_names = F)

# Variants
targVarsBed <- read_tsv(paste0(localPath, "data/germanTrioTargVars.tsv"), col_types = cols()) |> 
  mutate(bedStart = start - 1) |> select(chr, bedStart, start)
write_tsv(targVarsBed, file = paste0(path, "sperm/germanTrio_data/postNextflow/cov/mutRegion/muts/germanTrioTargVars.bed"), col_names = F)

TUKtargVarsBed <- read_tsv(paste0("<local_working_dir>/twinsUK_SpermPaperFiles/", "data/analysisTargVars.tsv"), col_types = cols()) |> 
  mutate(bedStart = start - 1) |> select(chr, bedStart, start)
write_tsv(TUKtargVarsBed, file = paste0(path, "sperm/germanTrio_data/postNextflow/cov/mutRegion/twinsUK/muts/twinsUKtargVars.bed"), col_names = F)
```

#### Merge regions
```{bash maf, eval=FALSE}
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/cov/mutRegion/regions/
module load bedtools2-2.29.0/
sort -k1,1 -k2,2n nonCoding.bed | bedtools merge > nonCoding.merged.bed
sort -k1,1 -k2,2n coding.bed | bedtools merge > coding.merged.bed
sort -k1,1 -k2,2n exclDrivers.bed | bedtools merge > exclDrivers.merged.bed

cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/cov/mutRegion
sort -k1,1 -k2,2n muts/germanTrioTargVars.bed > muts/germanTrioTargVars.sorted.bed
sort -k1,1 -k2,2n twinsUK/muts/twinsUKtargVars.bed > twinsUK/muts/twinsUKtargVars.sorted.bed

```


#### Mut rate by region
```{bash mutRegion, eval=FALSE}

cd <hpc_scratch>/sperm/germanTrio_data/postNextflow
rm cov/jobLogs/mutRegion*
for sample in `cat ../germanTrioSamplesFiltered.txt | cut -f1`;
  do echo $sample
  for region in cov/mutRegion/regions/*.merged.bed
    do 
    # Remove directory path
    filename=$(basename "$region")
    # Remove file extension
    basename="${filename%.merged.*}"
    echo $basename
    bsub -o cov/jobLogs/mutRegion.$sample.$basename.out -e cov/jobLogs/mutRegion.$sample.$basename.err -q normal -R 'select[mem>=200] rusage[mem=200] span[hosts=1]' -M 200 "module load bedtools2-2.29.0/; bedtools intersect -a cov/targCut/$sample.cov.bed -b cov/mutRegion/regions/$basename.merged.bed | cut -f 4 | paste -sd+ - | bc | sed 's/$/\t$sample\t$basename/'  >  cov/mutRegion/$sample.$basename.txt"
  done
done


cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/cov/mutRegion/
cat *.txt > mutRegionsCovs.tsv


cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/cov/mutRegion/

for region in regions/*.merged.bed
  do 
  # Remove directory path
  filename=$(basename "$region")
  # Remove file extension
  basename="${filename%.merged.*}"
  echo $basename
  module load bedtools2-2.29.0/; bedtools intersect -a muts/germanTrioTargVars.sorted.bed -b regions/$basename.merged.bed | cut -f 1,3 | sed "s/$/\t$basename/" >  muts/$basename.txt
done

```

##### Comp Mut rate by region TwinsUK exome samples
```{bash mutRegion, eval=FALSE}

cd <hpc_scratch>/sperm/germanTrio_data/postNextflow
rm cov/jobLogs/mutRegionEx*
for sample in `grep "exome" <hpc_scratch>/sperm/targeted/targetedSamplesFiltered.txt | cut -f1`;
  do echo $sample
  for region in cov/mutRegion/regions/*.merged.bed
    do 
    # Remove directory path
    filename=$(basename "$region")
    # Remove file extension
    basename="${filename%.merged.*}"
    echo $basename
    bsub -o cov/jobLogs/mutRegionEx.$sample.$basename.out -e cov/jobLogs/mutRegionEx.$sample.$basename.err -q normal -R 'select[mem>=2000] rusage[mem=2000] span[hosts=1]' -M 2000 "module load bedtools2-2.29.0/; bedtools intersect -a $T/cov/exCut/$sample.cov.bed -b cov/mutRegion/regions/$basename.merged.bed | cut -f 4 | paste -sd+ - | bc | sed 's/$/\t$sample\t$basename/'  >  cov/mutRegion/twinsUK/$sample.$basename.txt"
  done
done

cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/cov/mutRegion/twinsUK/
cat *.txt > mutRegionsCovs.tsv

cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/cov/mutRegion/

for region in regions/*.merged.bed
  do 
  # Remove directory path
  filename=$(basename "$region")
  # Remove file extension
  basename="${filename%.merged.*}"
  echo $basename
  module load bedtools2-2.29.0/; bedtools intersect -a twinsUK/muts/twinsUKtargVars.sorted.bed -b regions/$basename.merged.bed | cut -f 1,3 | sed "s/$/\t$basename/" >  twinsUK/muts/$basename.txt
done

```



## c) dNdS
First obtain dNdS_basePairCov_mini.zip & unzip 
### Examine gene level coverage
Creates the simple gene level coverage file that can be used in default dnds duplex cov correction
#### Generate files
```{bash dndscov, eval=FALSE}
# Run for targ 
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/dNdS_basePairCov/
mkdir -p covTarg
mkdir -p inputFor_dNdS

cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/dNdS_basePairCov/covTarg 
for sample in `grep "PD6" ../../../germanTrioSamplesFiltered.txt | cut -f1 `;
  do ln -s <hpc_scratch>/sperm/germanTrio_data/postNextflow/cov/source/$sample.cov.bed .
done

cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/dNdS_basePairCov/covTarg 
bsub -e %J.err -o %J.out -q normal -R 'select[mem>=1000] rusage[mem=1000] span[hosts=1]' -M 1000 'module load bedtools2-2.29.0/; perl ../coverage4dndscv.pl *.bed'
```

#### Edit gene lists
```{r dndscovGenes}
# Mean coverage per gene in targeted
covPerGene <- read_tsv(paste0(dndsPath, "covTarg/coverage_info.tsv"), col_types = cols()) |> 
  # I typically filter genes under a certain threshold that have barely any coverage, you can skip this if you want
  filter(dplx_cov_total > 10000000) |> 
  # 
  filter(!gene %in% c("RP11-729L2.2", "CSNK2B-LY6G5B-1181"))
  
vCov <- covPerGene |> pull(dplx_cov_mean)
names(vCov) <- covPerGene |> pull(gene)

save(vCov, file = paste0(localPath, "data/dndsCov.Rdata"))
save(vCov, file = paste0(dndsPath, "inputFor_dNdS/dndsCov.Rdata"))

covPerGene |> mutate(target = "targeted") |> 
  ggplot(aes(dplx_cov_mean, colour = target)) +
  geom_density() +
  scale_x_log10() +
  theme_minimal()
```

### Prep input files
#### Variant files
```{r variantsPenta}
allVars <- germanTrioTargVars |> filter(Cohort == "DD fathers") |> 
  select(sampleID = PD_ID, chr, pos = start, ref, mut = alt) |> 
  distinct()

# TwinsUK control exome vars for global dnds test comparison
tukVars <- read_tsv(paste0(TUKpath, "data/analysisTargVars.tsv"), col_types = cols()) |> filter(nanoseq == "Exome") |> 
  filter(!Sample_PD_ID %in% c("PD50508f_tds0001", "PD50527b_tds0001")) |> #exclude 2 samples which fail this studies QC criteria
  select(sampleID = PD_ID, chr, pos = start, ref, mut = alt) |> 
  distinct()

# Age cut off for comp without age bias
tukVarsAge <- read_tsv(paste0(TUKpath, "data/analysisTargVars.tsv"), col_types = cols()) |> filter(nanoseq == "Exome" & age_at_sampling <63) |> 
    filter(!Sample_PD_ID %in% c("PD50508f_tds0001", "PD50527b_tds0001")) |> #exclude 2 samples which fail this studies QC criteria
  select(sampleID = PD_ID, chr, pos = start, ref, mut = alt) |> 
  distinct()

# Write to cluster for running penta models
write_tsv(allVars, paste0(dndsPath, "inputFor_dNdS/allVars.tsv"))
write_tsv(tukVars, paste0(dndsPath, "inputFor_dNdS/tukVars.tsv"))
write_tsv(tukVarsAge, paste0(dndsPath, "inputFor_dNdS/tukVarsAge.tsv"))
```

###  Build reference
Can run once to create new opportunity matrices to the reference object.
Update if coverage file changes
```{r bp_dnds_setup}
source(paste0(dndsPath, "R/build_bp_refdb.R"))
source(paste0(dndsPath, "R/dndscvAnnotPenta.R"))

RefCDS_bp <- build_bp_refdb(
  refcds_path = paste0(dndsPath, "dataFiles/refcds_hg19.rda"), 
  coverage_path =  paste0(path, "sperm/germanTrio_data/postNextflow/cov/totalCov.bed"), 
  gene_list = names(vCov),
  build_tri = TRUE,
  build_penta = TRUE, 
  build_meth = TRUE,
  methylation_path = paste0(path, "genomicRegions/methylation/testisMeth19.germanTrioCov.bed"),
  genome_fasta = paste0(path, "genomicRegions/genome.fa"), 
  out_path = paste0(dndsPath, "output/germanTrio_refcds_bp.Rdat")
)
```

### Run dNdS
#### Function to save as dndsPath/run_dnds_gene.R
```{r bp_dnds_setup}
#!/usr/bin/env Rscript

args <- commandArgs(trailingOnly = TRUE)

if (length(args) < 2) {
  stop(
    paste0(
      "Usage:\n",
      "  Rscript run_dnds_gene.R <referenceVersion> <geneSet>\n\n",
      "Examples:\n",
      "  Rscript run_dnds_gene.R Tri Panel\n",
      "  Rscript run_dnds_gene.R TriMeth Panel\n",
      "  Rscript run_dnds_gene.R Penta Panel\n",
      "  Rscript run_dnds_gene.R PentaMeth Panel\n"
    )
  )
}

referenceVersion <- args[1]  # Tri / TriMeth / Penta / PentaMeth
geneSet          <- args[2]  # Panel or a filename (without .tsv) in geneLists/

suppressPackageStartupMessages({
  library(dndscv)
  library(tidyverse)
})

# -----------------------------
# Paths (edit once)
# -----------------------------
dndsPath <- "<hpc_scratch>/sperm/germanTrio_data/postNextflow/dNdS_basePairCov/"
inDir    <- file.path(dndsPath, "inputFor_dNdS")
resDir   <- file.path(dndsPath, "output")


# -----------------------------
# Inputs you already have in the new workflow
# -----------------------------
# Keep these exactly as per your current setup
refdb      <- file.path(dndsPath, "output/germanTrio_refcds_bp.Rdat")
genomefile <- "<hpc_references>/pipeline_ref/Human/GRCh37d5/genome.fa"

# Covariates and gene universe
load(file.path(inDir, "covariates_hg19_hg38_epigenome_pcawg.rda"))  # loads covs
load(file.path(inDir, "dndsCov.Rdata"))                             # expects vCov

# Mutations (choose which one you want as the default run input)
mutations_file <- file.path(inDir, "allVars.tsv")
mutations <- read_tsv(mutations_file, col_types = cols()) |>
  dplyr::select(sampleID, chr, pos, ref, mut) |>
  dplyr::distinct()

# Gene list
gene_list <- NULL
if (geneSet == "Panel") {
  gene_list <- names(vCov)
} else {
  gene_list <- read_tsv(file.path(inDir, paste0(geneSet, ".tsv")), col_types = cols()) |>
    dplyr::filter(gene %in% names(vCov)) |>
    dplyr::pull(gene)
}

# Methylation file (only used for *Meth models)
methylationFile <- "<hpc_scratch>/genomicRegions/methylation/testisMeth19.germanTrioCov.bed"

# -----------------------------
# Source your current code (must define dndscv_BP)
# -----------------------------
# Change this to wherever dndscv_BP() lives in your repo
source(file.path(dndsPath, "R/dndsPentaMethBPfunction.R"))

# -----------------------------
# Run
# -----------------------------
common_args <- list(
  mutations = mutations,
  refdb = refdb,
  referenceVersion = referenceVersion,
  cv = covs,
  gene_list = gene_list,
  genomefile = genomefile,
  max_muts_per_gene_per_sample = Inf,
  max_coding_muts_per_sample = Inf,
  outmats = TRUE
)

if (grepl("Meth", referenceVersion)) {
  common_args$methylationFile <- methylationFile
}

out <- do.call(dndscv_BP, common_args)

# -----------------------------
# Write outputs
# -----------------------------
tag <- paste(referenceVersion, geneSet, sep = "_")

write_tsv(out$annotmuts,    file.path(resDir, paste0("annotmuts_", tag, ".tsv")))
write_tsv(out$globaldnds,   file.path(resDir, paste0("globaldnds_", tag, ".tsv")))
write_tsv(out$sel_cv,       file.path(resDir, paste0("sel_cv_", tag, ".tsv")))
write_tsv(out$mle_submodel, file.path(resDir, paste0("mle_submodel_", tag, ".tsv")))

aic <- AIC(out$poissmodel)
write_tsv(tibble(AIC = aic), file.path(resDir, paste0("aic_", tag, ".tsv")))

save(out, file = file.path(resDir, paste0("dndsout_", tag, ".Rdata")))
```

#### Submit jobs
```{bash dndsComp, eval=FALSE}
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/dNdS_basePairCov
mkdir -p jobLogs
rm jobLogs/*

bsub -o jobLogs/Tri_Panel.out -e jobLogs/Tri_Panel.err -q normal -R 'select[mem>=2000] rusage[mem=2000] span[hosts=1]' -M 2000 \
  Rscript run_dnds_gene.R Tri Panel

bsub -o jobLogs/TriMeth_Panel.out  -e jobLogs/TriMeth_Panel.err  -q normal -R 'select[mem>=2000] rusage[mem=2000] span[hosts=1]' -M 2000 \
  Rscript run_dnds_gene.R TriMeth Panel

bsub -o jobLogs/Penta_Panel.out    -e jobLogs/Penta_Panel.err    -q normal -R 'select[mem>=12000] rusage[mem=12000] span[hosts=1]' -M 12000 \
  Rscript run_dnds_gene.R Penta Panel

bsub -o jobLogs/PentaMeth_Panel.out -e jobLogs/PentaMeth_Panel.err -q normal -R 'select[mem>=12000] rusage[mem=12000] span[hosts=1]' -M 12000 \
  Rscript run_dnds_gene.R PentaMeth Panel

```

##### Submit TUK comp 
edited run_dnds_gene.R to run_dnds_geneTUK.R which has "inputFor_dNdS/mutationsInput/tukVarsMeth.tsv" as mutations input file and TUK output name
```{bash dndsComp, eval=FALSE}
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/dnds/inputFor_dNdS
bsub -o jobLogs/methtriTUK.out -e jobLogs/methtriTUK.err -q normal -R 'select[mem>=1200] rusage[mem=1200] span[hosts=1]' -M 1200 Rscript runPentaTUK.R TriMeth Panel
bsub -o jobLogs/methtriTUKage.out -e jobLogs/methtriTUKage.err -q normal -R 'select[mem>=1200] rusage[mem=1200] span[hosts=1]' -M 1200 Rscript runPentaTUKage.R TriMeth Panel
```

#### Gene lists
```{r genelists}

# Create the methylation matrix (same as before)
#Tri meth sub model
# TRI -------------------------------------------------------------
nt = c("A", "C", "G", "T")
trinucs = paste(rep(nt, each = 16, times = 1), rep(nt, each = 4, times = 4), rep(nt, each = 1, times = 16), sep = "")
trinucinds = setNames(1:64, trinucs)
trinucsubs = NULL
for (j in 1:length(trinucs)) {
  trinucsubs = c(trinucsubs, paste(trinucs[j], paste(substr(trinucs[j], 
                                                            1, 1), setdiff(nt, substr(trinucs[j], 2, 2)), substr(trinucs[j], 
                                                                                                                 3, 3), sep = ""), sep = ">"))}
trinucsubsind = setNames(1:192, trinucsubs)
# Meth
methSubs = trinucsubs
for (j in 1:length(trinucsubs)) {
  # Check for C>T substitution at CpG site: C>T means 6th character is "T"
  if (substr(trinucsubs[j], 6, 6) == "T" && substr(trinucsubs[j], 2, 2) == "C" && substr(trinucsubs[j], 3, 3) == "G") {
    # Add z_high and z_low prefixed substitutions for C>T at CpG
    methSubs = c(methSubs, 
                 paste0("z_High", trinucsubs[j]),
                 paste0("z_Low", trinucsubs[j]))
  }
  # Check for G>A substitution at reverse CpG site: G>A means 6th character is "A"
  if (substr(trinucsubs[j], 6, 6) == "A" && substr(trinucsubs[j], 2, 2) == "G" && substr(trinucsubs[j], 1, 1) == "C") {
    # Add z_high and z_low prefixed substitutions for G>A at reverse CpG
    methSubs = c(methSubs, 
                 paste0("z_High", trinucsubs[j]),
                 paste0("z_Low", trinucsubs[j]))
  }
}
# Create indices for methSubs
methSubsInd = setNames(1:length(methSubs), methSubs)
M_matrix = array("",dim=c(length(methSubs), 4))
M_matrix[,1] = methSubs
M_matrix[,2] = paste(methSubs,"*wmis",sep="")
M_matrix[,3] = paste(methSubs,"*wnon",sep="")
M_matrix[,4] = paste(methSubs,"*wspl",sep="")
substmodelMeth <- M_matrix

# Source the custom function
source(paste0("<hpc_scratch>/sperm/germanTrio_data/postNextflow/dnds/", "inputFor_dNdS/geneSetEnrich/genesetdndsMutCount.R"))

# For new genes in TUK data
load(paste0("<hpc_scratch>/sperm/germanTrio_data/postNextflow/dnds/", "inputFor_dNdS/results/dndsout_TriMethPanel.tuk.Rdata"))
geneList <- sigComp |> filter(dndsLabel == "DD fathers") |> pull(gene_name)
geneSetEnrichNew <- genesetdndsMutCount(output, gene_list = geneList, sm = substmodelMeth)

# For old genes in DD data
load(paste0(dndsPath, "output/dndsout_TriMeth_Panel.Rdata"))
geneList2 <- sigComp |> filter(dndsLabel == "Controls") |> pull(gene_name)
geneSetEnrichOld <- genesetdndsMutCount(out, gene_list = geneList2, sm = substmodelMeth)
```


### Model comp summary
#### Basic model
```{r dndsGlobalComp}
# No corrections
load(paste0(dndsPath, "inputFor_dNdS/dndsCov.Rdata"))
load(paste0(path, "dndsObjects/covariates_hg19_hg38_epigenome_pcawg.rda"))  # loads covs
allVars <- read_tsv(paste0(dndsPath, "inputFor_dNdS/allVars.tsv"), col_types = cols())

dndsBasic <- dndscv(allVars, max_muts_per_gene_per_sample = Inf, max_coding_muts_per_sample = Inf,
                            outmats = T, cv = covs, gene_list = names(vCov))

dndsGeneLevel <- dndscv(allVars, max_muts_per_gene_per_sample = Inf, max_coding_muts_per_sample = Inf,
                            outmats = F, dc = vCov, cv = covs, gene_list = names(vCov))

```

#### Compare global
```{r dndsGlobalComp}

# Collect globaldnds values
compGlobal_dNdS <- dndsBasic$globaldnds  |> mutate(test = "Basic") |> 
  bind_rows(read_tsv(paste0(dndsPath, "output/globaldnds_Tri_Panel.tsv"), col_types = cols()) |> 
            mutate(test = "BasePairCov")) |> 
  bind_rows(read_tsv(paste0(dndsPath, "output/globaldnds_TriMeth_Panel.tsv"), col_types = cols()) |> 
          mutate(test = "BasePairCov + CpGmeth")) |> 
  bind_rows(read_tsv(paste0(dndsPath, "output/globaldnds_PentaMeth_Panel.tsv"), col_types = cols()) |> 
              mutate(test = "BasePairCov + CpGmeth + Penta")) |> 
  mutate(name = str_remove(name, "w"))
write_tsv(compGlobal_dNdS, paste0(localPath, "data/compGlobal_dNdS.tsv"))

p_compGlobal_dNdS <- compGlobal_dNdS |> 
  filter(name != "tru") |> 
  mutate(test = fct_inorder(test)) |> 
  ggplot(aes(x = name, y = mle, colour = test)) +
  geom_pointrange(aes(ymin = cilow, ymax = cihigh), position=position_dodge(width = 0.5)) +
  geom_hline(yintercept = 1, linetype = "dashed") + 
  scale_colour_discrete(name = "dNdScv model") +
  ylab("Global dN/dS") + xlab("Mutation class") +
  theme_pubr() +
  theme(legend.position = c(0.3, 0.8), legend.text = element_text(size = 10), legend.title = element_text(size = 12), axis.title.x = element_blank())

p_compGlobal_dNdS
```

#### Compare sig genes
```{r dndsGlobalComp, fig.height = 7, fig.width=4}

basic_ranked_genes <- read_tsv(paste0(dndsPath, "output/sel_cv_TriMeth_Panel.tsv"), col_types = cols()) |>
  arrange(-qglobal_cv) |> 
  pull(gene_name)
compGene_dNdS <- read_tsv(paste0(dndsPath, "output/sel_cv_Tri_Panel.tsv"), col_types = cols()) |> 
  mutate(test = "BasePairCov") |> 
  bind_rows(read_tsv(paste0(dndsPath, "output/sel_cv_TriMeth_Panel.tsv"), col_types = cols()) |> 
              mutate(test = "BasePairCov +\nCpGmeth")) |> 
  bind_rows(
    read_tsv(paste0(dndsPath, "output/sel_cv_PentaMeth_Panel.tsv"), col_types = cols()) |> 
      mutate(test = "BasePairCov +\nCpGmeth +\nPenta")) |> 
  bind_rows(
    dndsBasic$sel_cv |> 
      mutate(test = "Basic")) |> 
  filter(qglobal_cv < 0.01) |>
  select(gene_name, test, qglobal_cv) |> 
  mutate(Significant = ifelse(qglobal_cv < .0001, "***", ifelse(qglobal_cv < .001, "**", ifelse(qglobal_cv < .01, "*", " ")))) |> 
  complete(gene_name, test, fill = list(Significant = "NS")) |> 
  mutate(Significant = factor(Significant, levels = c("***", "**", "*", "NS"))) |> # Set order for significance
  mutate(test = factor(test, levels = c("Basic", "BasePairCov", "BasePairCov +\nCpGmeth", "BasePairCov +\nCpGmeth +\nPenta"))) |> # Set order for significance
  mutate(gene_name = factor(gene_name, levels = basic_ranked_genes)) |> 
  filter(!gene_name %in% c("SEMG1", "MECP2"))


p_compGene_dNdS <- ggplot(compGene_dNdS, aes(x = test, y = gene_name, label = Significant)) +
  geom_tile(color = "black", fill = "white") +
  geom_text() +
  labs(x = "Test Type", y = "Gene", text = "Significance") +
  theme_pubr() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 9), axis.text.y = element_text(size = 9), legend.position = "right",
        legend.text = element_text(size = 8), legend.title = element_text(size = 12),axis.title.x = element_blank(),
        axis.line.x = element_blank(), axis.line.y = element_blank(), axis.ticks = element_blank())

p_compGene_dNdS

```

#### Penta context
Check if any penta contexts are signficantly different from trinucleotide contexts
```{r dndsPentaContext}
tri <- read_tsv(paste0(dndsPath, "output/mle_submodel_TriMeth_Panel.tsv"), col_types = cols())|> 
  filter(!name %in% c("wmis", "wspl", "wnon")) |> 
    # Handle meth
  mutate(meth = if_else(str_detect(name, "z_High"), "High", "Mid")) |> mutate(meth = if_else(str_detect(name, "z_Low"), "Low", meth)) |> 
  mutate(name = if_else(str_detect(name, "z_High"), str_sub(name, 7,14), name)) |> mutate(name = if_else(str_detect(name, "z_Low"), str_sub(name, 6,13), name)) |> 
  dplyr::rename(tri = name, triRate = mle, tricilow = cilow, tricihigh = cihigh)

penta <- read_tsv(paste0(dndsPath, "output/mle_submodel_PentaMeth_Panel.tsv"), col_types = cols()) |> 
  # Remove no estimate
  filter(!name %in% c("wmis", "wspl", "wnon")) |> 
  # Handle meth
  mutate(meth = if_else(str_detect(name, "z_High"), "High", "Mid")) |> mutate(meth = if_else(str_detect(name, "z_Low"), "Low", meth)) |> 
  mutate(name = if_else(str_detect(name, "z_High"), str_sub(name, 7,18), name)) |> mutate(name = if_else(str_detect(name, "z_Low"), str_sub(name, 6,17), name)) |> 
  dplyr::rename(penta = name, pentaRate = mle, pentacilow = cilow, pentacihigh = cihigh) |> 
  mutate(tri = paste0(str_sub(penta,2,4), ">", str_sub(penta,8,10)))

# Function to calculate standard error from confidence interval
calc_se <- function(ci_high, ci_low) {
  (ci_high - ci_low) / (2 * 1.96)  # CI to SE conversion assuming 95% CI
}

# Calculate the difference in rates, their standard errors, z-scores, and FDR-corrected p-values
pentaCompare <- penta |> left_join(tri, join_by(meth, tri)) |>
  mutate(
    # Calculate standard errors for pentanucleotide and trinucleotide contexts
    penta_se = calc_se(pentacihigh, pentacilow),
    tri_se = calc_se(tricihigh, tricilow),
    
    # Calculate the difference in mutation rates
    rate_diff = pentaRate - triRate,
    
    # Combined standard error for the difference
    diff_se = sqrt(penta_se^2 + tri_se^2),
    
    # Calculate z-score for the difference
    z_score = rate_diff / diff_se,
    
    # p-value from z-score (two-tailed test)
    p_value = 2 * pnorm(-abs(z_score)),
    
    # FDR-corrected p-value (Benjamini-Hochberg correction)
    p_adj_fdr = p.adjust(p_value, method = "fdr")
  )

pentaSig = pentaCompare |> arrange(p_adj_fdr) |> mutate(sig = if_else(p_adj_fdr < 0.05, T, F))

```

### Site dnds
#### Basic (default dnds)
```{r standardHostpot_dnds}
hotspots_dndsBasic = sitednds(dndsBasic, min_recurr = 3, gene_list = names(vCov))

hotspots_dndsBasicSig <- hotspots_dndsBasic$recursites |>
  mutate(across(c(freq, dnds, pval), \(x) as.numeric(vapply(x, \(z) unname(z)[1], numeric(1)))),
         significant = if_else(qval < 0.01, "*", "")) |> 
  filter(significant == "*")

write_tsv(hotspots_dndsBasicSig, paste0(localPath, "data/siteHotspots_dnds_sig.tsv"))
```

#### Base pair coverage corrected
BP-aware sitednds uses duplex exposure at the exact base (dx_cov): expected mu is proportional to how many cells/molecules were observed there.
```{r bpHostpot_dnds}

source(paste0(dndsPath, "R/sitedndsBP.R"))
load(paste0(dndsPath, "output/dndsout_Tri_Panel.Rdata"))

hotspotsBP <- sitednds(
  dndsout = out,
  refdb_bp = paste0(dndsPath, "output/germanTrio_refcds_bp.Rdat"),
  coverageFile = "<hpc_scratch>/sperm/germanTrio_data/postNextflow/cov1/totalCov.bed",
  min_recurr = 3,
  gene_list = names(vCov)
)

hotspotsBPSig <- hotspotsBP$recursites |>
  mutate(significant = if_else(qval < 0.01, "*", "")) 

write_tsv(hotspotsBPSig |> filter(significant == "*") |> select(-significant), paste0(localPath, "data/siteHotspots_dndsBP_sig.tsv"))
write_tsv(hotspotsBPSig, paste0(localPath, "data/siteHotspots_dndsBP_all.tsv"))

```

##### SupFig 3
```{r dndsGlobalComp, fig.height = 6, fig.width=10}
p_compGlobal_dNdS + p_compGene_dNdS + plot_layout(widths = c(2,1)) + plot_annotation(tag_levels = "A") &
    theme(plot.tag = element_text(size = 14, face = "bold"))

png(paste0(manuscriptPath, 'Figures/SupFig3_dndsModel.png'), height=6, width=10, res=300, units = "in")
p_compGlobal_dNdS + p_compGene_dNdS + plot_layout(widths = c(2,1)) + plot_annotation(tag_levels = "A") &
    theme(plot.tag = element_text(size = 14, face = "bold"))
dev.off()
pdf(paste0(manuscriptPath, 'Figures/SupFig3_dndsModel.pdf'), height=6, width=10, onefile = F)
p_compGlobal_dNdS + p_compGene_dNdS + plot_layout(widths = c(2,1)) + plot_annotation(tag_levels = "A") &
    theme(plot.tag = element_text(size = 14, face = "bold"))
dev.off()
```

### Review Analysis: SCN1A dN/dS exclusion
#### Prepare filtered variant input
```{r scn1a_dnds_prep}
# Filter germanTrioTargVars to DD fathers excluding SCN1A-ascertained sperm samples.
# scn1a_sperm_ids is defined in the scn1a_ids chunk above.
allVars_scn1a <- germanTrioTargVars |>
  filter(Cohort == "DD fathers", !Sample_PD_ID %in% scn1a_sperm_ids) |>
  select(sampleID = PD_ID, chr, pos = start, ref, mut = alt) |>
  distinct()

cat("Variants in manuscript allVars:    ", nrow(germanTrioTargVars), "\n")
cat("Variants in SCN1A-excluded allVars:", nrow(allVars_scn1a), "\n")
cat("Samples removed:", length(scn1a_sperm_ids), "\n")

write_tsv(allVars_scn1a, paste0(dndsPath, "inputFor_dNdS/allVars_scn1a.tsv"))
```

#### Write SCN1A-excluded run script for HPC
```{r scn1a_dnds_script, eval=FALSE}
# Writes run_dnds_gene_scn1a.R to dndsPath — identical to run_dnds_gene.R except
# the mutations input points to allVars_scn1a.tsv and outputs are tagged _scn1a.
script_lines <- readLines(paste0(dndsPath, "run_dnds_gene.R"))
script_lines <- gsub(
  'mutations_file <- file.path(inDir, "allVars.tsv")',
  'mutations_file <- file.path(inDir, "allVars_scn1a.tsv")',
  script_lines, fixed = TRUE
)
script_lines <- gsub(
  'tag <- paste(referenceVersion, geneSet, sep = "_")',
  'tag <- paste(referenceVersion, geneSet, "scn1a", sep = "_")',
  script_lines, fixed = TRUE
)
writeLines(script_lines, paste0(dndsPath, "run_dnds_gene_scn1a.R"))
cat("Written to", paste0(dndsPath, "run_dnds_gene_scn1a.R"), "\n")
```

#### Submit HPC job
```{bash scn1a_dnds_submit, eval=FALSE}
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/dNdS_basePairCov
bsub -e jobLogs/%J_scn1a_TriMeth.err \
     -o jobLogs/%J_scn1a_TriMeth.out \
     -q long \
     -R 'select[mem>=8000] rusage[mem=8000] span[hosts=1]' \
     -M 8000 \
     "Rscript run_dnds_gene_scn1a.R TriMeth Panel"
```

#### Compare global dN/dS: manuscript vs SCN1A excluded
```{r scn1a_dnds_global, fig.width=4, fig.height=4}
# Extend p_globaldnds (Fig 2a) with a third bar for SCN1A-excluded DD fathers.
# Matches Fig 2a exactly: wall only, geom_col + errorbar, dataset_colours palette.
scn1a_colour <- "#f4b88c"  # lighter shade of DD fathers orange

global_scn1a_data <- bind_rows(
  read_tsv(paste0(dndsPath, "output/globaldnds_TriMeth_Panel.tsv"), col_types = cols()) |>
    mutate(dataset = "DD fathers"),
  read_tsv(paste0("<hpc_scratch>/sperm/germanTrio_data/postNextflow/dnds/",
                  "inputFor_dNdS/results/globaldnds_TriMethPanel.tukAge.tsv"),
           col_types = cols()) |>
    mutate(dataset = "Controls"),
  read_tsv(paste0(dndsPath, "output/globaldnds_TriMeth_Panel_scn1a.tsv"), col_types = cols()) |>
    mutate(dataset = "DD fathers\n(SCN1A excl.)")
) |>
  filter(name == "wall") |>
  mutate(dataset = factor(dataset,
                          levels = c("DD fathers", "DD fathers\n(SCN1A excl.)", "Controls")))

p_global_scn1a <- ggplot(global_scn1a_data,
       aes(x = dataset, y = mle, fill = dataset)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(aes(ymin = cilow, ymax = cihigh),
                position = position_dodge(width = 0.8),
                width = 0.2, colour = "black") +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_fill_manual(values = c(dataset_colours,
                                "DD fathers\n(SCN1A excl.)" = scn1a_colour)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = NULL, y = "Global dN/dS (220 gene panel)") +
  theme_pubr() +
  theme(axis.title.x = element_blank(), legend.position = "none",
        axis.text.x  = element_text(size = 9))

p_global_scn1a
```

#### Compare gene-level significance: manuscript vs SCN1A excluded
```{r scn1a_dnds_genes, fig.width=5, fig.height=6}
# Manuscript gene-level results (TriMeth)
genes_manuscript <- read_tsv(paste0(dndsPath, "output/sel_cv_TriMeth_Panel.tsv"),
                              col_types = cols()) |>
  mutate(dataset = "Manuscript")

# SCN1A-excluded gene-level results
genes_scn1a <- read_tsv(paste0(dndsPath, "output/sel_cv_TriMeth_Panel_scn1a.tsv"),
                         col_types = cols()) |>
  mutate(dataset = "SCN1A excluded")

# Universe: genes significant in the manuscript (q < 0.01), same filter as main figure
sig_genes_manuscript <- genes_manuscript |>
  filter(qglobal_cv < 0.01) |>
  pull(gene_name)

gene_comp <- bind_rows(genes_manuscript, genes_scn1a) |>
  filter(gene_name %in% sig_genes_manuscript) |>
  select(gene_name, dataset, qglobal_cv) |>
  mutate(
    sig_label = case_when(
      qglobal_cv < 0.0001 ~ "***",
      qglobal_cv < 0.001  ~ "**",
      qglobal_cv < 0.01   ~ "*",
      TRUE                ~ "NS"
    ),
    sig_label = factor(sig_label, levels = c("***", "**", "*", "NS")),
    dataset   = factor(dataset, levels = c("Manuscript", "SCN1A excluded")),
    gene_name = factor(gene_name,
                       levels = genes_manuscript |>
                         arrange(qglobal_cv) |>
                         pull(gene_name))
  )

p_genes_scn1a <- ggplot(gene_comp, aes(x = dataset, y = gene_name, label = sig_label)) +
  geom_tile(colour = "black", fill = "white") +
  geom_text(size = 4) +
  labs(x = NULL, y = NULL) +
  theme_pubr() +
  theme(axis.text.x  = element_text(angle = 45, hjust = 1),
        axis.line.x  = element_blank(), axis.line.y  = element_blank(),
        axis.ticks   = element_blank())

p_genes_scn1a
```


## d) Observed VAFs
### TwinsUK control rates
```{r twinsUKrates}
tukExomeVars <- read_tsv(paste0(TUKpath, "data/analysisTargVars.tsv"), col_types = cols()) |> 
  filter(nanoseq == "Exome") |> filter(!Sample_PD_ID %in% c("PD50508f_tds0001", "PD50527b_tds0001"))
  
tukVars0 <- tukExomeVars |> 
  mutate(varID = paste(chr, start,ref,alt,Sample_PD_ID, sep = "_")) |> 
  mutate(bedStart = start - 1) 
write_tsv(tukVars0 |> select(chr, bedStart, start, varID), paste0(path, "sperm/germanTrio_data/postNextflow/diseaseRates/tukVars0.bed"), col_names = F)

tukAllVars <- read_tsv(paste0(TUKpath, "data/analysisTargVars.tsv"), col_types = cols())
tukAllVars0 <- tukAllVars |> 
  mutate(varID = paste(chr, start,ref,alt,Sample_PD_ID, sep = "_")) |> 
  mutate(bedStart = start - 1) 
write_tsv(tukAllVars0 |> select(chr, bedStart, start, varID), paste0(path, "sperm/germanTrio_data/postNextflow/diseaseRates/tukVarsAll0.bed"), col_names = F)
```

#### Vars intersect with german panel
```{bash intPanel, eval=FALSE}
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/diseaseRates
sort -k1,1 -k2,2n tukVars0.bed > tukVars0_sorted.bed

module load bedtools2-2.29.0/
bedtools intersect -a tukVars0_sorted.bed -b ../cov/totalCovMerged.bed -wa > tukVarsGermanPanel.bed
```

#### Cov intersect with german panel
```{bash intPanel, eval=FALSE}
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/cov/twinsUKcontrol/

for sample in `grep "exome" <hpc_scratch>/sperm/targeted/targetedSamplesFiltered.txt | cut -f1`;
  do echo $sample
  bsub -o jobLogs/covCut.$sample.out -e jobLogs/covCut.$sample.err -q normal -R 'select[mem>=4000] rusage[mem=4000] span[hosts=1]' -M 4000 "module load bedtools2-2.29.0/;  bedtools intersect -sorted -a $T/cov/exCut/$sample.cov.bed -b ../totalCovMerged.bed > $sample.germanPanelCovEx.bed"
done

cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/cov/twinsUKcontrol/
for sample in `grep "targeted" <hpc_scratch>/sperm/targeted/targetedSamplesFiltered.txt | cut -f1`;
  do echo $sample
  bsub -o jobLogs/covCut.$sample.out -e jobLogs/covCut.$sample.err -q normal -R 'select[mem>=100] rusage[mem=100] span[hosts=1]' -M 100 "module load bedtools2-2.29.0/;  bedtools intersect -sorted -a $T/cov/targCut/$sample.cov.bed -b ../totalCovMerged.bed > $sample.germanPanelCovTarg.bed"
done

cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/cov/twinsUKcontrol/
rm twinsUK_germanPanelCov.tsv

for sample in $(cut -f1 <hpc_scratch>/sperm/targeted/targetedSamplesFiltered.txt); do
  echo $sample
  sum=$(awk '{sum+=$4} END {print sum}' $sample.*.bed)
  count=$(wc -l < $sample.*.bed | awk '{sum+=$1} END {print sum}')
  echo -e "$sum\t$count\t$sample" >> twinsUK_germanPanelCov.tsv
done

```

### Calculate obs VAFs
```{r vafSummary}
generate_vaf_summary <- function(variant_data) {
  variant_data |> 
    filter(csq == "synonymous" & TYPE == "snv") |> mutate(vafGroup = "synonymous") |> 
    bind_rows(variant_data |> filter(csq == "missense" & TYPE == "snv") |> mutate(vafGroup = "missense"))  |> 
    bind_rows(variant_data |> filter(IMPACT != "MODIFIER" & TYPE == "snv") |> mutate(vafGroup = "coding"))  |> 
    bind_rows(variant_data |> filter(IMPACT == "MODIFIER" & TYPE == "snv") |> mutate(vafGroup = "non-coding"))  |> 
    bind_rows(variant_data |> filter(IMPACT == "HIGH" & TYPE == "snv") |> mutate(vafGroup = "LOF_snv"))  |> 
    bind_rows(variant_data |> filter(ClinVar) |> mutate(vafGroup = "Clin")) |> 
    bind_rows(variant_data |> filter((ClinVar | (gene %in% dddLOF & IMPACT == "HIGH"))) |> mutate(vafGroup = "ClinDDlof")) |> 
    bind_rows(variant_data |> filter((ClinVar | (gene %in% dddLOF & IMPACT == "HIGH") | (gene %in% dddLOF & cadd >= 30 & gnomADe_AF < 5e-6))) |> mutate(vafGroup = "ClinDDlofCADD")) |> 
    bind_rows(variant_data |> filter(ClinVar & str_detect(PHEN, "RASopathy")) |> mutate(vafGroup = "Rasopathy")) |> 
    bind_rows(variant_data |> filter(ClinVar & str_detect(PHEN, "Hereditary_cancer-predisposing_syndrome")) |> mutate(vafGroup = "CancerPredisposing")) |> 
    bind_rows(variant_data |> filter(start == 1806119 & str_detect(PHEN, "Achondroplasia") & ClinVar) |>  mutate(vafGroup = "Achondroplasia")) |> 
    bind_rows(variant_data |> filter(str_detect(PHEN, "oonan")) |> filter(!str_detect(PHEN, "eurofibromatosis") & ClinVar) |> mutate(vafGroup = "Noonan")) |>
    bind_rows(variant_data |> mutate(var = paste(chr,start,ref,alt, sep = "_")) |> 
      filter((gene %in% lofSelectionGenes & (IMPACT == "HIGH" | cadd >= 30)) | 
             (var %in% hotspots$var)) |> mutate(vafGroup = "allPos")) |> 
    bind_rows(variant_data |> mutate(var = paste(chr,start,ref,alt, sep = "_")) |> 
      filter((ClinVar | (gene %in% dddLOF & IMPACT == "HIGH") | (gene %in% dddLOF & cadd >= 30 & gnomADe_AF < 5e-6)) & (gene %in% lofSelectionGenes & (IMPACT == "HIGH" | cadd >= 30)) | 
             (var %in% hotspots$var)) |> mutate(vafGroup = "posPath")) |>
    bind_rows(variant_data |> filter(gene == "NF1") |> filter((ClinVar & str_detect(PHEN, "eurofibromatosis")) | IMPACT == "HIGH" | cadd >= 30) |>
      mutate(vafGroup = "NF1")) |> 
    bind_rows(variant_data |> filter((chr == "10" & start == 123279677 & alt %in% c("C")) | (chr == "10" & start == 123279674 & alt %in% c("C"))) |>
      mutate(vafGroup = "Apert")) |>
    bind_rows(variant_data |> filter((str_detect(PHEN, "rouzon") | str_detect(PHEN, "feiffe")) & gene == "FGFR2" & ClinVar) |> mutate(vafGroup = "Crouzon/Pfeiffer")) |> 
    bind_rows(variant_data |> filter((str_detect(PHEN, "Thanatophoric_dysplasia")) & start!= 1806119 & ClinVar)|> mutate(vafGroup = "Thanatophoric dysplasia")) |> 
    bind_rows(variant_data |> filter(IMPACT != "MODIFIER" & TYPE != "snv") |> mutate(vafGroup = "coding_indel"))  |> 
    bind_rows(variant_data |> filter(IMPACT == "MODIFIER" & TYPE != "snv") |> mutate(vafGroup = "non-coding_indel"))  |> 
    group_by(Sample_PD_ID, Age, vafGroup) |> 
    summarize(vafSum = sum(DUPLEX_VAF, na.rm = TRUE), countSum = sum(TIMES_CALLED), .groups = "drop") |> 
    complete(nesting(Sample_PD_ID, Age), vafGroup, fill = list(countSum=0, vafSum = 0)) |> 
    mutate(method = "obs")
}

dddLOF <- read_tsv(paste0(localPath, "data/dddLOF.txt"), col_types = cols()) |> pull(gene)
lofSelectionGenes <- read_tsv(paste0(localPath, "data/lofSelectionGenes.tsv"), col_types = cols()) |> pull(gene)
hotspots <- read_tsv(paste0(localPath, "data/hotspots.tsv"), col_types = cols())

tukExomeVars <- read_tsv(paste0(TUKpath, "data/analysisTargVars.tsv"), col_types = cols()) |> 
  filter(nanoseq == "Exome") |> filter(!Sample_PD_ID %in% c("PD50508f_tds0001", "PD50527b_tds0001"))
tukGermanPanelVars <- tukVars0 |> filter(varID %in% (read_tsv(paste0(path, "sperm/germanTrio_data/postNextflow/diseaseRates/tukVarsGermanPanel.bed"), col_types = 'cccc', col_names = c("a", "b", "c", "varID")) |> pull(varID)))

tukSamplesFiltered <- read_tsv(paste0("<local_working_dir>/nanoSampleMeta/", "targetedSamplesFiltered.tsv"), col_types = cols())
germanTrioTargVars <- read_tsv(paste0(localPath, "data/germanTrioTargVars.tsv"), col_types = cols())

tukVafSums <- generate_vaf_summary(tukGermanPanelVars |> rename(Age = age_at_sampling, gnomADe_AF = gnomAD_AF)) |> 
  left_join(tukExomeVafSums) |> filter(!Sample_PD_ID %in% c("PD50508f_tds0001", "PD50527b_tds0001"))
vafSumsVars <- generate_vaf_summary(germanTrioTargVars)

tukVafSumsExp <- read_tsv(paste0(path, "sperm/twinsUK/expVafSumsTUK.tsv"), col_types = 'dcc', col_names = c("vafSum", "Sample_PD_ID", "vafGroup")) |> 
  mutate(vafGroup = str_remove(str_remove(vafGroup, "vepCovered"), "_")) |> 
  # Rename
  mutate(vafGroup = str_replace(vafGroup, "MODIFIER", "non-coding")) |> 
  mutate(vafGroup = str_replace(vafGroup, "HIGH", "LOF_snv")) |> 
  mutate(vafGroup = str_replace(vafGroup, "Thanatophoricdysplasia", "Thanatophoric dysplasia")) |> 
  mutate(vafGroup = str_replace(vafGroup, "CrouzonPfeiffer", "Crouzon/Pfeiffer")) |> 
  # Add indel only expectations
  bind_rows(read_tsv(paste0(path, "sperm/twinsUK/valsExpectedIndelTarg.tsv"), col_types = cols())) |> 
  # Summarize to combine indel + snv combos
  group_by(Sample_PD_ID, vafGroup) |> summarize(vafSum = sum(vafSum), .groups = "drop") |> 
  left_join(tukSamplesFiltered |> select(Sample_PD_ID, age_at_sampling, nanoseq), by = join_by(Sample_PD_ID)) |> 
  mutate(method = "exp") |> 
  filter(nanoseq == "Exome")


write_tsv(tukVafSumsExp, paste0(localPath, "data/tukVafSumsExp.tsv"))
write_tsv(tukVafSums, paste0(localPath, "data/tukGermanPanelVafSums.tsv"))
write_tsv(vafSumsVars, paste0(localPath, "data/vafSumsVars.tsv"))
```

### Mosaic Extend to regions
```{r mosaicSources, fig.width=7, fig.height=5}

# Trio vars coding fraction
exonicFraction <- (trioDNM_Vars |> filter(IMPACT != "MODIFIER") |> dim())[1]/dim(trioDNM_Vars)[1]
exonicFraction

tukExomes <- read_tsv(paste0( "<local_working_dir>/nanoSampleMeta/","targetedSamplesFiltered.tsv"), col_types = cols()) |> 
  filter(nanoseq == "Exome") |> pull(Sample_PD_ID)

# Fraction of path
pathFractionExp <- read_tsv(paste0(localPath, "data/tukVafSumsExp.tsv"), col_types = cols()) |>
  filter(Sample_PD_ID %in% tukExomes) |> 
  group_by(vafGroup) |> summarize(vafSum = mean(vafSum))
pathFractionExpInd <- read_tsv(paste0(path, "sperm/twinsUK/valsExpectedIndelTarg.tsv"), col_types = cols()) |> 
  filter(Sample_PD_ID %in% tukExomes) |> 
  group_by(vafGroup) |> summarize(vafSum = mean(vafSum))
# Mean coding vafSum. Have for indel directly, SNVs can add up
codingExp <- (pathFractionExp |> filter(vafGroup == "HIGH_vepCovered") |> pull(vafSum)) + 
  (pathFractionExp |> filter(vafGroup == "missense_vepCovered") |> pull(vafSum)) + 
  (pathFractionExp |> filter(vafGroup == "synonymous_vepCovered") |> pull(vafSum)) + 
  (pathFractionExpInd |> filter(vafGroup == "coding_indel") |> pull(vafSum)) 
# Mean pathogenic
pathExp <- (pathFractionExp |> filter(vafGroup == "vepCoveredClinDDlofCADD") |> pull(vafSum)) + 
  (pathFractionExpInd |> filter(vafGroup == "ClinDDlofCADD") |> pull(vafSum))

pathFraction <- pathExp/codingExp
pathFraction

# Relative fractions for each region
fractions <- tribble(
  ~region,       ~mult,
  "Genome", 1,
  "Exome",       exonicFraction,
  "Pathogenic",  exonicFraction*pathFraction
  #"GermanPanel",        0.0002,
  #"GermanPanelPath",   0.001 * 0.02
)

regions_overall_individual <- overall_individual |>
  rename(base_ID = ID) |>  
  crossing(fractions) |>
  mutate(
    ID = factor(region, levels = fractions$region),
    class = "Exp",
    n_variants            = n_variants * mult,
    ci_low_n_variants     = ci_low_n_variants * mult,
    ci_high_n_variants    = ci_high_n_variants * mult,
    expected_transmitted  = expected_transmitted * mult,
    ci_low_transmitted    = ci_low_transmitted * mult,
    ci_high_transmitted   = ci_high_transmitted * mult) 

pMosaicCount <- ggplot(regions_overall_individual, aes(x = ID, y = n_variants)) +
  geom_point() +
  geom_errorbar(aes(ymin = ci_low_n_variants, ymax = ci_high_n_variants), width = 0.1) +
  scale_y_log10() +
  ylab("Number of 1%-30% VAF variants in sperm") +
  theme_pubr() +
  theme(axis.title.x = element_blank())

pMosaicTransmitted <- ggplot(regions_overall_individual, aes(x = ID, y = expected_transmitted)) +
  geom_point() +
  geom_errorbar(
    aes(ymin = ci_low_transmitted, ymax = ci_high_transmitted),
    width = 0.1
  ) +
  scale_y_log10() +
  ylab("Transmitted 1%–30% VAF variants as DNMs") +
  theme_pubr() +
  theme(axis.title.x = element_blank())

pMosaicCount + pMosaicTransmitted

```




## e) Signatures
Create file postNextflow/signatures/run_SigProfiler_Extractor.v1.8.py
### Sigprofiler script
```{python signatures, eval=FALSE}
#!/usr/bin/env python
import sys, argparse
from SigProfilerExtractor import sigpro as sig

parser = argparse.ArgumentParser()
parser.add_argument("input", help="input matrix file or folder contating vcf files")
parser.add_argument("-o", "--output", type=str, default="output_folder", help="output folder")
parser.add_argument("-d", "--data_type", type=str, default="vcf")
parser.add_argument("-r", "--reference", type=str, default="GRCh37")
args = parser.parse_args()
print(args)

if __name__=="__main__":
	sig.sigProfilerExtractor(args.data_type, args.output, args.input, cpu=8, gpu=True, reference_genome=args.reference, opportunity_genome=args.reference, minimum_signatures=1, maximum_signatures=10)

```

### Submit standard
```{bash sigs, eval=FALSE}

#Move VCFs to input folder
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/
mkdir -p signatures
cd signatures
mkdir -p sigprofiler_in headerless jobLogs 
mkdir -p sigprofiler_Aug25 sigprofiler_Aug25norm

for sample in $(cut -f1 <hpc_scratch>/sperm/germanTrio_data/germanTrioSamplesFiltered.txt); do
  echo $sample
  if grep -q "$sample" <hpc_scratch>/sperm/germanTrio_data/postNextflow/vep/sigInput.vcf; then
    grep "$sample" <hpc_scratch>/sperm/germanTrio_data/postNextflow/vep/sigInput.vcf > headerless/$sample.vcf
    cat <hpc_scratch>/genomicRegions/vcfHeader_GRCh37d5.txt headerless/$sample.vcf > sigprofiler_in/$sample.vcf
  fi
done

#Run signature extraction with no matrix normalization
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/signatures
rm sigprofiler_in/input/*.vcf
bsub -J sigprofiler_Aug25 -o jobLogs/sig.%J.out -e jobLogs/sig.%J.err -q gpu-normal -gpu - -R 'select[mem>=32000] rusage[mem=32000]' -M32000 "module load sigprofiler/extractor-1.1.24-GRCh38-GRCh37; python run_SigProfiler_Extractor.v1.8.py sigprofiler_in -o sigprofiler_Aug25"
```

### Matrix normalization 
#### Create Count File
``` {bash matrixNorm}
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/signatures/
rm trinucCounts.tsv
#Get trinucleotide counts for matrix normalization
for sample in `cut -f1 <hpc_scratch>/sperm/germanTrio_data/germanTrioSamplesFiltered.txt`;
  do  echo $sample
  grep -v  'tri_bg' <hpc_scratch>/sperm/germanTrio_data/canAppsOutput/$sample/*.summary/$sample*.trint_counts_and_ratio2genome.tsv | sed "s/$/\t$sample/" >> trinucCounts.tsv
done

```

#### Normalize matrix by nanoseq coverage
```{r signaturesNormalize}

#Read in trinuc counts and ratio to full genome
backgroundRates <- read_tsv("<hpc_scratch>/sperm/germanTrio_data/postNextflow/signatures/trinucCounts.tsv", col_types = cols(), col_names = c("trinuc", "trinucCount", "ratio2genome", "Sample_PD_ID")) 
  
#Get mutation matrix generated by sigprofiler and then correct for ratio to genome
matrixNorm <- read_tsv("<hpc_scratch>/sperm/germanTrio_data/postNextflow/signatures/sigprofiler_in/output/SBS/sigprofiler_in.SBS96.all", col_types = cols()) |> 
  pivot_longer(cols = starts_with("PD"), names_to = "Sample_PD_ID", values_to = "mutCount") |> 
  mutate(trinuc = paste0(str_sub(MutationType, 1,1), str_sub(MutationType, 3,3),str_sub(MutationType, 7,7))) |> 
  left_join(backgroundRates, by = c("Sample_PD_ID", "trinuc")) |> 
  mutate(normCount = round(mutCount * 1/ratio2genome)) |> 
  select(MutationType, Sample_PD_ID,normCount) |> drop_na() |> 
  pivot_wider(names_from = Sample_PD_ID, values_from = normCount) 

write_tsv(matrixNorm, "<hpc_scratch>/sperm/germanTrio_data/postNextflow/signatures/sigprofiler_in/output/SBS/sigprofiler_in.SBS96.normalized", col_names = T)

# Get normalized for all samples combined due to low mutation count per sample
matrixNormCombined <- read_tsv("<hpc_scratch>/sperm/germanTrio_data/postNextflow/signatures/sigprofiler_in/output/SBS/sigprofiler_in.SBS96.all", col_types = cols())|> 
  pivot_longer(cols = starts_with("PD"), names_to = "Sample_PD_ID", values_to = "mutCount") |> 
  mutate(trinuc = paste0(str_sub(MutationType, 1,1), str_sub(MutationType, 3,3),str_sub(MutationType, 7,7))) |> 
  left_join(backgroundRates, by = c("Sample_PD_ID", "trinuc")) |> 
  mutate(normCount = mutCount * 1/ratio2genome) |> 
  select(MutationType, Sample_PD_ID,normCount) |> drop_na() |> 
  mutate(dataset = if_else(str_detect(Sample_PD_ID, "PD5"), "Oxford", "GermanTrio")) |> 
  group_by(MutationType) |> 
summarize(
  Oxford = round(sum(normCount[dataset == "Oxford"], na.rm = TRUE)),
  GermanTrio = round(sum(normCount[dataset == "GermanTrio"], na.rm = TRUE)),
  .groups = "drop"
)

write_tsv(matrixNormCombined, "<hpc_scratch>/sperm/germanTrio_data/postNextflow/signatures/sigprofiler_in/output/SBS/sigprofiler_in.SBS96.normalized.combined", col_names = T)


```

#### Submit normalized
```{bash sigs, eval=FALSE}
# Per sample
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/signatures/
bsub -J sigprofiler_Aug25norm -o jobLogs/sig.%J.out -e jobLogs/sig.%J.err -q gpu-normal -gpu - -R 'select[mem>=32000] rusage[mem=32000]' -M32000 "module load sigprofiler/extractor-1.1.24-GRCh38-GRCh37; python run_SigProfiler_Extractor.v1.8.py sigprofiler_in/output/SBS/sigprofiler_in.SBS96.normalized --data_type matrix -o sigprofiler_Aug25norm"

# All samples
cd <hpc_scratch>/sperm/germanTrio_data/postNextflow/signatures/
bsub -J sigprofiler_Aug25normAll -o jobLogs/sig.%J.out -e jobLogs/sig.%J.err -q gpu-normal -gpu - -R 'select[mem>=32000] rusage[mem=32000]' -M32000 "module load sigprofiler/extractor-1.1.24-GRCh38-GRCh37; python run_SigProfiler_Extractor.v1.8.py sigprofiler_in/output/SBS/sigprofiler_in.SBS96.normalized.combined --data_type matrix -o sigprofiler_Aug25normAll"


```

#### Trinuc DNM counts
```{r trinucDNMs}
trinucDNMs <- read_tsv(paste0(path, "sperm/germanTrio_data/dnmData/trinucCountsDNMs.tsv"), col_types = cols()) |>
  select(-c(PD67073b, PD64157a)) |>
  rowwise() |>
  mutate(dnmCount = sum(c_across(-MutationType))) |>
  ungroup() |>
  mutate(TRI = str_replace(MutationType, "([ACGT])\\[([ACGT])>([ACGT])\\]([ACGT])", "\\1\\2\\4>\\3")) |>
  select(TRI, dnmCount)
sum(trinucDNMs$dnmCount)

write_tsv(trinucDNMs, paste0(localPath, "data/trinucDNMs.tsv"))
```


## f) TwinsUK model fits (for public code fallback)
TwinsUK individual-level VAF sums (tukVafSums etc.) are under TwinsUK's controlled-access
policy and can't be shared publicly (see README in PublicCode/). The age-trend models
fitted on them are pure aggregate summaries (an Age grid + fitted value + 95% CI, or a
handful of scalar stats) with no way to recover individual data points, so those are
safe to export and let the public figures code draw the TwinsUK trend line/reported
stats without needing the underlying samples.

### Basic burdens
```{r targBurden, fig.height=5, fig.width=8}
 
burdenTwinsUKTarg <- read_tsv(paste0("<local_working_dir>/nanoseqTwinsUK_Targ/data/burdensFiltered.tsv"), col_types = cols()) |>
  left_join(read_tsv(paste0("<local_working_dir>/nanoSampleMeta/targetedSamplesFiltered.tsv"), col_types = cols()), by = "Sample_PD_ID") |>
  mutate(nanoseq = "Controls") |>
  rename(Age = age_at_sampling) |> drop_na(Age)

# TwinsUK burden data is individual-level and controlled-access (see the
# "TwinsUK model fits" note in ## d) Observed VAFs). Save the age-trend fits
# only (aggregate Age grid + fit + CI, using the same age_grid_fine/fit_lm_grid
# helpers defined there) so the public figures code can draw the TwinsUK trend
# line without the underlying samples.
twinsUK_burdenFits <- bind_rows(
  fit_lm_grid(lm(burdenCorr ~ Age, data = burdenTwinsUKTarg)) |> mutate(model = "burdenCorr"),
  fit_lm_grid(lm(indelBurden ~ Age, data = burdenTwinsUKTarg)) |> mutate(model = "indelBurden")
)
write_tsv(twinsUK_burdenFits, paste0(localPath, "data/twinsUK_burdenFits.tsv"))

burdenCombined <- germanTrioBurdens |>
  mutate(burdenCorr = burdenCorr1percent, indelBurden = burdenIndelPost1percent) |> 
  mutate(nanoseq = if_else(str_detect(Sample_PD_ID, "PD5"),"Apert fathers", "DD fathers")) |> 
  bind_rows(burdenTwinsUKTarg) 

p_germanTrioBurdens <- ggplot(burdenCombined, aes(x = Age, y = burdenCorr, fill = nanoseq)) + 
  geom_point(size=3.5, pch=21, col='white', stroke=0.25) + 
  geom_smooth(method = "lm") + 
  scale_fill_manual(values = dataset_colours) +
   theme(text = element_text(size=16)) + 
  xlab('Age (years)') + 
  ylab('Mutations per bp') + 
  labs(fill='Dataset') + 
  theme_minimal() + 
  ylim(0,NA) 

p_germanTrioBurdens

p_indelBurden<- ggplot(burdenCombined, aes(x = Age, y = indelBurden, fill = nanoseq)) + 
  geom_point(size=3.5, pch=21, col='white', stroke=0.25) + 
  scale_alpha_manual(values = c(1, 1, 0.8), guide = "none") + 
   theme(text = element_text(size=16)) + 
  geom_smooth(method = "lm") + 
  scale_fill_manual(values = dataset_colours) +
  xlab('Age (years)') + 
  ylab('Indels per bp') + 
  labs(fill='Dataset') + 
  theme_minimal() + 
  ylim(0,NA) 

p_indelBurden
```

### VafSums
```{r twinsUKVafFits}
# Wider than the observed TwinsUK age range so ages back-inferred to conception
# (Age - paternalAgeGap) on the public side stay within the saved grid.
age_grid_fine <- tibble(Age = seq(10, 80, by = 1))

# Predict a model over the age grid with 95% CI - works for any simple lm(y ~ Age).
# Linear interpolation between grid points reproduces predict() exactly for a
# straight-line model, so this same grid also supports point predictions at
# arbitrary (non-grid) ages downstream via approx().
fit_lm_grid <- function(model, ages = age_grid_fine$Age) {
  pred <- predict(model, newdata = tibble(Age = ages), se.fit = TRUE, interval = "confidence", level = 0.95)
  tibble(Age = ages, fit = pred$fit[,"fit"], ci_lo = pred$fit[,"lwr"], ci_hi = pred$fit[,"upr"])
}

twins_model <- lm(exome_vafSum ~ Age, data = tukVafSums |> filter(vafGroup == "ClinDDlofCADD"))
twins_model_panel <- lm(vafSum ~ Age, data = tukVafSums |> filter(vafGroup == "ClinDDlofCADD"))
twins_model_driver <- lm(vafSum ~ Age, data = tukVafSums |> filter(vafGroup == "posPath"))
model_exome_posPath <- tukVafSums |> filter(vafGroup == "posPath") |> lm(exome_vafSum ~ Age, data = _)

# Quasibinomial smooth (used for the driver-rate panels) is not a simple lm, so the
# grid is predicted on the link scale then back-transformed, matching what
# stat_smooth(method = "glm", family = "quasibinomial") does internally.
fitQB_allPos <- glm(vafSum ~ Age, family = quasibinomial, data = tukVafSums |> filter(vafGroup == "allPos"))
predsQB <- predict(fitQB_allPos, newdata = age_grid_fine, type = "link", se.fit = TRUE)
gridQB <- tibble(
  Age = age_grid_fine$Age,
  fit = plogis(predsQB$fit),
  ci_lo = plogis(predsQB$fit - 1.96 * predsQB$se.fit),
  ci_hi = plogis(predsQB$fit + 1.96 * predsQB$se.fit)
) |> mutate(model = "vafSum_allPos_quasibinom")

# Note: "ClinDDlofCADD" is the internal vafGroup label from generate_vaf_summary();
# it's the same selection as what the public figures code calls "likelyDisease",
# so the saved model names use that more readable label instead.
twinsUK_vafFits <- bind_rows(
  fit_lm_grid(twins_model) |> mutate(model = "exome_vafSum_likelyDisease"),
  fit_lm_grid(twins_model_panel) |> mutate(model = "vafSum_likelyDisease"),
  fit_lm_grid(twins_model_driver) |> mutate(model = "vafSum_posPath"),
  fit_lm_grid(model_exome_posPath) |> mutate(model = "exome_vafSum_posPath"),
  gridQB
)
write_tsv(twinsUK_vafFits, paste0(localPath, "data/twinsUK_vafFits.tsv"))

# Exome-to-panel capture fraction: a single scalar plus reporting stats, all
# aggregate (see "Calibrate exome correction" in the public/DD-father section
# for how this constant is used).
fractionWithinPanel <- tukVafSums |>
  group_by(vafGroup) |>
  summarize(totalVAFs = sum(vafSum, na.rm = TRUE), totalVAFsExome = sum(exome_vafSum, na.rm = TRUE), .groups = "drop") |>
  mutate(percentageInPanel = totalVAFs / totalVAFsExome)
exomePanelCorrection <- fractionWithinPanel |> filter(vafGroup == "ClinDDlofCADD") |> pull(percentageInPanel)

panel_exome_df <- tukVafSums |>
  filter(vafGroup == "ClinDDlofCADD") |>
  mutate(percentageInPanel = vafSum / exome_vafSum) |>
  filter(is.finite(percentageInPanel), percentageInPanel > 0)

cor_test <- cor.test(panel_exome_df$vafSum, panel_exome_df$exome_vafSum, method = "pearson")
fit_age <- lm(percentageInPanel ~ Age, data = panel_exome_df)
fit_age_sum <- summary(fit_age)

twinsUK_exomePanelCorrection_stats <- tibble(
  exomePanelCorrection = exomePanelCorrection,
  n_samples = nrow(panel_exome_df),
  n_sources = dplyr::n_distinct(panel_exome_df$source),
  cor_estimate = unname(cor_test$estimate),
  cor_p_value = cor_test$p.value,
  age_beta_per_year = coef(fit_age)["Age"],
  age_se = coef(fit_age_sum)["Age", "Std. Error"],
  age_p_value = coef(fit_age_sum)["Age", "Pr(>|t|)"],
  age_r_squared = fit_age_sum$r.squared,
  mean_frac = mean(panel_exome_df$percentageInPanel, na.rm = TRUE),
  median_frac = median(panel_exome_df$percentageInPanel, na.rm = TRUE),
  sd_frac = sd(panel_exome_df$percentageInPanel, na.rm = TRUE),
  iqr_frac = IQR(panel_exome_df$percentageInPanel, na.rm = TRUE)
)
write_tsv(twinsUK_exomePanelCorrection_stats, paste0(localPath, "data/twinsUK_exomePanelCorrection_stats.tsv"))

twinsUK_vafFits <- bind_rows(twinsUK_vafFits, fit_lm_grid(fit_age) |> mutate(model = "percentageInPanel_exomeCorrection"))
write_tsv(twinsUK_vafFits, paste0(localPath, "data/twinsUK_vafFits.tsv"))
```

# 5. Supplementary Tables
### SupTable 1: Overview
```{r supTable1}

trioIDs <- read_xlsx(paste0(metaPath, "Tables/controls_samplesEGA.xlsx"))

id_lut <- setNames(trioIDs$Sanger_Sample_ID, trioIDs$Canapps_Sample_name)

controlTrios <- read_tsv(paste0(metaPath, "Tables/metadata_14Apr2025_completefamilies_DNMs_counts-Jan2026.txt"), col_types = cols()) |>
  mutate(Cohort = "ControlTrios") |>
  mutate(sex = if_else(`sex(F=2,M=1)` == 1, "M", "F")) |>
  mutate(FAM_ID = paste0("SFHS_", famid)) |>
  rename(PD_ID = Sample, Pat_ID = father, Mat_ID = mother) |>
  mutate(across(c(PD_ID, Pat_ID, Mat_ID), ~{ m <- id_lut[.x]; ifelse(is.na(m), .x, unname(m)) })) |>
  rename(Pat_Age_at_birth = m_age, Mat_Age_at_birth = f_age)|> group_by(FAM_ID) |>
  arrange(Pat_Age_at_birth, .by_group = TRUE) |>
  mutate(FAM_ID = if (dplyr::n() > 1) paste0(FAM_ID, "_", dplyr::row_number()) else FAM_ID) |>
  ungroup()

sperm0 <- covSummaryTarg |>
  mutate(SpermNanoseqData = T) |> 
  mutate(SpermNanoseqFilter = if_else(sampleFilter == "FULL", "Pass", sampleFilter)) |> 
  select(FAM_ID, Cohort, SpermNanoseqData, SpermNanoseqFilter)

SupTable1_Overview <- read_tsv(paste0(metaPath, "Tables/metadata_DNMs_counts-Jan2026.txt"), col_types = cols()) |> 
  mutate(Cohort = "DD fathers", FAM_ID = str_replace(FamID, "F", "FAM_")) |> 
  filter(FAM_ID != "FAM_179") |> # Accidentally sent sister's blood and not index
  bind_rows(controlTrios) |> 
  mutate(TrioWGSdata = T) |> 
  mutate(TrioWGSfilter = if_else(PD_ID == "PD64157a" & Cohort == "DD fathers", "QCfail_metadata", "Pass")) |> 
  arrange(FAM_ID) |> 
  select(FAM_ID, Cohort, TrioWGSdata, TrioWGSfilter) |> 
  full_join(sperm0, by = join_by(FAM_ID, Cohort)) |> 
  arrange(FAM_ID) |> 
  mutate(TrioWGSdata = coalesce(TrioWGSdata, F)) |> 
  mutate(SpermNanoseqData = coalesce(SpermNanoseqData, F))

write_tsv(SupTable1_Overview, paste0(manuscriptPath, "SupTables/SupTable1_Overview.tsv"))

```

### SupTable 3: Trios
```{r supTable3}

SupTable3_Trios <- read_tsv(paste0(metaPath, "Tables/metadata_DNMs_counts-Jan2026.txt"), col_types = cols()) |> 
  mutate(Cohort = "DD fathers", FAM_ID = str_replace(FamID, "F", "FAM_")) |> 
  filter(FAM_ID != "FAM_179") |> # Accidentally sent sister's blood and not index
  filter(PD_ID != "PD64157a") |> # Removed for QC
  bind_rows(controlTrios) |> 
  mutate(Age = as.integer(Age)) |> suppressWarnings() |> 
  arrange(FamID) |> 
  select(FAM_ID, Cohort, 
         Pat_age_at_birth = Pat_Age_at_birth, Mat_age_at_birth = Mat_Age_at_birth, 
         Proband_median_cov = MEDIAN_COVERAGE, Pat_median_cov = Pat_seq_x, Mat_median_cov = Mat_seq_x, 
         SNP_DNMs_0.3VAF, Indel_DNMs_0.3VAF, DNM_0.3_medianVAF = sample.median.0.3vaf, SBS1, SBS5,
         PD_ID_Proband = PD_ID, PD_ID_Pat = Pat_ID, PD_ID_Mat = Mat_ID)

write_tsv(SupTable3_Trios, paste0(manuscriptPath, "SupTables/SupTable3_Trios.tsv"))

```

### SupTable 2: DNM calls
```{r supTable2}

trioDNMPhase <- read_tsv(paste0(path, "sperm/germanTrio_data/dnmData/All_PhaseMyDeNovo_output_0.3VAF.txt"), col_types = cols()) |>
  filter(id != "id") |>
  select(PD_ID = id, chr = chrom, start = pos, ref, alt, phase = `phase...14`) |>
  mutate(start = as.numeric(start))

SupTable2_DNMs <- read_tsv(paste0(metaPath, "trioDNM_Vars.tsv"), col_types = cols())|>
  left_join(trioDNMPhase, by = join_by(chr, start, ref, alt, PD_ID)) |>
  mutate(phase = coalesce(phase, "unknown")) |>
  mutate(phase = if_else(phase == "F,M", "unknown", phase)) |>
  left_join(SupTable3_Trios |> select(FAM_ID, PD_ID = PD_ID_Proband), by = join_by(PD_ID, FAM_ID)) |>
  select(chr, start_hg19 = start, ref, alt, FAM_ID, PD_ID_Proband = PD_ID, Cohort, phase, VEP_csq = csq, VEP_IMPACT = IMPACT, gene, gnomADe_AF, cadd,
         ClinVar_CLNDN = PHEN, ClinVar_CLNSIG = CLNSIG, ClinVar_CLNSIGCONF = CLNSIGCONF, likelyDisease, likelyDriver) |>
  arrange(FAM_ID, chr, start_hg19) |>
  distinct()

write_tsv(SupTable2_DNMs, paste0(manuscriptPath, "SupTables/SupTable2_DNMs.tsv"))
```

### SupTable 6: Sperm Samples
```{r supTable6}
SupTable6_SpermSamples <- read_tsv(paste0(metaPath,"germanTrioSamplesFiltered.tsv"), col_types = cols()) |> 
  mutate(Tissue = "Semen") |> 
  mutate(Cohort = if_else(str_detect(Sample_PD_ID, "PD6"), "DD fathers", "Apert fathers")) |> 
  left_join(read_tsv(paste0(localPath, "data/germanTrioBurdens.tsv"), col_types = cols()), by = join_by(Sample_PD_ID)) |> 
  select(PD_ID, FAM_ID, Cohort, Tissue, Age, total_duplex_bp_cov = total, 
         snvs = muts, snv_burden = burden, snv_burden_corr = burdenCorr, snv_burden_corr_0.01vaf = burdenCorr1percent, 
         indels, indel_burden = indelBurden, indel_burden_0.01vaf = burdenIndelPost1percent)

write_tsv(SupTable6_SpermSamples, paste0(manuscriptPath, "SupTables/SupTable6_SpermSamples.tsv"))

```

### SupTable 4: Diag DNMs
```{r supTable4}

SupTable4_DiagDNMs <- read_tsv(paste0(metaPath, "diagnosticVarsPos.txt"), col_types = cols()) |>
  distinct() |>
  filter(FAM_ID %in% SupTable1_Overview$FAM_ID | FAM_ID %in% SupTable6_SpermSamples$FAM_ID | Cohort == "Apert fathers") |>
  rename(inDNMcallSet = called_TrioWGS, reasonNotInDNMcallSet = missing_TrioWGS) |>
  mutate(reasonNotInDNMcallSet = if_else(is.na(inDNMcallSet), "No Trio WGS data", reasonNotInDNMcallSet)) |>
  left_join(read_xlsx(paste0(metaPath, "SupTable4_DiagDNM_SoNs.xlsx"), sheet = 1) |> select(VAR_ID, ANALYSIS_PERFORMED)|>
  separate_rows(VAR_ID, sep = ",\\s*") |> drop_na(VAR_ID), by = join_by(VAR_ID)) |>
  mutate(ANALYSIS_PERFORMED = if_else(Cohort == "Apert fathers", "Apert ascertained", ANALYSIS_PERFORMED)) |>
  mutate(ANALYSIS_PERFORMED = if_else(gene == "SCN1A" & !FAM_ID %in% c("FAM_050", "FAM_087", "FAM_091", "FAM_144"), "SCN1A ascertained", ANALYSIS_PERFORMED)) |>
  mutate(patSpermSequenced = if_else(FAM_ID %in% SupTable6_SpermSamples$FAM_ID, "Yes", "No")) |>
    left_join(read_tsv(paste0(localPath, "data/diagnosticVarsCalls.tsv"), col_types = cols()) |>
              filter(group == "Target") |>
              select(VAR_ID, FAM_ID, sperm_duplexCalls = duplexCalls, sperm_duplexCov = duplexCov, sperm_duplexVAF = duplexVAF,
                     sperm_BAM_calls = BAMcalls, sperm_BAM_cov = BAM_COV, sperm_BAM_VAF = BAM_VAF),
            by = join_by(VAR_ID, FAM_ID))

write_tsv(SupTable4_DiagDNMs, paste0(manuscriptPath, "SupTables/SupTable4_DiagDNMs.tsv"))

write.xlsx(SupTable4_DiagDNMs |> filter(Cohort == "DD fathers"), paste0(manuscriptPath, "SupTables/SupTable4_DiagDNMs.xlsx"))
table(SupTable4_DiagDNMs$reasonNotInDNMcallSet)

```

### SupTable 5: Clinical Review
```{r supTable5}

SupTable5_ClinicalReview <- read_csv(paste0(metaPath, "Tables/JoeClinicalReview3.0.csv"), col_types = cols()) |> 
  filter(!Group %in% c("Diagnostic variant", "Diagnostic variant (Uncalled)", "No known  clinical phenotype", "Unlikely  clinical relevance")) |> 
  mutate(across(`Variant classification`:Comments, ~replace_na(.x, "N/A")))
table(SupTable5_ClinicalReview$Group)

write_tsv(SupTable5_ClinicalReview, paste0(manuscriptPath, "SupTables/SupTable5_ClinicalReview.tsv"))

```



### SupTable 7: Target Panels
```{r supTable6}
SupTable7_TargetPanels <- read_tsv("<local_working_dir>/genePanel/GermlinePanelV2/germlineV2_hg19.bed", col_types = 'cii--', col_names= c("chr", "start_hg19", "end_hg19")) |> mutate(DD_fathersPanel = T) |> 
  full_join(read_tsv("<local_working_dir>/genePanel/GermlinePanelV1/germlineV1_hg19.bed", col_types = 'cii--', col_names= c("chr", "start_hg19", "end_hg19")) |> mutate(Apert_fathersPanel = T)) |> 
  arrange(chr, start_hg19) |> 
  mutate(DD_fathersPanel = coalesce(DD_fathersPanel, FALSE),
    Apert_fathersPanel = coalesce(Apert_fathersPanel, FALSE))
  
write_tsv(SupTable7_TargetPanels, paste0(manuscriptPath, "SupTables/SupTable7_TargetPanels.tsv"))

```

### SupTable 8: Sperm Vars
```{r supTable8}
SupTable8_SpermVars <- read_tsv(paste0(localPath, "data/germanTrioTargVars.tsv"), col_types = cols()) |> 
  mutate(Cohort = if_else(str_detect(Sample_PD_ID, "PD6"), "DD fathers", "Apert fathers")) |> 
  select(chr, start_hg19 = start, ref, alt, TRI, TYPE, DUPLEX_MUT = TIMES_CALLED, DUPLEX_COV, DUPLEX_VAF,
         BAM_MUT, BAM_COV, BAM_VAF, PD_ID, Age, FAM_ID, Cohort, gene, VEP_csq = csq, VEP_IMPACT = IMPACT, gnomADe_AF, cadd,
         ClinVar_CLNDN = PHEN, ClinVar_CLNSIG = CLNSIG, ClinVar_CLNSIGCONF = CLNSIGCONF, likelyDisease, likelyDriver)

write_tsv(SupTable8_SpermVars, paste0(manuscriptPath, "SupTables/SupTable8_SpermVars.tsv")) 

```

### SupTable 9: dNdS gene level
```{r supTable9}
SupTable9_dNdSgeneLevel <- read_tsv(paste0(localPath, "data/SupTable_sig_dndsAll.tsv"), col_types = cols())
write_tsv(SupTable9_dNdSgeneLevel, paste0(manuscriptPath, "SupTables/SupTable9_dNdSgeneLevel.tsv"))

```

### SupTable 10: dNdS hotspot
```{r supTable10}
SupTable10_dNdShotspot <- read_tsv(paste0(localPath, "data/siteHotspots_dndsBP_all.tsv"), col_types = cols())
  
write_tsv(SupTable10_dNdShotspot, paste0(manuscriptPath, "SupTables/SupTable10_dNdShotspot.tsv"))

```



#### Combined xlsx
Run this to rewrite all to excel at once
```{r supTables}
supTables <- createWorkbook()

# Add each table as a new sheet
addWorksheet(supTables, "Table S1 Overview")
writeData(supTables, "Table S1 Overview", SupTable1_Overview)

addWorksheet(supTables, "Table S2 DNMs")
writeData(supTables, "Table S2 DNMs", SupTable2_DNMs)

addWorksheet(supTables, "Table S3 Trios")
writeData(supTables, "Table S3 Trios", SupTable3_Trios)

addWorksheet(supTables, "Table S4 DiagDNMs")
writeData(supTables, "Table S4 DiagDNMs", SupTable4_DiagDNMs)

addWorksheet(supTables, "Table S5 ClinicalReview")
writeData(supTables, "Table S5 ClinicalReview", SupTable5_ClinicalReview)

addWorksheet(supTables, "Table S6 SpermSamples")
writeData(supTables, "Table S6 SpermSamples", SupTable6_SpermSamples)

addWorksheet(supTables, "Table S7 TargetPanels")
writeData(supTables, "Table S7 TargetPanels", SupTable7_TargetPanels)

addWorksheet(supTables, "Table S8 SpermVars")
writeData(supTables, "Table S8 SpermVars", SupTable8_SpermVars)

addWorksheet(supTables, "Table S9 dNdSgeneLevel")
writeData(supTables, "Table S9 dNdSgeneLevel", SupTable9_dNdSgeneLevel)

addWorksheet(supTables, "Table S10 dNdShotspot")
writeData(supTables, "Table S10 dNdShotspot", SupTable10_dNdShotspot)

# Save the combined workbook
saveWorkbook(supTables, paste0(manuscriptPath, "Supplementary_Tables.xlsx"), overwrite = TRUE)

# Save each table as an individual xlsx file
write.xlsx(list("Table S1 Overview"       = SupTable1_Overview),       paste0(manuscriptPath, "SupTables/TableS1.xlsx"), overwrite = TRUE)
write.xlsx(list("Table S2 DNMs"           = SupTable2_DNMs),           paste0(manuscriptPath, "SupTables/TableS2.xlsx"), overwrite = TRUE)
write.xlsx(list("Table S3 Trios"          = SupTable3_Trios),          paste0(manuscriptPath, "SupTables/TableS3.xlsx"), overwrite = TRUE)
write.xlsx(list("Table S4 DiagDNMs"       = SupTable4_DiagDNMs),       paste0(manuscriptPath, "SupTables/TableS4.xlsx"), overwrite = TRUE)
write.xlsx(list("Table S5 ClinicalReview" = SupTable5_ClinicalReview), paste0(manuscriptPath, "SupTables/TableS5.xlsx"), overwrite = TRUE)
write.xlsx(list("Table S6 SpermSamples"   = SupTable6_SpermSamples),   paste0(manuscriptPath, "SupTables/TableS6.xlsx"), overwrite = TRUE)
write.xlsx(list("Table S7 TargetPanels"   = SupTable7_TargetPanels),   paste0(manuscriptPath, "SupTables/TableS7.xlsx"), overwrite = TRUE)
write.xlsx(list("Table S8 SpermVars"      = SupTable8_SpermVars),      paste0(manuscriptPath, "SupTables/TableS8.xlsx"), overwrite = TRUE)
write.xlsx(list("Table S9 dNdSgeneLevel"  = SupTable9_dNdSgeneLevel),  paste0(manuscriptPath, "SupTables/TableS9.xlsx"), overwrite = TRUE)
write.xlsx(list("Table S10 dNdShotspot"   = SupTable10_dNdShotspot),   paste0(manuscriptPath, "SupTables/TableS10.xlsx"), overwrite = TRUE)
```
